### 2017-06-02 monocle_analysis.Rmd
first commit

---
title: "Monocle analysis of influenza-infected cells"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

## Setup
Load or install necessary packages.
If `Monocle` is not already available, install it using `Bioconductor`.
Note that Cole suggested that we might want to install the development version of GitHub; right now this does **not** do this.
```{r, message=FALSE, warning=FALSE}

if (!require("monocle")) {
  source("http://bioconductor.org/biocLite.R")
  biocLite()
  biocLite("monocle")
  }

if (!require("ggExtra")) install.packages("ggExtra")
if (!require("cowplot")) install.packages("cowplot")
```
```{r}
cat("Using Monocle version", as.character(packageVersion("monocle")), "\n")
```

## Create and filter cell-gene dataset
Read in cell-gene data of all samples.
This is the merged cell-gene data set constructed essentially from `cellranger` using the iPython notebook.
The cells are annotated by the number of wildtype and synonymous-barcoded influenza UMIs called for each cell.
```{r}
matrixfile <- "results/cellgenecounts/merged_humanplusflu_matrix.mtx"
genesfile <- "results/cellgenecounts/merged_humanplusflu_genes.tsv"
cellsfile <- "results/cellgenecounts/merged_humanplusflu_cells.tsv"

all_cells <- newCellDataSet(readMM(matrixfile), 
      phenoData=new("AnnotatedDataFrame", data=read.delim(cellsfile)), 
      featureData=new("AnnotatedDataFrame", data=read.delim(genesfile)),
      expressionFamily=negbinomial.size())
```

Filter on just the samples that interest us for this analysis.
```{r}
samples = c("Uninfected", "8hr-2", "8hr-2repeat", "10hr-2")

cellfilter <- row.names(subset(pData(all_cells), Sample %in% samples))
cells <- all_cells[,cellfilter]

cat("Filtered all", nrow(pData(all_cells)), "cells down to the", nrow(pData(cells)), 
      "in samples of interest.")
```

Compute the total number of mRNAs, cellular mRNAs, and influenza mRNAs per cell.
```{r}
pData(cells)$total_mRNAs <- Matrix::colSums(exprs(cells))

flugenes = c("fluPB2", "fluPB1", "fluPA", "fluHA", "fluNP", "fluNA", "fluM", "fluNS")
flu_id = row.names(subset(fData(cells), gene_short_name %in% flugenes))
pData(cells)$flu_mRNAs <- Matrix::colSums(exprs(cells[flu_id,]))

flugenes_id = row.names(subset(fData(cells), !(gene_short_name %in% flugenes)))
pData(cells)$cellular_mRNAs <- Matrix::colSums(exprs(cells[flugenes_id,]))
```

```{r}
scatterplot <- ggplot(pData(cells), aes(cellular_mRNAs, flu_mRNAs)) + geom_point()

# next line uses eval substitute to get value of maxy as here: https://stackoverflow.com/a/10659563
densityplot <- eval(substitute(
    {stat_density(aes(x=cellular_mRNAs, y=maxy*(..scaled..)), geom="line", color='blue')}, 
    list(maxy=layer_scales(scatterplot)$y$range$range[2])))

scatterplot + densityplot + theme(axis.text.x=element_text(angle=90, vjust=0.5)) + facet_wrap(~Sample)
```
The scatterplots above show the number of influenza mRNAs versus the number of cellular mRNAs for each sample. 
The blue curves show the overall density on the number of cellular mRNAs.
Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
The `Monocle` vignette suggests possibly filtering out cells with far more / less mRNAs than most cells.
I have **not** done that here for the reason that this will filter out quite a few of the influenza-infected cells.
This fact is made more clear by this next plot:
```{r}
all_densityplot <- ggplot(pData(cells), aes(total_mRNAs)) + geom_density()
flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 100))
flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
flugt100_densityplot + facet_wrap(~Sample)
```

```{r}
min_expr <- 5
num_cells <- 5
cells <- detectGenes(cells, min_expr=min_expr)
expressed_genes <- row.names(subset(fData(cells), num_cells_expressed >= num_cells))
cat("Filtering to the", length(expressed_genes), "genes expressed at >=", min_expr,
      "transcripts in at least", num_cells, "cells.\n")
cells <- cells[expressed_genes,]

cat("Plotting number of RNAs per cell for each sample\n")
print(table(pData(cells)$Sample))
pData(cells)$Total_mRNAs <- Matrix::colSums(exprs(cells))
qplot(Total_mRNAs, data=pData(cells), geom="density", color=Sample)

cat("Filtering out 6hr sample since it has a different distribution of RNA per cell")
cells <- cells[, row.names(subset(pData(cells), Sample != "6hr"))]
print(table(pData(cells)$Sample))
```

## Look at distribution of influenza genes.
```{r}
flu_id <- row.names(subset(fData(cells), substr(ShortGeneName, 1, 3) == "flu"))

pData(cells)$Total_flu <- Matrix::colSums(exprs(cells[flu_id,]))
qplot(Total_flu, data=pData(cells), color=Sample, log="x")

cth <- newCellTypeHierarchy()
cth <- addCellType(cth, "noFlu", classify_func=function(x) {colSums(x[flu_id,]) == 0})
cth <- addCellType(cth, "highFlu", classify_func=function(x) {colSums(x[flu_id,]) >= 100}) 
cells <- classifyCells(cells, cth)
print(table(pData(cells)$Sample, pData(cells)$CellType))
```

### 2017-06-02 data/flu_sequences/make_fasta_and_gtf.py

@@ -1,0 +1,83 @@
+"""Make FASTA and GTF files for vRNAs from Genbank plasmid maps."""
+
+
+import re
+import os
+import glob
+import Bio.SeqIO
+import Bio.Seq
+import Bio.SeqRecord
+import Bio.SeqFeature
+import BCBio.GFF
+
+
+def main():
+    """Main body of script."""
+    u12 = re.compile('AGC[GA]AAAGCAGG') # U12 with position 4 polymorphism
+    u13 = 'CCTTGTTTCTACT' # reverse of U13
+    segs = ['PB2', 'PB1', 'PA', 'HA', 'NP', 'NA', 'M', 'NS']
+
+    for syntype in ['', '-syn']:
+        vrnas = []
+        for seg in segs:
+            fplasmid = glob.glob('*_pHW18*-{0}{1}.gb'.format(seg, syntype))
+            assert len(fplasmid) == 1, "Didn't find 1 map for {0}{1}".format(
+                    seg, syntype)
+            fplasmid = fplasmid[0]
+            plasmid = str(Bio.SeqIO.read(fplasmid, 'genbank').seq)
+            assert plasmid.count(u13) == 1, "Not exactly one U13 for {0}{1}".format(
+                    seg, syntype)
+            u12matches = list(u12.finditer(plasmid))
+            assert len(u12matches) == 1, "Not exactly one U12 for {0}{1}".format(
+                    seg, syntype)
+            vrna = plasmid[u12matches[0].start(0) : plasmid.index(u13) + len(u13)]
+            print("vRNA of length {0} for {1}{2}".format(len(vrna), seg, syntype))
+            vrna = Bio.SeqRecord.SeqRecord(Bio.Seq.Seq(vrna), 
+                    id='flu' + seg + syntype,
+                    description='{0} vRNA from plasmid {1}'.format(seg, fplasmid),
+                    features=[Bio.SeqFeature.SeqFeature(
+                        Bio.SeqFeature.FeatureLocation(0, len(vrna)),
+                        type='exon',
+                        strand=1,
+                        qualifiers={
+                            "source":"JesseBloom",
+                            "gene_id":'flu' + seg,
+                            "transcript_id":'flu' + seg,
+                            "gene_biotype":"vRNA",
+                            },
+                        )],
+                    )
+            vrnas.append(vrna)
+        fastafile = 'flu-wsn{0}.fasta'.format(syntype)
[…]

### 2017-06-02 pymodules/__init__.py

@@ -1,0 +1,1 @@
+"""Custom Python modules used in this analysis."""

### 2017-06-02 pymodules/utils.py

@@ -1,0 +1,282 @@
+"""Python functions used in 10X single-cell analysis.
+
+Written by Jesse Bloom."""
+
+
+import math
+import os
+import json
+import time
+import subprocess
+import csv
+import scipy.io
+import scipy.sparse
+import pandas
+import matplotlib
+matplotlib.use('pdf')
+import matplotlib.pyplot as plt
+plt.style.use('ggplot')
+from IPython.display import Image, display
+
+
+def mergeCellGeneMatrices(mergedgenes, mergedcells, mergedmatrix,
+        samples, cellbarcodes, cellannotations, genes, matrices,
+        removeprefix=None):
+    '''Merges cell-gene matrices for a variety of ``cellranger`` samples.
+
+    Creates file suitable for analysis with ``Monocle``.
+
+    Does **not** perform any type of normalization, just merges the
+    counts.
+
+    Args:
+        `mergedgenes`, `mergedcells`, `mergedmatrix` (str)
+            Names of created file containing the gene names and
+            cell names (as ``*.tsv``) and gene-cell matrix (in
+            Market Exchange Format).
+        `samples` (list)
+            List of samples that we are merging.
+        `cellbarcodes`, `cellannotations`, `genes`, `matrices` (dict)
+            Each is a dictionary keyed by the each sample in `samples`.
+            The values are the names of the files giving the cell barcodes,
+            cell annotations, genes, and matrices for that sample. The
+            annotations can be `None`. The annotations file should have a
+            header, and the first column should give the cell barcode.
+        `removeprefix` (str or `None`)
+            If not `None`, should specify a prefix associated with each
+            gene. This prefix is removed in `mergedgenes`.
+
+    Result: 
+        Creates the files `mergedgenes`, `mergedcells`, and `mergedmatrix`.
+    '''
+    assert len(samples) == len(set(samples)) >= 1
+    genelist = annotations = None
+    cells = []
+    matrixlist = []
+    for sample in samples:
+        sample_m = scipy.io.mmread(matrices[sample])
+        with open(genes[sample]) as f:
+            sample_genes = f.readlines()
+        if genelist is None:
+            genelist = sample_genes
+        elif genelist != sample_genes:
+            raise ValueError("different gene sets for different matrices")
+        with open(cellbarcodes[sample]) as f:
[…]

### 2017-06-02 scripts/annotate_synbarcodes.py

@@ -1,0 +1,362 @@
+#!/usr/bin/python
+
+"""Annotate synonymous barcodes on reads."""
+
+
+import os
+import sys
+import subprocess
+import argparse
+import math
+import numpy
+import Bio.SeqIO 
+import pandas
+import matplotlib
+matplotlib.use('pdf')
+import matplotlib.pyplot as plt
+import scipy.misc
+import HTSeq 
+
+
+class ArgumentParserNoArgHelp(argparse.ArgumentParser):
+    """Like *argparse.ArgumentParser*, but prints help when no arguments."""
+    def error(self, message):
+        """Prints error message, then help."""
+        sys.stderr.write('error: %s\n\n' % message)
+        self.print_help()
+        sys.exit(2)
+
+
+def parseArgs():
+    """Parse arguments for script."""
+    parser = ArgumentParserNoArgHelp(description="Annotate synonymous barcodes "
+            "on influenza reads and make some plots.")
+    parser.add_argument('--bamfiles', help='BAM file generated by cellranger',
+            required=True, nargs='+')
+    parser.add_argument('--flugenes', help='FASTA file with flu genes',
+            required=True)
+    parser.add_argument('--synflugenes', help='FASTA file with synonymously '
+            'barcoded flu genes', required=True)
+    parser.add_argument('--cellbarcodes', help='Cell barcodes file created '
+            'by cellranger', required=True)
+    parser.add_argument('--outprefix', help='Prefix for output files.',
+            required=True)
+    parser.add_argument('--samplename', help='Name of sample, used as title for some plots', required=True)
+    return parser
+
+
+def main():
+    """Main body of script."""
+    parser = parseArgs()
+    args = vars(parser.parse_args())
+    bamfiles = args['bamfiles']
+    genefile = args['flugenes']
+    syngenefile = args['synflugenes']
+    cellbarcodesfile = args['cellbarcodes']
+    outprefix = args['outprefix']
+
+    print("Annotating synonymously barcoded reads.\n")
+
+    # read genes and synonymous variants
+    genes = dict([(seq.name, str(seq.seq)) for seq in Bio.SeqIO.parse(
+            genefile, 'fasta')])
+    genenames = [seq.name for seq in Bio.SeqIO.parse(genefile, 'fasta')]
+    print("Read {0} genes from {1}".format(len(genes), genefile))
+    syngenes = dict([(seq.name, str(seq.seq)) for seq in Bio.SeqIO.parse(
+            syngenefile, 'fasta')])
[…]

### 2017-06-04 monocle_analysis.Rmd
Plot total mRNAs for infected and un-infected cells

@@
 
-## Create and filter cell-gene dataset
+## Create cell-gene dataset and examine some basic statistics about the number of influenza and cellular reads per cell
 Read in cell-gene data of all samples.
@@
 Filter on just the samples that interest us for this analysis.
+Right now this is all the samples in the cell-gene dataset.
 ```{r}
-samples = c("Uninfected", "8hr-2", "8hr-2repeat", "10hr-2")
+samples = c("Uninfected", "6hr-1", "8hr-2", "8hr-2repeat", "10hr-2")
 
@@
 Compute the total number of mRNAs, cellular mRNAs, and influenza mRNAs per cell.
+Also compute the fraction of total mRNAs that are from influenza.
 ```{r}
@@
 
-flugenes = c("fluPB2", "fluPB1", "fluPA", "fluHA", "fluNP", "fluNA", "fluM", "fluNS")
-flu_id = row.names(subset(fData(cells), gene_short_name %in% flugenes))
-pData(cells)$flu_mRNAs <- Matrix::colSums(exprs(cells[flu_id,]))
+flugenes <- c("fluPB2", "fluPB1", "fluPA", "fluHA", "fluNP", "fluNA", "fluM", "fluNS")
+flugenes_id <- row.names(subset(fData(cells), gene_short_name %in% flugenes))
+pData(cells)$flu_mRNAs <- Matrix::colSums(exprs(cells[flugenes_id,]))
 
-flugenes_id = row.names(subset(fData(cells), !(gene_short_name %in% flugenes)))
-pData(cells)$cellular_mRNAs <- Matrix::colSums(exprs(cells[flugenes_id,]))
+pData(cells)$frac_flu_mRNAs <- pData(cells)$flu_mRNAs / pData(cells)$total_mRNAs
+
+cellgenes_id <- row.names(subset(fData(cells), !(gene_short_name %in% flugenes)))
+pData(cells)$cellular_mRNAs <- Matrix::colSums(exprs(cells[cellgenes_id,]))
 ```
 
+Plot the number of influenza mRNAs versus the number of cellular RNAs as scatterplots.
+The blue curves show the overall density on the number of cellular mRNAs.
+Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
 ```{r}
@@
 ```
-The scatterplots above show the number of influenza mRNAs versus the number of cellular mRNAs for each sample. 
-The blue curves show the overall density on the number of cellular mRNAs.
-Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
+
 The `Monocle` vignette suggests possibly filtering out cells with far more / less mRNAs than most cells.
-I have **not** done that here for the reason that this will filter out quite a few of the influenza-infected cells.
-This fact is made more clear by this next plot:
+I have **not** done that for the reason that this will filter out quite a few of the influenza-infected cells.
+This fact is made more clear by this next plot, which shows that the cells that have greater than 1% of their reads derived from influenza tend to have more total mRNAs than other cells in the two samples with the greatest number of influenza reads
+I can think of several possible reasons for this trend: 
+  1. Cells infected with influenza are simply "bigger" or are stimulated to produce more mRNAs
+  2. Influenza infection preferentially leads to doublets due to cell clumping, perhaps mediated by hemagglutinin-based physical linkage of cells.
+In any case, it seems safer **not** to exclude the influenza infected cells as we'd then be preferentially excluded cells with high viral burden.
 ```{r}
-all_densityplot <- ggplot(pData(cells), aes(total_mRNAs)) + geom_density()
-flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 100))
-flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
-flugt100_densityplot + facet_wrap(~Sample)
+pData(cells)$frac_flu_gt_0.01 <- pData(cells)$frac_flu_mRNAs > 0.01
+
+ggplot(pData(cells[, row.names(subset(pData(cells), Sample %in% c("8hr-2", "10hr-2")))]), 
+        aes(total_mRNAs)) +
+    geom_density(aes(color=frac_flu_gt_0.01)) +
+    theme(axis.text.x=element_text(angle=90, vjust=0.5)) +
+    facet_wrap(~Sample)
+#flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 50))
+#flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
 ```

### 2017-06-05 monocle_analysis.Rmd
clean plots of basic flu vs cellular mRNAs

@@
 title: "Monocle analysis of influenza-infected cells"
+author: "Alistair Russel, Cole Trapnell, Jesse Bloom"
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
@@
 Load or install necessary packages.
-If `Monocle` is not already available, install it using `Bioconductor`.
-Note that Cole suggested that we might want to install the development version of GitHub; right now this does **not** do this.
+Note that Cole suggested that we might want to install the development version of [Monocle](http://cole-trapnell-lab.github.io/monocle-release/) from GitHub; right now this does **not** do this.
 ```{r, message=FALSE, warning=FALSE}
@@
 if (!require("cowplot")) install.packages("cowplot")
+if (!require("scales")) install.packages("scales")
+
+# scientific notation formatting: https://stackoverflow.com/a/24241954
+fancy_scientific <- function(l) {
+     l <- format(l, scientific=TRUE)
+     l <- gsub("^0e\\+00","0", l)
+     l <- gsub("^1e\\+00", "1", l)
+     l <- gsub("^(.*)e", "'\\1'e", l)
+     l <- gsub("e\\+","e", l)
+     l <- gsub("e", "%*%10^", l)
+     l <- gsub("^\'1\'\\%\\*\\%", "", l)
+     parse(text=l)
+}
 ```
@@
 
-## Create cell-gene dataset and examine some basic statistics about the number of influenza and cellular reads per cell
-Read in cell-gene data of all samples.
-This is the merged cell-gene data set constructed essentially from `cellranger` using the iPython notebook.
+## Create cell-gene dataset and examine basic statistics about number of influenza and cellular mRNAs per cell
+Read in cell-gene data for all samples.
+This is the merged cell-gene data set constructed with `cellranger` using the iPython notebook.
 The cells are annotated by the number of wildtype and synonymous-barcoded influenza UMIs called for each cell.
@@
 
-Filter on just the samples that interest us for this analysis.
+Filter on the samples that interest us.
 Right now this is all the samples in the cell-gene dataset.
@@
 
-Compute the total number of mRNAs, cellular mRNAs, and influenza mRNAs per cell.
-Also compute the fraction of total mRNAs that are from influenza.
+Compute the number of total mRNAs, cellular mRNAs, influenza mRNAs, and fraction of influenza mRNAs per cell.
 ```{r}
@@
 
-Plot the number of influenza mRNAs versus the number of cellular RNAs as scatterplots.
-The blue curves show the overall density on the number of cellular mRNAs.
-Note that these numbers are not equivalent across samples as the different samples were sequenced to different depths.
-```{r}
+Plot the number of influenza mRNAs versus the number of cellular RNAs.
+The blue curves show the overall density of the number of cellular mRNAs.
+Note that different samples were sequenced to somewhat different depths.
+```{r, fig.height=4, fig.width=6.5}
 scatterplot <- ggplot(pData(cells), aes(cellular_mRNAs, flu_mRNAs)) + geom_point()
@@
 
-scatterplot + densityplot + theme(axis.text.x=element_text(angle=90, vjust=0.5)) + facet_wrap(~Sample)
+scatterplot + densityplot + theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1)) + 
+    facet_wrap(~Sample) + xlab("cellular mRNAs per cell") + ylab("influenza mRNAs per cell")
 ```
 
-The `Monocle` vignette suggests possibly filtering out cells with far more / less mRNAs than most cells.
-I have **not** done that for the reason that this will filter out quite a few of the influenza-infected cells.
-This fact is made more clear by this next plot, which shows that the cells that have greater than 1% of their reads derived from influenza tend to have more total mRNAs than other cells in the two samples with the greatest number of influenza reads
-I can think of several possible reasons for this trend: 
-  1. Cells infected with influenza are simply "bigger" or are stimulated to produce more mRNAs
-  2. Influenza infection preferentially leads to doublets due to cell clumping, perhaps mediated by hemagglutinin-based physical linkage of cells.
-In any case, it seems safer **not** to exclude the influenza infected cells as we'd then be preferentially excluded cells with high viral burden.
-```{r}
-pData(cells)$frac_flu_gt_0.01 <- pData(cells)$frac_flu_mRNAs > 0.01
+The `Monocle` vignette suggests filtering cells with far more / less mRNAs than most cells.
+I have **not** done that because it filters out a lot of the influenza-infected cells.
+This fact is made more clear by the next plot, which shows that cells with >1% of their mRNAs from influenza tend to have more total mRNAs than other cells.
+I can think of several possible explanations:
+
+  1. Influenza preferentially infects bigger cells.
+
+  2. Influenza infection leads cells to produce more mRNAs.
+
+  2. Influenza infection preferentially causes "doublets"" due to cell clumping, perhaps mediated by hemagglutinin-based cross-linking of cells.
+
+In any case, it seems safer **not** to exclude the influenza-infected cells as we'd then be preferentially excluded cells with high viral burden.
+```{r, fig.height=2.75, fig.width=5.5}
+pData(cells)$frac_flu_0.01 <- ifelse(pData(cells)$frac_flu_mRNAs > 0.01, 
+    "more than 1%", "less than 1%")
 
 ggplot(pData(cells[, row.names(subset(pData(cells), Sample %in% c("8hr-2", "10hr-2")))]), 
-        aes(total_mRNAs)) +
-    geom_density(aes(color=frac_flu_gt_0.01)) +
-    theme(axis.text.x=element_text(angle=90, vjust=0.5)) +
+        aes(total_mRNAs, color=frac_flu_0.01, fill=frac_flu_0.01)) +
+    geom_density(alpha=0.2) +
+    theme(axis.text.x=element_text(angle=90, vjust=0.5, hjust=1)) +
+    scale_y_continuous(label=NULL, name="density of cells", breaks=NULL, limits=c(0, NA)) + 
+    scale_x_continuous(limits=c(0, NA)) +
+    xlab("total mRNAs per cell") + 
+    labs(color="mRNAs from\ninfluenza", fill="mRNAs from\ninfluenza") + 
     facet_wrap(~Sample)
-#flugt100cells_id <- row.names(subset(pData(cells), flu_mRNAs > 50))
-#flugt100_densityplot <- ggplot(pData(cells[,flugt100cells_id]), aes(total_mRNAs)) + geom_density()
 ```
 
-```{r}
-min_expr <- 5
-num_cells <- 5
-cells <- detectGenes(cells, min_expr=min_expr)
-expressed_genes <- row.names(subset(fData(cells), num_cells_expressed >= num_cells))
-cat("Filtering to the", length(expressed_genes), "genes expressed at >=", min_expr,
-      "transcripts in at least", num_cells, "cells.\n")
-cells <- cells[expressed_genes,]
+Here are some cumulative distribution plots showing the fraction of mRNAs derived from influenza across cells for each sample.
+For the 6 and 8 hour timepoints, less than a quarter of cells have any appreciable influenza mRNAs.
+For the 10 hour timepoint, about a quarter the cells have at least a tiny fraction of influenza mRNAs, but only about 10% have more than 0.1% of mRNAs from influenza.
+```{r, fig.width=5, fig.height=2.75}
+# some trickery around x minimum to avoid taking log of 0 while plotting x on log scale
+minx <- 1e-4 # minimum value on y-axis
+pData(cells)$log_frac_flu_plus <- (function(x){ifelse(x < minx, log10(minx), log10(x))}
+    )(pData(cells)$frac_flu_mRNAs)
 
-cat("Plotting number of RNAs per cell for each sample\n")
-print(table(pData(cells)$Sample))
-pData(cells)$Total_mRNAs <- Matrix::colSums(exprs(cells))
-qplot(Total_mRNAs, data=pData(cells), geom="density", color=Sample)
-
-cat("Filtering out 6hr sample since it has a different distribution of RNA per cell")
-cells <- cells[, row.names(subset(pData(cells), Sample != "6hr"))]
-print(table(pData(cells)$Sample))
+ggplot(pData(cells), aes(log_frac_flu_plus, color=Sample)) + 
+    stat_ecdf() + 
+    scale_x_continuous(labels=function(x){fancy_scientific(10**x)},
+        limits=c(log10(minx), 0), expand=c(0, 0), name="fraction mRNAs from influenza") +
+    scale_y_continuous(limits=c(0, 1), name="fraction cells with > this\nmany influenza mRNAs") 
 ```
-
-## Look at distribution of influenza genes.
-```{r}
-flu_id <- row.names(subset(fData(cells), substr(ShortGeneName, 1, 3) == "flu"))
-
-pData(cells)$Total_flu <- Matrix::colSums(exprs(cells[flu_id,]))
-qplot(Total_flu, data=pData(cells), color=Sample, log="x")
-
-cth <- newCellTypeHierarchy()
-cth <- addCellType(cth, "noFlu", classify_func=function(x) {colSums(x[flu_id,]) == 0})
-cth <- addCellType(cth, "highFlu", classify_func=function(x) {colSums(x[flu_id,]) >= 100}) 
-cells <- classifyCells(cells, cth)
-print(table(pData(cells)$Sample, pData(cells)$CellType))
-```

### 2017-06-11 monocle_analysis.Rmd
Made purity plot

@@
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
-output: html_document
+output: github_document
 ---
@@
 
-if (!require("ggExtra")) install.packages("ggExtra")
-if (!require("cowplot")) install.packages("cowplot")
-if (!require("scales")) install.packages("scales")
+if (!require("pacman")) install.packages("pacman")
+pacman::p_load("ggplot2", "ggthemes", "ggExtra", "cowplot", "scales", "reshape2", "dplyr", "magrittr", "rmarkdown")
 
-# scientific notation formatting: https://stackoverflow.com/a/24241954
-fancy_scientific <- function(l) {
-     l <- format(l, scientific=TRUE)
-     l <- gsub("^0e\\+00","0", l)
-     l <- gsub("^1e\\+00", "1", l)
-     l <- gsub("^(.*)e", "'\\1'e", l)
-     l <- gsub("e\\+","e", l)
-     l <- gsub("e", "%*%10^", l)
-     l <- gsub("^\'1\'\\%\\*\\%", "", l)
-     parse(text=l)
+# scientific notation formatting, based loosely on this: https://stackoverflow.com/a/24241954
+fancy_scientific <- function(x) {
+     x %>% format(scientific=TRUE) %>% gsub("^0e\\+00","0", .) %>% gsub("^1e\\+00", "1", .) %>%
+        gsub("^(.*)e", "'\\1'e", .) %>% gsub("e\\+","e", .) %>% gsub("e", "%*%10^", .) %>%
+        gsub("^\'1\'\\%\\*\\%", "", .) %>% parse(text=.)
 }
@@
 
-cellfilter <- row.names(subset(pData(all_cells), Sample %in% samples))
-cells <- all_cells[,cellfilter]
+cells <- all_cells[, row.names(subset(pData(all_cells), Sample %in% samples))]
 
@@
 ```{r, fig.height=4, fig.width=6.5}
-scatterplot <- ggplot(pData(cells), aes(cellular_mRNAs, flu_mRNAs)) + geom_point()
+scatterplot <- ggplot(
+        transform(pData(cells), Sample=factor(Sample, levels=samples)), # order by sample 
+        aes(cellular_mRNAs, flu_mRNAs)) + 
+    geom_point()
 
@@
 
-ggplot(pData(cells[, row.names(subset(pData(cells), Sample %in% c("8hr-2", "10hr-2")))]), 
+ggplot(
+        transform(pData(cells[, row.names(subset(pData(cells), Sample %in% c("8hr-2", "10hr-2")))]),
+            Sample=factor(Sample, samples)), # just keep two samples with most flu and order them 
         aes(total_mRNAs, color=frac_flu_0.01, fill=frac_flu_0.01)) +
@@
 minx <- 1e-4 # minimum value on y-axis
-pData(cells)$log_frac_flu_plus <- (function(x){ifelse(x < minx, log10(minx), log10(x))}
-    )(pData(cells)$frac_flu_mRNAs)
+pData(cells)$log_frac_flu_plus <- ifelse(pData(cells)$frac_flu_mRNAs < minx, log10(minx), log10(pData(cells)$frac_flu_mRNAs))
 
-ggplot(pData(cells), aes(log_frac_flu_plus, color=Sample)) + 
+ggplot(
+        transform(pData(cells), Sample=factor(Sample, samples)), # order by sample 
+        aes(log_frac_flu_plus, color=Sample)) + 
     stat_ecdf() + 
@@
 ```
+
+## Look at levels of influenza genes, using synonymous barcodes to detect co-infection and mRNA leakage
+Roughly half of the infecting influenza viruses had a synonymous barcode at the 3' end of each mRNA.
+If cells infected by a single virus (which we expect to be the most common case for infected cells since infection itself is rare), then they should generally have either all wildtype or all synonymously barcoded viruses.
+Looking for deviations from this expectation can identify both true co-infection and cells that just got some influenza mRNA by leakage, since such leakage is expected to introduce a mix of both wildtype and synonymously barcoded.
+Establishing when cells start to have influenza mRNA due to true infection versus simple leakage is important for figuring out the actual variation in influenza mRNA levels in truly infected cells.
+
+Note that we have counted influenza mRNAs from the standard 10X sequencing of the full transcriptome.
+However, we have also called the synonymous versus wildtype barcodes using some extra semi-specific PCR which is not included in the cell-gene count data due to the fact that it is biased towards some genes, but does provide extra depth on the barcode counts.
+We limit the analysis below to cells that meet the following criteria:
+
+  1. They have at least two influenza reads in the cell-gene matrix.
+
+  2. They have at least two called barcodes (synonymous or wildtype) in the cell-gene matrix plus extra semi-specific PCR.
+
+These criteria are a way to limit the analysis just to cells where there is actual information about the synonymous barcodes.
+First, we filter down to these cells and plot some information about them.
+Note that these plots show the number of influenza mRNAs on a log scale, and then color the bars by the fraction of reads from each barcode.
+So the heights of each color in the bars are proportional to the fraction of the reads with that barcode, they are **not** the log of the number of reads with each barcode.
+```{r, warning=FALSE}
+# filter on sufficiently barcoded cells and samples with such cells
+bc_cells <- filter(pData(cells), flu_mRNAs >= 2, flu.wt + flu.syn >= 2)
+samples_to_plot <- bc_cells[["Sample"]] %>% unlist %>% unique
+
+# make variables with product of log flu mRNAs times fraction with each barcode
+bc_cells <- bc_cells %>% mutate(frac_wt=(flu.wt / (flu.wt + flu.syn))) %>% 
+    mutate(wildtype=log10(flu_mRNAs) * frac_wt, synonymous=log10(flu_mRNAs) * (1 - frac_wt)) %>%
+    mutate(major=pmax(wildtype, synonymous), minor=pmin(wildtype, synonymous))
+
+# make some dummy rows with 0 or many influenza to "pad" between samples on the plot
+npad <- 5 # pad this many cells on each side of each sample
+bc_cells <- lapply(samples_to_plot, function(s){data.frame(Sample=rep(s , 2 * npad), flu_mRNAs=c(rep(0, npad), rep(1e8, npad)))}) %>% 
+    bind_rows(., bc_cells) %>% mutate_each(funs(replace(., is.na(.), 0)))
+
+# order bc_cells by sample and flu abundance
+bc_cells$Sample <- factor(bc_cells$Sample, samples_to_plot)
+bc_cells <- arrange(bc_cells, Sample, flu_mRNAs)
+bc_cells$indx <- as.numeric(row.names(bc_cells))
+
+# get min and max indx for each sample
+indxrange <- bc_cells %>% filter(major > 0) %>% group_by(Sample) %>% summarize(imax=max(indx), imin=min(indx)) %>% 
+    mutate(center=(imin + imax) / 2, label=sprintf("%s\n(%d cells)", Sample, imax - imin + 1))
+
+#' plot purity of barcoded cells
+#'
+#' @param barcodes type of barcodes to plot, \code{c("major", "minor")} or \code{c("wildtype", "synonymous")}
+#' @return plot of purity
+purityplot <- function (barcodes) {
+    ggplot(melt(bc_cells, id.vars=c("indx", "Sample"), measure.vars=barcodes, variable.name="barcode"), 
+        aes(x=indx, y=value)) +
+    geom_bar(aes(fill=barcode), stat="identity", position="stack", width=1) +
+    scale_y_continuous(name="influenza mRNAs", expand=c(0, 0), labels=function(y){fancy_scientific(10**y)}) +
+    scale_x_continuous(expand=c(0, 0), name=NULL, breaks=indxrange$center, labels=indxrange$label) +
+    theme(text=element_text(size=13), axis.ticks.x=element_blank(), legend.position=c(0.01, 1.05), 
+          legend.justification=c(0, 1), legend.direction="horizontal", 
+          legend.background=element_rect(fill="white", size=0.5, linetype="solid", color="black")) +
+    scale_fill_discrete("") +
+    geom_rect(data=data.frame(xmin=indxrange$imin[seq(2, nrow(indxrange), 2)] - npad,
+            xmax=indxrange$imax[seq(2, nrow(indxrange), 2)] + npad, ymin=-Inf, ymax=Inf),
+        aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), alpha=0.1, inherit.aes=FALSE)
+}
+
+p_wt_syn <- purityplot(c("wildtype", "synonymous"))
+p_major_minor <- purityplot(c("major", "minor"))
+plot_grid(p_wt_syn + theme(axis.text.x=element_blank()), 
+    p_major_minor + scale_fill_manual(values=c("darkmagenta", "gold")), 
+    nrow=2, align='h', labels="AUTO", label_size=16, vjust=1)
+```

### 2017-07-26 paper/key_refs.tex
Added some refs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}

\begin{document}

\section{Key background references}
Here we list some key background references on some relevant points.

\subsection{Stochasticity during viral infection}
\begin{itemize}

\item \citet{delbruck1945burst} showed that the amount of progeny virions released from bacteria infected with bacteriophage differed by over 2 orders of magnitude.

\item \citet{zhu2009growth} showed that the number of progeny virions produced by mammalian cells infected with VSV differ by over 2 orders of magnitude. 
They further suggest that the differences are mostly due to cellular factors, since they are not heritable.

\item \citet{schulte2014single} examine single cells infected with poliovirus.
They found that the amount of both positive-sense and negative-sense genome and progeny virus differed by over an order of magnitude across cells.
Higher MOI was associated with more positive-sense genome but \emph{not} higher yields of progeny virus.

\item \citet{akpinar2016high} examine VSV infection of single cells both in the presence and absence of defective interfering particles (DIPs). 
They track both the amount of produced viral progeny and fluorescence of a reporter.
One interesting finding is that the viral yield from individual isolated cells is consistently lower than that from populations of cells, suggesting that presence of some cell-cell signaling or secreted factor might help viral production.
They find that adding more DIPs decreases both viral production and the expression of the reporter gene, although the former is decreased more.

\end{itemize}

\subsection{Influenza co-infection and defective particles}
\begin{itemize}

\item \citet{saira2013sequence} shows that defective influenza particles are present in infected humans.

\end{itemize}

\subsection{Interferon induction}
\begin{itemize}

\item \citet{shalek2013single} and \citet{shalek2014single} using single-cell RNA-seq to show that interferon induction is highly variable across individual cells in bone-marrow derived dendritic cells.
Some cells express lots of IFN related genes, others express almost none.
They suggest that some ``precocious'' cells are responsible for much of the IFN signaling.

\item \citet{bhushal2017cell} look at expression of IFN in single cell using an RFP reporter.
In epithelial cells, type I IFN induction is bimodal, with both the fraction of cells responding and the magnitude of response increasing with more IFNbeta, and saturating at near 100\% response.
In contrast, for type III IFN, although the response is also bimodal, it saturates at less than 100\%, and there is less dose-response in the magnitude.
They also did interesting experiments looking at ``memory'' in the response by separating responders and non-responders and then re-stimulating.
Cells that did not express IFN first time still had a fraction the responded the second time, whereas responders the first time usually although not always responded the second time.
So there is some but incomplete memory among responders.
They show that at least for type III IFN, the difference between responders and non-responders is not at the level of STAT1 activation, but is rather downstream and may involve histone acetylation as it is affected by HDAC inhibitors.
They also look in organoids and polarized cells, and find that in this more physiological setting the cells are more type III IFN responsive.

\item \citet{lopez2014defective} reviews the idea that defective genomes are important for IFN induction by many viruses including influenza.

\item \citet{baum2010preference} argues that RIG-I preferentially binds to short (defective) influenza segments.

\item \citet{tapia2013defective} shows that mice infected with influenza with more defective particles have more IFN induction.

\end{itemize}

\subsection{Transcriptional dynamics}

\begin{itemize}

\item \citet{hatada1989control} examine the accumulation of each influenza mRNA by Northern blot.
They find the same general hierarchy of expression as us: M and NS are the highest, then NP, then NA, then HA, then the three polymerases.
Specifically estimates that the polymerase proteins reach about $10^3$ copies / cell, HA gets to about $7 \times 10^3$ per cell, NA to about $10^4$ per cell, NP to about $1.5 \times 10^4$ per cell, NS to almost $2 \times 10^4$ / cell, and M to over $2 \times 10^4$ cell.

\item \citet{kawakami2011strand} use qPCR to look at the accumulation of NP and NA in infected cells.
They suggest that NP mRNA reaches a peak around 6 hours of $5 \times 10^4$ / cell, whereas NA reaches a much lower peak and after more time.

\item \citet{shapiro1987influenza} examines gene expression using Northern blots.
They find that the \emph{rate} of mRNA synthesis (at least for NP, M1, and NS1) generally peaks around 2.5 hours, with protein levels appearing to be near maximal by about 5.5 hours.
The rate of mRNA synthesis had fallen to 5\% of the maximal rate by 4.5 hours. 
Note that these times are for infection intiated \emph{after} absorption has been allowed to occur.

\end{itemize}

\nocite{*} % This command displays all refs in the bib file
\bibliography{references}

\end{document}

### 2017-08-08 paper/key_refs.tex

@@
 
+\item \citet{combe2015single} also look at the production of VSV from single infected cells, although immediately some of these cells are likely multiply infected. 
+They find that the yield of viral progeny differs by about two orders of magnitude across cells.
+
+\item \citet{heldt2015single} examine stochasticity during influenza virus infection.
+They show that the amount of infectious progeny virions produced per infection (at high MOI) ranged from one to almost 1000.
+They also show that the amount of vRNA per cell (again at high MOI infection) ranges by almost 3 orders of magnitude.
+The amount of vRNA and progeny virions is strongly correlated.
+For most segments, the amounts of different mRNAs is correlated.
+They also have a bunch of simulations that I looked at less closely.
+
+\item \citet{dou2017analysis} use an RNA-labeling method to look at the presence of vRNAs in single infected cells.
+By their method, they find that only about 20\% of cells contain all 8 vRNPs.
+They also show that there is a very limited window for co-infection.
+
 \end{itemize}
@@
 
+\item \citet{davis1980influenza} is an old paper that appears to be the original one showing that influenza defective RNAs have both termini and so represent some type of internal deletion.
+
+\item \citet{odagiri1997segment} look at a PA defective segment in the presence of an NS2 mutant.
+They show that the shorter defective PA segment is favored at both the levels of segment replication and genome packaging. 
+I remain a bit unclear about how the NS2 mutant plays into all of this, other than that apparently this NS2 mutant generates lots of PA defectives...
+
+\item \citet{hutchinson2008mutational} look at how many infected cells lack expression of at least one of two segments tested.
+They find that for wildtype virus this is about 20\%, but when there are synonymous mutations in the packaging signal of M, then it is about 50\%.
+
 \item \citet{saira2013sequence} shows that defective influenza particles are present in infected humans.
+
+\item \citet{brooke2013most} show that over half of cells infected with PR8 fail to express one of four proteins tested.
+High MOI infection alleviates this incomplete expression, presumably via co-infection with viruses each of which fails to express different genes.
+
+\item \citet{brooke2014influenza} find a mutation in NP that specifically reduces the packaging of NA in a way that is advantageous in guinea pigs.
+They do various things showing that co-infection occurs \textit{in vivo}, and also show that there is at least some amount of dose-response in terms of gene expression as cells that get lots of no-NA virus during co-infection express relatively less NA protein.
 
@@
 \begin{itemize}
+
+\item \citet{perez2014unbiased} looks at type I IFN induction in A549 cells with a fluorescent reporter.
+The wildtype virus induces IFN in a few percent of cells.
+A NS1-deletion virus induces in 15-20\% of cells.
+They passage in IFN-deficient cells to isolate various mutants that induce more than wildtype but less than the delta-NS1 virus.
+Most of these mutations are in proteins other than NS1, including some in M1 that were shown to induce IFN more effectively.
+They also argue that some of the inducing mutations lead to more defective segments to accumulate.
+
+\item \citet{killip2015influenza} reviews IFN induction by influenza, including the point that only a fraction of cells induce IFN upon infection.
+
+\item \citet{killip2017single} look at IFN induction by wildtype and NS1-defective influenza virus.
+Although some viruses (like Sendai) are able to induce IFN expression in almost all their cells, influenza only induces IFN in less than one percent of cells.
+An NS1-defective virus induced IFN in many more cells, but it was still only 5 to 20\%.
+In neither case is there an obvious correlation between NP protein production and IFN activation -- IFN is both present and absent in cells with little and lots of NP.
 
@@
 
+\citet{vreede2004model} This paper examines production of mRNA and cRNA in the presence of cycloheximide.
+They demonstrate the cRNA production requires the presence of the polymerase proteins (PB2, PB1, PA, NP) in excess of those in the initial vRNAs, although the PB1 can be inactive.
+
 \item \citet{kawakami2011strand} use qPCR to look at the accumulation of NP and NA in infected cells.
 They suggest that NP mRNA reaches a peak around 6 hours of $5 \times 10^4$ / cell, whereas NA reaches a much lower peak and after more time.
+The paper also shows that mRNA levels peak at 4-6 hours regardless of the presence or absence of cycloheximide, but the mRNA production is less in the presence of cycloheximide indicating secondary transcription.
 

### 2017-08-08 pymodules/utils.py
slight change to fix compatability issue with Python3

@@ -77,9 +77,8 @@ def mergeCellGeneMatrices(mergedgenes, mergedcells, mergedmatrix,
             elif (list(sample_annotations.columns.values[1 : ]) !=
                     annotations[2 : ]):
                 raise ValueError("Different annotations for different samples")
-            sample_annotations['cellbarcode'] = map(lambda bc: 
-                    '{0}-{1}'.format(bc.strip().split('-')[0], sample), 
-                    sample_annotations['cellbarcode'])
+            sample_annotations['cellbarcode'] = sample_annotations.apply(lambda row:
+                   '{0}-{1}'.format(row['cellbarcode'].strip().split('-')[0], sample), axis = 1)
             annotated_barcodes = []
             for (bc, sample) in map(lambda line: line.split('\t'),
                     sample_barcodes):

### 2017-08-10 paper/figures/Pseudovirus_flow_cytometry/Analysis.Rmd
Figures and data added to paper

---
title: "Flow_Analysis"
author: "Alistair Russell"
date: "7/10/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup}
# pipe operators
suppressPackageStartupMessages(library(magrittr))
# for manipulating factors
suppressPackageStartupMessages(library(forcats))
# tab data processing
suppressPackageStartupMessages(library(tidyverse))
# supress warnings
options(warn = -1)
# for low-level manipulation of ggplot
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggforce))
library(grid)
#loading flow cytometry data
suppressPackageStartupMessages(library(flowCore))
```

## Including Plots

You can also embed plots, for example:

```{r load data}
# look for files ending in .fcs
fcsfiles <- 'Data/HA' %>% 
    list.files(full.names = T, pattern = '.fcs$')

flowdata  <- fcsfiles %>% 
    lapply(. %>%
           # read FCS file without any transformations
           flowCore::read.FCS(transformation = F, alter.names = T) %>% 
           # extract counts alone
           flowCore::exprs() %>% 
           # convert to dplyr tibble
           as_tibble() 
          )

names(flowdata) <- fcsfiles %>% 
    # extract file name without .fcs extension and parent folders
    lapply(. %>% 
           stringr::str_match('([^/]+).fcs$') %>% 
           extract2(2))

# join all samples by 'file'
flowdata %<>% bind_rows(.id = 'file')

print(flowdata)
write.table(flowdata, 'coinfection.tsv', sep='\t', row.names=FALSE)
```

### 2017-08-25 paper/paper.tex
revisions to all sections of Results

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.

\usepackage{lipsum} % Required to insert dummy text
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\DeclareSIUnit\Molar{M}
\usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document

\renewcommand{\floatpagefraction}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Transcriptional dynamics of influenza virus infection in single cells}

\author[1]{Alistair B. Russell}
\author[2]{Cole Trapnell}
\author[1,2*]{Jesse D. Bloom}
\affil[1]{Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States}
\affil[2]{Department of Genome Sciences, University of Washington, Seattle, United States}
\corr{jbloom@fredhutch.org}{}

% \presentadd[\authfn{5}]{eLife Sciences editorial Office, eLife Sciences, Cambridge, United Kingdom}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

\begin{abstract}
Influenza virus infection induces large changes in cellular transcription.
Previously this has mostly been looked at using bulk measurements
Here we examine the process at the level of single cells.
We find extremely wide variation in the extent of viral gene transcription across infected cells.
IFN induction is very rare.
Some cellular pathways may be consistently altered in cells with high burden of viral transcripts.
Overall, highlights remarkable heterogeneity in the outcome of infection.
\end{abstract}

\section{Introduction}

Heterogeneity is important in a lot of cellular processes even when isogenic~\citep{shalek2013single,shalek2014single}.

Population (genetic) heterogeneity is also important. 
Viral quasispecies, cancer single-cell, etc.
Salmonella paper (PhoP).

Literature on viral burst-size heterogeneity.
This goes back to Delbruck, Andino polio paper~\citep{schulte2014single}, the MDCK / flu paper.

Discuss segmented nature of influenza.
Maybe in the context of how this could further increase heterogeneity because there is a lot of potential for entire genes to be missing.
Includes Yewdell and Lowen papers.

\section{Results}

\subsection{Strategy to measure mRNA in single virus-infected cells.}
We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription.
Each droplet is associated with a unique \emph{cell barcode} that tags all mRNAs from that cell during reverse-transcription.
A random \emph{unique molecular identifier (UMI)} is additionally appended to each mRNA molecule during reverse transcription.
The 3' ends of the mRNAs are sequenced and mapped to the human and influenza virus transcriptomes to determine transcript identities.
This information is combined with that provided by the UMIs and cell barcodes to quantify the number of molecules of each mRNA species that have been captured for each cell.

Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/Workflow/workflow.pdf}}
\caption{\label{fig:workflow} Experimental design.
{\bf (A)}  
We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
At top are the specific mutations for PB2.
At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
{\bf (B)} 
The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
{\bf (C)}  
A549 cells were infected with an equal mixture of mutant and wild-type virus. 
Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
The libraries were deep sequenced, and the data processed to create a matrix that gives the number of molecules of each transcript observed in each cell.
Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
}
\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.}
\end{figure}

We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes.
These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes.
To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{xue2016propagation}.
We then validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to both virion RNA (Figure~\ref{fig:viruspopulations}A) and particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
In addition, viral stocks with many defective particles are more immunostimulatory~\citep{tapia2013defective}.
We confirmed that our viral stocks induced much less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/validating_populations_D02.pdf}}
\caption{\label{fig:viruspopulations} The viral stocks in our experiments are relatively pure of defective particles. 
{\bf (A)}
Our viral stocks have a higher ratio of infectious particles to HA virion RNA compared to a high-defective stock propagated at high MOI.
HA viral RNA was quantified by qPCR on virions. 
{\bf (B)} 
Our viral stocks have a higher ratio of infectious particles to particles capable of expressing the viral HA protein.
A549 cells were infected with all three viral stocks at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
{\bf (C)} 
Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
MOIs were calculated by TCID50 on MDCK-SIAT1 cells.}
\figsupp[Flow cytometry data for calculating the percentage of HA-expressing cells.]{Longer caption}{\emph{HA-expression flow cytometry supplemental figure needs to be created.}}
\end{figure}

\subsection{Single cells show a wide range of expression of viral transcripts.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted close to half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}C).
%Therefore, there is extraordinary variability in the number of viral transcripts per infected cell.

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_cell_mRNA_summary.pdf}}
\caption{\label{fig:cells}
There is a very wide distribution in the number of viral mRNAs detected per cell.
{\bf (A)} 
Number of cells sequenced for each sample.
{\bf (B)} 
Cumulative distribution of the fraction of mRNA derived from influenza virus for each sample.
For each point on the x-axis, the y-axis indicates the fraction of cells with at least that much of their mRNA from virus.
{\bf (C)} 
The number of cellular and viral mRNAs for each cell is plotted as a point.
The blue lines show the overall distribution of the number of cellular mRNAs per cell.
Cells outside the red lines were considered outliers (possibly derived from two cells per droplet), and were excluded from subsequent analyses.
}
\end{figure}

A complicating factor is that uninfected cells could have small amounts of viral mRNA due to leakage of transcripts from lysed cells.
It is therefore important to establish a threshold for identifying truly infected cells.
We can do this by taking advantage of the fact that roughly half the infecting virions bear synonymous barcodes.
Reads derived from lysed cells will be drawn from both wild-type and synonymously barcoded viral transcripts.
However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
Figure~\ref{fig:viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and Figure~\ref{fig:viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection with both variants.

We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
Because the bulk of cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.

Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
In fact, a bare few infected cells contribute the majority of influenza transcripts.
Specifically, at 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_frac_flu_summary.pdf}}
\caption{\label{fig:viralbarcodes}
Synonymous barcodes on the viral mRNAs distinguish true infections from cells that contain viral mRNAs derived from leakage of lysed cells.
{\bf (A)}
Cells with at least two viral mRNAs for which the barcode could be called, arranged in order of increasing influenza transcript counts.
Bar heights denote the number viral mRNAs on a log\textsubscript{10} scale, bar coloring is linearly proportional to the fractions of viral mRNAs derived from wild-type and synonymously barcoded virus.
{\bf (B)}
Same as (A), but each bar is colored according to the relative fraction of the more common (major) and less common (minor) virus variant.
At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single particle.
A few cells are also obviously co-infected with both viral variants.
{\bf (C)}
We determined a cutoff for calling ``true'' infections by fitting a curve to the mean barcode purity of all cells with greater than a given fraction of their mRNA derived from virus.
We called the cutoff at the point at which purity stops increasing with the fraction of viral mRNA.
{\bf (D)}
The number of cells identified as infected and co-infected for each sample, as well as the number of cells with any viral read.
For all subsequent analyses, we subsampled the number of uninfected cells per sample to the greater of 50 or the number of infected cells.
{\bf (E)} 
Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
}
\figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
The number of viral barcodes called for each sample and influenza gene segment. 
Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
}{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.
}{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}
\end{figure}

\subsection{Absence of viral genes partially explains cell-to-cell variability in viral transcript abundance.}
The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{brooke2013most}. 
We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP).
Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins.
This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).

\begin{figure}[t!]
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
\caption{\label{fig:fluburdenbyflugene}
The absence of viral genes explains some of the variability in the amount of viral mRNA per cell.
{\bf (A)} 
Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
Cells with high viral burden always express all the RNP genes, but some cells with high viral burden lack each of the other viral genes.
{\bf (B)}
Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each viral gene.
A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$.
Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
{\bf (C)}
Fraction of infected cells that express all four RNP genes and also express each of the four other genes, as well as the fraction that express \emph{all} four of the other genes.
The fraction of cells that express all four other genes is well predicted by simply multiplying the frequencies of cells that express each of these genes individually, indicating that gene absence is approximately independent across these four genes.
}
\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cels in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide.
}{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
\figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
The absence of viral RNP genes but \emph{not} non-RNP genes remains significantly associated with reduced viral burden when we examine only the 10-hr sample, which is the single time point with the most data points.
The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
}{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
\end{figure}

We examined the total fraction of mRNA in each cell that derives from virus versus the relative expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A). 
Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, demonstrating that the presence of these RNP genes is essential for high levels of viral transcription.
However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any of the other viral genes is associated with reduced amounts of viral mRNA. 
However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express. 
The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.

We also sought to quantify the fraction of infected cells that completely failed to express a given gene.
We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C).
The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C). 
Our measurements of the frequency at which infected cells fail to express individual viral genes are roughly consistent with estimates made by others using different means~\citep{brooke2013most}.

\subsection{The relative amounts of different viral mRNAs are fairly consistent across cells.}
The results above show that the total amount of viral mRNA in infected cells varies over orders of magnitude.
Does the relative expression of viral genes exhibit similar cell-to-cell variability?
To address this question, we focused on cells that derived $>$5\% of their mRNA from virus, since estimates of relative viral gene expression will be less noisy in cells with more viral mRNAs.

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_aledit.pdf}}
\caption{\label{fig:fluexpr}
Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
{\bf (A)} 
The fraction of viral mRNA from each viral gene for each cell. 
{\bf (B)}
Box plots showing the distribution of the fraction of viral mRNA per cell from each viral gene.
The black lines at the notches are the medians, and the tops and bottoms of boxes indicate the first and third quartiles.
{\bf (C)}, {\bf (D)} 
The same plots, but only including cells for which we observed at least one molecule of each viral gene.
}
\figdata{The raw data are in \texttt{p\_flu\_expr.csv}.}
\end{figure}

In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
The cell-to-cell consistency in the relative expression of different viral genes generally becomes even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells.

\subsection{Co-infection can provide infected cells with the full complement of viral genes.}
Our sequencing enables us to identify the rare cells that were co-infected with both wild-type and synonymously barcoded viral variants.
Overall, we captured 10 such co-infected cells that had $>$5\% of their mRNA derived from virus (Figure~\ref{fig:coexpression}).
Seven of these 10 cells expressed all eight viral genes.
Remarkably, the majority (4 of 7) of these cells would \emph{not} have expressed all the viral genes in the absence of co-infection, since they have at least one gene exclusively derived from each viral variant.
For instance, the cell with 11.2\% of its mRNA from virus in the upper right of Figure~\ref{fig:coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{brooke2013most}.

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_coinfection.pdf}}
\caption{\label{fig:coexpression}
The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
The bars show the logarithm of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
}
\figsupp[Co-infected cells express roughly equal amounts of a gene from each infecting viral variant. \label{figsupp:flowcyto}]{
{\bf(A)} Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
At 10 hours post-infection, cells were analyzed by flow cytometry for HA and eGFP expression.
{\bf(B)} The expression of HA and GFP are correlated in co-infected cells. 
Shown are the quantile-normalized HA and eGFP signals for double-positive cells. 
Cells are colored by density, using a Gaussian kernel density estimate. 
{\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
\figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
\end{figure}

Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
This observation is consistent with the finding by \citet{dou2017analysis} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.

To support this idea with a larger dataset albeit at lower resolution, we generated a virus in which the HA coding sequence was replaced by GFP.
We then co-infected cells with a mix of wildtype and $\Delta$HA-GFP virus and used flow cytometry to score cells for the presence of HA only (infection by wildtype virus), GFP only (infection by $\Delta$HA-GFP virus), or both (co-infection) as shown in Figure~\ref{fig:coexpression}~Figure~supplement~\ref{figsupp:flowcyto}.
As in our single-cell sequencing data, we found that expression of HA and GFP were highly correlated, indicating that co-infected cells typically expressed roughly equal amounts of transcript from each viral variant.

\subsection{Activation of the anti-viral response is rare in single infected cells.}
Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
In particular, influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes.
In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection.
However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{shalek2013single,shalek2014single,bhushal2017cell}.
We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.

\begin{figure}
\centerline{\includegraphics[width=0.8\linewidth]{figures/p_small_tsne_merge.pdf}}
\caption{\label{fig:tsne}
A t-SNE plot created by semi-supervised clustering of the cells using genes that co-vary with viral infection status.
Each point is a single cell, and each panel shows an identical layout but colors the cells according to a different property.
{\bf (A), (B)}
Cells colored by the fraction of their mRNA that is derived from virus.
{\bf (C)}
Cells colored by the experimental sample from which they are derived.
{\bf (D), (E)}
Cells colored by the number of detected transcripts from type I and III interferons (IFN).
Only one cell has detectable interferon expression.
{\bf (E)}
Cells colored by the expression of the interferon-stimulated gene IFIT1.
{\bf (F)}
Cells colored by whether they express the viral NS gene.
The one interferon-positive cell is lacking NS, but many interferon-negative cells also lack NS.
}

\end{figure}

To examine whether there were distinct sub-populations of virus-infected cells, we used a semi-supervised t-SNE approach to cluster cells by genes that co-varied with viral infection status.
As shown in Figure~\ref{fig:tsne}A,B, this approach effectively grouped cells by the amount of viral mRNA that they expressed.
Sample-to-sample variation was regressed away during the clustering, as cells did not obviously group by time-point, with expected exception that the uninfected and 6-hour samples had few cells in the region of the plot corresponding to large amounts of viral mRNA (Figure~\ref{fig:tsne}C).

But to our surprise, we did not see a prominent clustering of infected cells into sub-populations as expected if the interferon response was strongly activated in some cells.
To investigate further, we annotated each cell by the total number of type I and III interferon transcripts detected.
Remarkably, only a single cell expressed detectable interferon (Figure~\ref{fig:tsne}D).
We also examined interferon-stimulated genes, which are induced by autocrine and paracrine interferon signaling.
Figure~\ref{fig:tsne}E shows expression of one such gene, IFIT1.
As with interferon itself, expression of IFIT1 was rare and most prominent in the single interferon-positive cell, presumably due to the higher efficiency of autocrine versus paracrine signaling.
Notably, interferon and interferon-stimulated genes were also relatively ineffective at blocking viral transcription in the single cell in which they were potently induced, since $>$10\% of the mRNA in this cell was derived from virus (Figure~\ref{fig:tsne}A,B,D,E).

We posited that the paucity of interferon induction might be due to the activity of influenza virus's major interferon antagonist, the NS1 protein.
We therefore identified cells that expressed substantial amounts of viral mRNA but lacked the NS gene (Figure~\ref{fig:tsne}F).
Consistent with the idea that NS1 is important for suppressing interferon, the one interferon-positive cell lacked detectable expression of the NS gene.
But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having $>$10\% of their mRNA derived from virus (Figure~\ref{fig:tsne}).
Therefore, most infected cells fail to express interferon even when the virus lacks its major known interferon antagonist.
Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{killip2017single}.
Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus clearly activate innate-immune responses much less frequently than might be presumed from bulk studies that flood cells with viral stocks full of defective particles.

\subsection{Genes involved in the oxidative-stress response co-vary with viral gene expression.}
We next examined whether any host genes were differentially expressed in cells with more viral mRNA.
We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of any effects due to the stochastic presence or absence of particular viral transcripts.
We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression along with the the amount of influenza virus mRNA in individual cells expressing all eight viral genes.
Shown are all genes that are differentially expressed at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_sig\_cellular\_genes.csv}.}
\figsupp[Many genes that co-vary with the amount of viral mRNA are involved in the response to oxidative stress. \label{figsupp:ROStable}]{
Table delineating genes in (Figure~\ref{fig:cellulargenes}) that are associated with the response to oxidative stress.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

Many of the genes that exhibit increased expression in cells with large amounts of viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}~Figure~supplement~\ref{figsupp:ROStable}). 
We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
Transient oxidative stress to occur during virus infection, and may act in a proviral fashion via MAPK activation driving vRNP export.
The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity.
Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.

\section{Discussion}
Still needs to be written.

\section{Methods and Materials}

\subsection{Design of synonymously barcoded virus.}

\subsection{Generation and characterization of viral stocks.}

\subsection{Infections and single-cell sequencing.}

\subsection{Data analysis}

\section{Acknowledgments}

Additional information can be given in the template, such as to not include funder information in the acknowledgments section.

\nocite{*} % This command displays all refs in the bib file
\bibliography{references}

\end{document}
ommand displays all refs in the bib file
\bibliography{references}

\end{document}

### 2017-08-25 paper/paper.tex
Changed Figure 1 to remove acronyms. Changed Figure 2A axis to reflect text. Added flow cytometry data for Figure 2B to supplement. Changed a few typosin text and changed title of subsection

@@
 MOIs were calculated by TCID50 on MDCK-SIAT1 cells.}
-\figsupp[Flow cytometry data for calculating the percentage of HA-expressing cells.]{Longer caption}{\emph{HA-expression flow cytometry supplemental figure needs to be created.}}
-\end{figure}
+\figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
+Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. All populations infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
+{\bf (A)} Uninfected gating control
+{\bf (B)} Cells infected with wild-type virus
+{\bf (C)} Cells infected with synonymously barcoded virus
+{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, therefore enriched in defective virus.
+}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}
 
@@
 We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
-Transient oxidative stress to occur during virus infection, and may act in a proviral fashion via MAPK activation driving vRNP export.
+Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export.
 The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity.

### 2017-08-30 paper/paper.tex
Materials and methods, first pass, added. Still need to add references and some of the data for materials and methods. Still determining scale on Monocle heatmap

@@
 Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
+Error bars $\pm$ S.D., n=3.
 {\bf (C)}  
@@
 HA viral RNA was quantified by qPCR on virions. 
+Error bars $\pm$ S.D., n=6 (qPCR replicates). 
 {\bf (B)} 
@@
 Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
-MOIs were calculated by TCID50 on MDCK-SIAT1 cells.}
+MOIs were calculated by TCID50 on MDCK-SIAT1 cells.
+Error bars $\pm$ S.D., n=3}
 \figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
@@
 Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
+Top and bottom of box bounds first and third quartiles of dataset, line dividing the box is at the median observation.
+Whiskers extend to the highest or lowest point within 1.5x the interquartile range, outliers shown as circles.
 {\bf (C)}
@@
 The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
-The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide.
+The bars show the relative amount of HA and PB2 mRNA in the absence versus the presence of cyclohexamide. Error $\pm$ S.D. n=3. 
 }{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
@@
 The black lines at the notches are the medians, and the tops and bottoms of boxes indicate the first and third quartiles.
+Whiskers extend to the highest or lowest data point observed within 1.5x the interquartile range, outliers shown as circles.
+Notches extend 1.58x the interquartile range divided by the square root of the number of observations. 
 {\bf (C)}, {\bf (D)} 
@@
 
-\subsection{Design of synonymously barcoded virus.}
+\subsection{Tissue culture and viral rescue}
+Cell lines used in this study were: the human lung epithelial carcinoma line, A549, obtained from ATCC, the canine kidney cell line overexpressing human SIAT1, MDCK-SIAT1, obtained from Sigma Aldrich, and the human embryonic kidney cell line, 293t, obtained from ATCC.
+All cells were maintained in D10 (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
+We generated viral stocks using a previously-defined reverse-genetics approach. 
+In brief, to generate wild-type virus we transfected bidirectional plasmids encoding the eight segments of A/WSN/1933 (kindly provided by Robert Webster of St.
+Jude Children's Research Hospital) into a coculture of 293t and MDCK-SIAT1 cells seeded at a ratio of 8:1.
+At 24 hours post-transfection, we changed media from D10 to influenza growth media (IGM Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
+At 48 hours post-transfection we harvested the virus-containing supernatant, removed cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at -80 \si{\degreeCelsius }. 
+We then determined the TCID50 of these stocks on MDCK-SIAT1 cells.
+Viral aliquots were then used to initiate an infection at an MOI of 0.01 on MDCK-SIAT1 cells in order to expand viral stocks and provide partial purging of defective viruses.
+Viral supernatant was harvested at 36 hours post-infection, prior to the onset of noticeable cytopathic effect. 
+Aliquots were again frozen at -80  \si{\degreeCelsius }, titered, and used in our single-cell experiments.
+Synonymously barcoded virus was generated identically to wild-type viral stocks, using plasmids encoding the synonymously-encoded viral segments described in (supplemental data).
+Mutant virus containing a gene encoding eGFP in place of the HA gene segment was generated using previously-described methods for growing HA-deficient virus, utilizing MDCK-SIAT1 cells expressing HA from a constituitive CMV promoter.
+A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21. 
+In order to obtain sufficient titers, HA-eGFP virus was expanded for 44 hours after initiating infection at an MOI of 0.01 rather than 36 hours.
 
-\subsection{Generation and characterization of viral stocks.}
+\subsection{qPCR measurements of infection kinetics}
+For qPCR measurements of viral transcriptional kinetics and interferon induction, A549 cells were seeded at 3 x 10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection.
+On the day of infection, a single well was trypsonized and enumerated, providing a measurement for MOI calculation.
+Immediately before infecting, D10 was replaced with IGM and the indicated treatment applied.
+For cells incubated with cyclohexamide, the compount was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production.
+RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's suggestions, and purity and concentration was measured using UV absorbance prior to cDNA synthesis. 
+cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
+Transcript abundance was measured using SYBR\textsuperscript{TM}  green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in (Supplemental table).
+Biological triplicates were performed for all samples.
+For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA using the manufacturer's protocol for random-hexamer priming.
+qPCR was then performed as for mRNA measurements.
+In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding bidirectional plasmid used in our viral rescue.
+All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.
 
-\subsection{Infections and single-cell sequencing.}
+\subsection{Flow cytometric analyses of infection}
+A549 cells were seeded in a 6-well plate and infected as described for qPCR analyses.
+Cells were harvested by trypsonization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10  \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study.
+After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
+Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
+After gating for single-cells in FlowJo, data were subsequently extracted using the R package, flowCore.
+Extracted data were then analyzed using a custom Python script, and gates determined.
+Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination.
 
-\subsection{Data analysis}
+\subsection{Generation of single cell infection data}
+All samples were processed using the 10x Chromium Single Cell 3' platform, both in initial, tagged, cDNA generation and in final library preparation.
+All time points, save for the 8 hour repeat, were performed on the same day as described below.
+A549 cells were seeded as described above in a 6-well plate, with two wells per time point. 
+A single well of cells was trypsonized and counted prior to initiation of the experiment for the purposes of calculating MOI.
+Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
+At the initiation of our experiment, all cells were changed from D10 to IGM, regardless of the desired time point.
+Cells were then infected with an intended, flow cytometry calculated, MOI of 0.3, in order of time point -- 10 hour, then 8 hour, then 6 hour time points, spaced two hours apart.
+This MOI was calculated in the absence of an additional washing step, and so potentially represents an overestimate of our final infectious dose.
+One hour post infection, media was changed to fresh IGM.
+All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media and normalizing potential batch effects across samples.
+For our replicate 8 hour time point, cells were infected as above, save for no wash step was included, the flow cytometry informed MOI was instead 0.1, and the sample was prepared on a different day.
+After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the guidlines provided by 10x genomics.
+All subsequent steps followed 10x genomics guidelines through library preparation.
+Samples were sequenced on an Illumina HiSeq.
+
+\subsection{Computational analysis of single cell transcriptomes}
+Sequencing data were aligned and a gene-cell-barcode matrix determined using the 10x Genomics software package, CellRanger v. 2.0.0. 
+Wild-type and synonymous viral transcripts were identified using a custom python script, and these annotations appended to the gene-cell-barcode matrix for further analysis.
+These matrices were then evaluated using the R package Monocle, version 2.4.0.
+Cell barcodes associated with 2.5-fold less or more UMI counts were considered to be potential doublets or inefficient/empty reactions, and were excluded from further downstream analyses.
+In order to empirically determine a cutoff for potential lysis-derived influenza reads we used the 10 hour time point, which had the highest influenza transcript count.
+The relative purity of barcodes was calculated out to the 35th percentile of influenza-associated reads.
+These values were plotted against the number of UMIs attributed to influenza, and a curve was fitted using the loess function in R. 
+The derivative of the resultant curve was calculated, and the threshold set at the minimum value wherein the derivative is less than 0 -- indicating the point below which purity drops indicating a significant contribution of lysis-derived rather than infection-derived reads.
+Cells meeting this threshold were defined as influneza-infected, and those failing to do so were excluded from further analysis.
+For each time point an additional number of cells, either 50 or the number of infected cells determined by our threshold above, whichever was larger, with no detectable influenza transcripts were also included for further analysis.
+
+Using Monocle?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influeza. 
+Influenza-infected cells were further described to possess <2\%, >2\%, and >20\% of reads derived from influenza using nested cell classes. 
+Candidate marker genes for t-distributed stochastic neighbor embedding (tSNE) dimensionality reduction that would allow us to seperate these populations were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
+The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
+
+Analyses of host expression required alteration to the standard Monocle pipeline and subsetting of our dataset.
+First, we only considered host genes in our analyses, as the massive contribution of viral variation could swamp out more nuanced effects.
+Second, as influenza has the capacity to suppress the host transcriptome, and itself ?takes up?
+UMIs that are attributed to a cell, we decided to assign the cell?s size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone.
+We excluded uninfected cells from these analyses, as well as those influenza-infected cells that lacked a gene from any of the eight segments. 
+Lastly, cells which contained a fraction of influenza-derived reads of greater than 25\% were assessed at a value of 25\% such that they did not contribute an undue weight to our analysis. 
+Genes whose expression covaried with the fraction of mRNAs in a cell derived from influenza were then identified, using the Monocle function differentialGeneTest, and removing variance explained by sample to sample variation.
+Retaining those meeting a q-value threshold of 0.1, the expression of these genes was visualized on a per-cell basis, arranging cells in order of increasing percentage of reads derived from influenza.
+
+As an additional analysis, we ranked cells within each sample according to the relative percentage of reads derived from influneza.
+As above, we only considered cells expressing genes from all eight viral genomic segments, and determined size factors based on the host transcriptome alone.
+Genes whose expression correlated with influenza ranking were then determined using the Monocel function deifferentialGeneTest, again removing variance explained by sample to sample variation and using a q-value threshold of 0.1.
+Cells were then arranged my ranking within each sample, in order of increasing percentage of reads derived from influenza within a sample, and differentially-regulated genes were visualized.
 

### 2017-08-30 paper/paper.tex
Fixed a few typos and sentence issues

@@
 These matrices were then evaluated using the R package Monocle, version 2.4.0.
-Cell barcodes associated with 2.5-fold less or more UMI counts were considered to be potential doublets or inefficient/empty reactions, and were excluded from further downstream analyses.
+Cell barcodes associated with 2.5-fold fewer or more UMI counts than the sample median were considered to be potential doublets or inefficient/empty reactions, and were excluded from further downstream analyses.
 In order to empirically determine a cutoff for potential lysis-derived influenza reads we used the 10 hour time point, which had the highest influenza transcript count.
@@
 These values were plotted against the number of UMIs attributed to influenza, and a curve was fitted using the loess function in R. 
-The derivative of the resultant curve was calculated, and the threshold set at the minimum value wherein the derivative is less than 0 -- indicating the point below which purity drops indicating a significant contribution of lysis-derived rather than infection-derived reads.
-Cells meeting this threshold were defined as influneza-infected, and those failing to do so were excluded from further analysis.
+The derivative of the resultant curve was calculated, and the threshold set at the minimum value wherein the derivative is less than 0 -- indicating the point below which there is a sacrifice in sample purity, indicating an increase in the relative proportion of extracellular-derived influenza reads.
+Cells meeting this threshold were defined as infected, and those failing to do so were excluded from further analysis.
 For each time point an additional number of cells, either 50 or the number of infected cells determined by our threshold above, whichever was larger, with no detectable influenza transcripts were also included for further analysis.
 
-Using Monocle?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influeza. 
-Influenza-infected cells were further described to possess <2\%, >2\%, and >20\% of reads derived from influenza using nested cell classes. 
-Candidate marker genes for t-distributed stochastic neighbor embedding (tSNE) dimensionality reduction that would allow us to seperate these populations were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
+Using Monocle?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influenza. 
+Influenza infected cells were further described to possess <2\%, >2\%, and >20\% of reads derived from influenza using nested cell classes. 
+Candidate marker genes for t-distributed stochastic neighbor embedding (tSNE) dimensionality reduction that would allow us to separate these populations were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
 The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
@@
 First, we only considered host genes in our analyses, as the massive contribution of viral variation could swamp out more nuanced effects.
-Second, as influenza has the capacity to suppress the host transcriptome, and itself ?takes up?
-UMIs that are attributed to a cell, we decided to assign the cell?s size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone.
+Second, as influenza has the capacity to suppress the host transcriptome, and moreover contributes significantly to the total number UMIs that are attributed to a cell, we decided to assign the cell?s size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone.
 We excluded uninfected cells from these analyses, as well as those influenza-infected cells that lacked a gene from any of the eight segments. 
 Lastly, cells which contained a fraction of influenza-derived reads of greater than 25\% were assessed at a value of 25\% such that they did not contribute an undue weight to our analysis. 
-Genes whose expression covaried with the fraction of mRNAs in a cell derived from influenza were then identified, using the Monocle function differentialGeneTest, and removing variance explained by sample to sample variation.
+Genes whose expression co-varied with the fraction of mRNAs in a cell derived from influenza were then identified, using the Monocle function differentialGeneTest, after removing variance explained by sample to sample variation.
 Retaining those meeting a q-value threshold of 0.1, the expression of these genes was visualized on a per-cell basis, arranging cells in order of increasing percentage of reads derived from influenza.
 
-As an additional analysis, we ranked cells within each sample according to the relative percentage of reads derived from influneza.
+As an additional analysis, we ranked cells within each sample according to the relative percentage of reads derived from influenza.
 As above, we only considered cells expressing genes from all eight viral genomic segments, and determined size factors based on the host transcriptome alone.
-Genes whose expression correlated with influenza ranking were then determined using the Monocel function deifferentialGeneTest, again removing variance explained by sample to sample variation and using a q-value threshold of 0.1.
+Genes whose expression correlated with influenza ranking were then determined using the Monocle function differentialGeneTest, again removing variance explained by sample to sample variation and using a q-value threshold of 0.1.
 Cells were then arranged my ranking within each sample, in order of increasing percentage of reads derived from influenza within a sample, and differentially-regulated genes were visualized.

### 2017-08-31 paper/paper.tex
Have added references through results section. Will start adding to methods next

@@
 \usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document
+\usepackage{natbib}
 
@@
 
-Heterogeneity is important in a lot of cellular processes even when isogenic~\citep{shalek2013single,shalek2014single}.
+Heterogeneity is important in a lot of cellular processes even when isogenic~\citep{Shalek:2014ey,Shalek:2013ej}.
 
@@
 Literature on viral burst-size heterogeneity.
-This goes back to Delbruck, Andino polio paper~\citep{schulte2014single}, the MDCK / flu paper.
+This goes back to Delbruck, Andino polio paper~\citep{Schulte:2014br}, the MDCK / flu paper.
 
@@
 \subsection{Strategy to measure mRNA in single virus-infected cells.}
-We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription.
+We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{Macosko:2015ht,Klein:2015ki}.
 Each droplet is associated with a unique \emph{cell barcode} that tags all mRNAs from that cell during reverse-transcription.
@@
 We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
-Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes.
-These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes.
-To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{xue2016propagation}.
+Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes ~\citep{Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
+These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes ~\citep{Lopez:2014en}.
+To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
 We then validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to both virion RNA (Figure~\ref{fig:viruspopulations}A) and particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
-In addition, viral stocks with many defective particles are more immunostimulatory~\citep{tapia2013defective}.
+In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf}.
 We confirmed that our viral stocks induced much less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
@@
 The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
-Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{brooke2013most}. 
+Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz}. 
 We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
 In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP).
-Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins.
+Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins~\citep{Vreede:2004ip}.
 This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).
@@
 The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C).
-Our measurements of the frequency at which infected cells fail to express individual viral genes are roughly consistent with estimates made by others using different means~\citep{brooke2013most}.
+Our measurements of the frequency at which infected cells fail to express individual viral genes are roughly consistent with estimates made by others using different means~\citep{Brooke:2013kb,Dou:2017cp}.
 
@@
 Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
-The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
+The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
 The cell-to-cell consistency in the relative expression of different viral genes generally becomes even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
@@
 For instance, the cell with 11.2\% of its mRNA from virus in the upper right of Figure~\ref{fig:coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
-Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{brooke2013most}.
+Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,Fonville:2015cg}.
 
@@
 Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
-This observation is consistent with the finding by \citet{dou2017analysis} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.
+This observation is consistent with the finding by \citet{Dou:2017cp} and \citet{Huang:2008gy} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.
 
@@
 Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
-In particular, influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes.
-In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection.
-However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{shalek2013single,shalek2014single,bhushal2017cell}.
+In particular, influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
+In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
+However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2014ey,Shalek:2013ej}.
 We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.
@@
 
-To examine whether there were distinct sub-populations of virus-infected cells, we used a semi-supervised t-SNE approach to cluster cells by genes that co-varied with viral infection status.
+To examine whether there were distinct sub-populations of virus-infected cells, we used a semi-supervised t-SNE approach to cluster cells by genes that co-varied with viral infection status \citep{Trapnell:2014kg,VanderMaaten:2008tm}.
 As shown in Figure~\ref{fig:tsne}A,B, this approach effectively grouped cells by the amount of viral mRNA that they expressed.
@@
 We also examined interferon-stimulated genes, which are induced by autocrine and paracrine interferon signaling.
-Figure~\ref{fig:tsne}E shows expression of one such gene, IFIT1.
+Figure~\ref{fig:tsne}E shows expression of one such gene, IFIT1~\citep{Fensterl:2011fp}.
 As with interferon itself, expression of IFIT1 was rare and most prominent in the single interferon-positive cell, presumably due to the higher efficiency of autocrine versus paracrine signaling.
@@
 Therefore, most infected cells fail to express interferon even when the virus lacks its major known interferon antagonist.
-Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{killip2017single}.
+Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
 Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus clearly activate innate-immune responses much less frequently than might be presumed from bulk studies that flood cells with viral stocks full of defective particles.
@@
 Shown are all genes that are differentially expressed at a false discovery rate of 0.1. 
-Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.}
+Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA. Scale is {\color{red} not sure what to put here} }
 \figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
@@
 \figsupp[Many genes that co-vary with the amount of viral mRNA are involved in the response to oxidative stress. \label{figsupp:ROStable}]{
-Table delineating genes in (Figure~\ref{fig:cellulargenes}) that are associated with the response to oxidative stress.
+Table delineating genes in (Figure~\ref{fig:cellulargenes}) that are associated with the response to oxidative stress ~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
 }{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
@@
 We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
-Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export.
-The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity.
+Transient oxidative stress is known to occur during viral infection, and may act in a proviral fashion via MAPK activation driving vRNP export~\citep{Amatore:2014cs}.
+The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
 Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
@@
 \end{document}
-ommand displays all refs in the bib file
-\bibliography{references}
 

### 2017-08-31 paper/paper.tex
added references to methods, fixed a few typos

@@
 All cells were maintained in D10 (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
-We generated viral stocks using a previously-defined reverse-genetics approach. 
+We generated viral stocks using a previously-defined reverse-genetics approach~\citep{Neumann:1999ws}. 
 In brief, to generate wild-type virus we transfected bidirectional plasmids encoding the eight segments of A/WSN/1933 (kindly provided by Robert Webster of St.
 Jude Children's Research Hospital) into a coculture of 293t and MDCK-SIAT1 cells seeded at a ratio of 8:1.
-At 24 hours post-transfection, we changed media from D10 to influenza growth media (IGM Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
+At 24 hours post-transfection, we changed media from D10 to influenza growth media (IGM, Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
 At 48 hours post-transfection we harvested the virus-containing supernatant, removed cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at -80 \si{\degreeCelsius }. 
@@
 Synonymously barcoded virus was generated identically to wild-type viral stocks, using plasmids encoding the synonymously-encoded viral segments described in (supplemental data).
-Mutant virus containing a gene encoding eGFP in place of the HA gene segment was generated using previously-described methods for growing HA-deficient virus, utilizing MDCK-SIAT1 cells expressing HA from a constituitive CMV promoter.
-A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21. 
+Mutant virus containing a gene encoding eGFP in place of the HA gene segment was generated using previously-described methods for growing HA-deficient virus, utilizing MDCK-SIAT1 cells expressing HA from a constituitive CMV promoter~\citep{Doud:2016gm}.
+A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21~\citep{Neumann:1999ws}. 
 In order to obtain sufficient titers, HA-eGFP virus was expanded for 44 hours after initiating infection at an MOI of 0.01 rather than 36 hours.
@@
 Immediately before infecting, D10 was replaced with IGM and the indicated treatment applied.
-For cells incubated with cyclohexamide, the compount was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production.
+For cells incubated with cycloheximide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
 RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's suggestions, and purity and concentration was measured using UV absorbance prior to cDNA synthesis. 
@@
 Biological triplicates were performed for all samples.
-For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA using the manufacturer's protocol for random-hexamer priming.
+For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
 qPCR was then performed as for mRNA measurements.
@@
 A549 cells were seeded in a 6-well plate and infected as described for qPCR analyses.
-Cells were harvested by trypsonization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10  \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study.
+Cells were harvested by trypsonization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10  \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
 After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
 Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
-After gating for single-cells in FlowJo, data were subsequently extracted using the R package, flowCore.
+After gating for single-cells in FlowJo, data were subsequently extracted using the R package, flowCore~\citep{LeMeur:2007uo}.
 Extracted data were then analyzed using a custom Python script, and gates determined.
-Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination.
+Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
 
@@
 For our replicate 8 hour time point, cells were infected as above, save for no wash step was included, the flow cytometry informed MOI was instead 0.1, and the sample was prepared on a different day.
-After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the guidlines provided by 10x genomics.
+After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the guidelines provided by 10x genomics.
 All subsequent steps followed 10x genomics guidelines through library preparation.
@@
 Wild-type and synonymous viral transcripts were identified using a custom python script, and these annotations appended to the gene-cell-barcode matrix for further analysis.
-These matrices were then evaluated using the R package Monocle, version 2.4.0.
+These matrices were then evaluated using the R package Monocle, version 2.4.0~\citep{Qiu:2017dw,Trapnell:2014kg}.
 Cell barcodes associated with 2.5-fold fewer or more UMI counts than the sample median were considered to be potential doublets or inefficient/empty reactions, and were excluded from further downstream analyses.
@@
 
-Using Monocle?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influenza. 
+Using Monocle'?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influenza. 
 Influenza infected cells were further described to possess <2\%, >2\%, and >20\% of reads derived from influenza using nested cell classes. 
@@
 First, we only considered host genes in our analyses, as the massive contribution of viral variation could swamp out more nuanced effects.
-Second, as influenza has the capacity to suppress the host transcriptome, and moreover contributes significantly to the total number UMIs that are attributed to a cell, we decided to assign the cell?s size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone.
+Second, as influenza has the capacity to suppress the host transcriptome ~\citep{BercovichKinori:2016iw}, and moreover contributes significantly to the total number UMIs that are attributed to a cell, we decided to assign the cell's size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone.
 We excluded uninfected cells from these analyses, as well as those influenza-infected cells that lacked a gene from any of the eight segments. 

### 2017-09-01 paper/paper.tex
a few minor citation additions, only had time to look at very beginning so far

--- 2017-08-31 paper/paper.tex
@@
 \subsection{Strategy to measure mRNA in single virus-infected cells.}
-We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{Macosko:2015ht,Klein:2015ki}.
+We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{zheng2017massively,Macosko:2015ht,Klein:2015ki}.
 Each droplet is associated with a unique \emph{cell barcode} that tags all mRNAs from that cell during reverse-transcription.
@@
 We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
-Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes ~\citep{Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
+Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes ~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
 These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes ~\citep{Lopez:2014en}.

### 2017-09-01 paper/paper.tex
small tweaks to results and references in it

@@
 \renewcommand{\floatpagefraction}{1}
+
+\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
 
@@
 Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
+\jdbcomment{Make the matrix more in line with the data that we actually see.
+Make the counts of the different host genes reasonably similar between cells.
+We think most host genes have pretty similar expression across cells, but your example shows all having at least 2-fold difference and some much larger.
+So adjust the numbers to be more similar among cells.
+Also, in A and the matrix in C, just write out ``wildtype'', ``synonymous'', and ``not detected'' (or ``none``) -- there is plenty of space, so no need to use acronyms.}
 }
@@
 We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
-Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes ~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
-These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes ~\citep{Lopez:2014en}.
+Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
+These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotype.
 To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
 We then validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to both virion RNA (Figure~\ref{fig:viruspopulations}A) and particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
-In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf}.
+In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
 We confirmed that our viral stocks induced much less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
@@
 Error bars $\pm$ S.D., n=6 (qPCR replicates). 
+\jdbcomment{Can you invert the y-axis on this plot so that it shows TCID50 / HA.}
 {\bf (B)} 
@@
 Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
-MOIs were calculated by TCID50 on MDCK-SIAT1 cells.
-Error bars $\pm$ S.D., n=3}
+Error bars $\pm$ S.D., n=3.
+Note that the MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
+}
 \figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
-Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. All populations infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
-{\bf (A)} Uninfected gating control
-{\bf (B)} Cells infected with wild-type virus
-{\bf (C)} Cells infected with synonymously barcoded virus
-{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, therefore enriched in defective virus.
+Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
+A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
+{\bf (A)} Uninfected gating control.
+{\bf (B)} Cells infected with the wild-type virus stock used in our experiments.
+{\bf (C)} Cells infected with synonymously barcoded virus stock used in our experiments.
+{\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
 }{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}
 
-\subsection{Single cells show a wide range of expression of viral transcripts.}
+\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
 We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
@@
 The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
-Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz}. 
+Indeed, several groups have reported that the majority of infected cells fail to express at least one viral gene~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}. 
 We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
-In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP).
-Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins~\citep{Vreede:2004ip}.
+In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\cite{huang1990determination}.
+Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
 This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).
@@
 Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each viral gene.
-A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$.
+A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
 Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
-Top and bottom of box bounds first and third quartiles of dataset, line dividing the box is at the median observation.
-Whiskers extend to the highest or lowest point within 1.5x the interquartile range, outliers shown as circles.
+Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
+Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
 {\bf (C)}
-Fraction of infected cells that express all four RNP genes and also express each of the four other genes, as well as the fraction that express \emph{all} four of the other genes.
-The fraction of cells that express all four other genes is well predicted by simply multiplying the frequencies of cells that express each of these genes individually, indicating that gene absence is approximately independent across these four genes.
+Among cells that express all four RNP genes, this panel shows the fraction of that express each of the four other genes, as well as the fraction that express \emph{all} four of the other genes.
+The fraction that express all four other genes is well predicted by simply multiplying the frequencies of cells that express each of these genes individually, indicating that gene absence is approximately independent across these four genes.
 }
@@
 
-We examined the total fraction of mRNA in each cell that derives from virus versus the relative expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A). 
-Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, demonstrating that the presence of these RNP genes is essential for high levels of viral transcription.
+We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A). 
+Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
 However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
-These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any of the other viral genes is associated with reduced amounts of viral mRNA.
+These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA.
 However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
@@
 Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
-This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells.
+This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells \jdbcomment{What is the reference here?}.
 
@@
 For instance, the cell with 11.2\% of its mRNA from virus in the upper right of Figure~\ref{fig:coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
-Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,Fonville:2015cg}.
+Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,Fonville:2015cg,aguilera2017plaques}.
 
@@
 The abundance of each viral transcript in cells that are co-infected with the two viral variants and have $>$5\% of their mRNA derived from virus.
-The bars show the logarithm of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
+The bars show the logarithms of the numbers of each viral mRNA detected, and are colored in linear proportion to the fraction of that mRNAs derived from wild-type or synonymously barcoded virus.
 }
@@
 
-\subsection{Activation of the anti-viral response is rare in single infected cells.}
+\subsection{Activation of the interferon response is rare in single infected cells.}
 Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
@@
 In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
-However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2014ey,Shalek:2013ej}.
+However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2014ey,Shalek:2013ej,PerezCidoncha:2014jr,bhushal2017cell}.
 We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.
@@
 
-To examine whether there were distinct sub-populations of virus-infected cells, we used a semi-supervised t-SNE approach to cluster cells by genes that co-varied with viral infection status \citep{Trapnell:2014kg,VanderMaaten:2008tm}.
+To examine whether there were distinct sub-populations of virus-infected cells, we used a semi-supervised t-SNE approach~\citep{VanderMaaten:2008tm} to cluster cells by genes that co-varied with viral infection status.
 As shown in Figure~\ref{fig:tsne}A,B, this approach effectively grouped cells by the amount of viral mRNA that they expressed.
@@
 \subsection{Genes involved in the oxidative-stress response co-vary with viral gene expression.}
-We next examined whether any host genes were differentially expressed in cells with more viral mRNA.
+We examined whether any host genes were differentially expressed in cells with more viral mRNA.
 We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of any effects due to the stochastic presence or absence of particular viral transcripts.
@@
 Shown are all genes that are differentially expressed at a false discovery rate of 0.1. 
-Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA. Scale is {\color{red} not sure what to put here} }
+Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA. 
+\jdbcomment{scale shows?} }
 \figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
@@
 Many of the genes that exhibit increased expression in cells with large amounts of viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
-These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}~Figure~supplement~\ref{figsupp:ROStable}).
+These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}).
 We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 

### 2017-09-01 paper/paper.tex
edits to first section of Methods

@@
 }
-\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.}
+\figdata{Sequences of wild-type and barcoded viruses are in \texttt{viralsequences.fasta}.\label{figdata:viralsequences}}
 \end{figure}
@@
 \figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
+\figdata{The sequence of the HA viral RNA carrying the GFP gene is in \texttt{HAflank-eGFP.fasta}.\label{figdata:HAflank}}
 \end{figure}
@@
 
-\subsection{Tissue culture and viral rescue}
-Cell lines used in this study were: the human lung epithelial carcinoma line, A549, obtained from ATCC, the canine kidney cell line overexpressing human SIAT1, MDCK-SIAT1, obtained from Sigma Aldrich, and the human embryonic kidney cell line, 293t, obtained from ATCC.
-All cells were maintained in D10 (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
-We generated viral stocks using a previously-defined reverse-genetics approach~\citep{Neumann:1999ws}. 
-In brief, to generate wild-type virus we transfected bidirectional plasmids encoding the eight segments of A/WSN/1933 (kindly provided by Robert Webster of St.
-Jude Children's Research Hospital) into a coculture of 293t and MDCK-SIAT1 cells seeded at a ratio of 8:1.
-At 24 hours post-transfection, we changed media from D10 to influenza growth media (IGM, Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
-At 48 hours post-transfection we harvested the virus-containing supernatant, removed cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at -80 \si{\degreeCelsius }. 
-We then determined the TCID50 of these stocks on MDCK-SIAT1 cells.
-Viral aliquots were then used to initiate an infection at an MOI of 0.01 on MDCK-SIAT1 cells in order to expand viral stocks and provide partial purging of defective viruses.
-Viral supernatant was harvested at 36 hours post-infection, prior to the onset of noticeable cytopathic effect. 
-Aliquots were again frozen at -80  \si{\degreeCelsius }, titered, and used in our single-cell experiments.
-Synonymously barcoded virus was generated identically to wild-type viral stocks, using plasmids encoding the synonymously-encoded viral segments described in (supplemental data).
-Mutant virus containing a gene encoding eGFP in place of the HA gene segment was generated using previously-described methods for growing HA-deficient virus, utilizing MDCK-SIAT1 cells expressing HA from a constituitive CMV promoter~\citep{Doud:2016gm}.
+\subsection{Cell lines and viruses}
+The following cell lines were used in this study: the human lung epithelial carcinoma line A549 (ATCC CCL-185), the MDCK-SIAT1 variant of the Madin Darby canine kidney cell line overexpressing human SIAT1 (Sigma-Aldrich 05071502), and the human embryonic kidney cell line 293T (ATCC CRL-3216).
+All cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
+
+Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
+The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
+Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
+The synonymous mutations were chosen to introduce nucleotides that are common in naturally occurring H1N1 viruses to avoid unwanted attenuation.
+The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
+To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
+At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
+At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }.
+We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
+To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
+
+The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by \jdbcomment{Alistair, add this.}
+
+For the experiments in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, we created a virus that carried an HA gene segment in which the carried GFP in place of most of the HA coding sequence by following a scheme first described by \citet{marsh2007specific}.
+Briefly, we created a plasmid encoding a viral RNA with GFP in place of the HA coding sequence in the context of the pHH21~\citep{Neumann:1999ws} reverse-genetics plasmid, removing potential start codons upstream of the GFP (see Figure~\ref{fig:coexpression}-source~data~\ref{figdata:HAflank} for the sequence of the viral RNA)
+We then generated GFP-carrying virus by reverse-genetics in previously described cells constitutively expressing HA~\citep{Doud:2016gm}.
 A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21~\citep{Neumann:1999ws}. 
-In order to obtain sufficient titers, HA-eGFP virus was expanded for 44 hours after initiating infection at an MOI of 0.01 rather than 36 hours.
+To obtain sufficient titers, this HA-eGFP virus was expanded for 44 than than 36 hours after initiating infection at an MOI of 0.01.
 
@@
 On the day of infection, a single well was trypsonized and enumerated, providing a measurement for MOI calculation.
-Immediately before infecting, D10 was replaced with IGM and the indicated treatment applied.
+Immediately before infecting, D10 was replaced with influenza growth media and the indicated treatment applied.
 For cells incubated with cycloheximide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
@@
 Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
-At the initiation of our experiment, all cells were changed from D10 to IGM, regardless of the desired time point.
+At the initiation of our experiment, all cells were changed from D10 to influenza growth media, regardless of the desired time point.
 Cells were then infected with an intended, flow cytometry calculated, MOI of 0.3, in order of time point -- 10 hour, then 8 hour, then 6 hour time points, spaced two hours apart.
 This MOI was calculated in the absence of an additional washing step, and so potentially represents an overestimate of our final infectious dose.
-One hour post infection, media was changed to fresh IGM.
+One hour post infection, media was changed to fresh influenza growth media.
 All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media and normalizing potential batch effects across samples.

### 2017-09-01 paper/paper.tex
revision to all methods sections except last one

@@
 
-\subsection{qPCR measurements of infection kinetics}
-For qPCR measurements of viral transcriptional kinetics and interferon induction, A549 cells were seeded at 3 x 10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection.
-On the day of infection, a single well was trypsonized and enumerated, providing a measurement for MOI calculation.
-Immediately before infecting, D10 was replaced with influenza growth media and the indicated treatment applied.
-For cells incubated with cycloheximide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
-RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's suggestions, and purity and concentration was measured using UV absorbance prior to cDNA synthesis. 
+\subsection{qPCR}
+For the qPCR in Figure~\ref{fig:viruspopulations} and Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection.
+On the day of infection, a single well was trypsinized and the cells were counted in order to determine the appropriate amount of virus to use to achieve the intended MOI.
+Immediately before infection, D10 was replaced with influenza growth media.
+For cells incubated with cyclohexamide, the compound was added to a final concentration of 50  \si{\micro}g/ml at the time of infection -- previously confirmed to be sufficient to halt viral protein production~\citep{Killip:2014kz}.
+RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's instructions. 
 cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
-Transcript abundance was measured using SYBR\textsuperscript{TM}  green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in (Supplemental table).
+Transcript abundance was measured using SYBR\textsuperscript{TM} green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in \jdbcomment{just list the primers here, as there aren't that many of them}.
 Biological triplicates were performed for all samples.
-For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
+For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA \jdbcomment{what are these primers?
+If they are the same ones given above, indicate that.
+If there are from one of the references that are at the end of this sentence, then indicate that.} using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
 qPCR was then performed as for mRNA measurements.
-In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding bidirectional plasmid used in our viral rescue.
+In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding reverse genetics plasmid. 
 All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.
 
-\subsection{Flow cytometric analyses of infection}
-A549 cells were seeded in a 6-well plate and infected as described for qPCR analyses.
-Cells were harvested by trypsonization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10  \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
+\subsection{Flow cytometry titering and analyses}
+To determine viral titers in terms of HA-expressing units and for the flow cytometry in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
+Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
 After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
-Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
-After gating for single-cells in FlowJo, data were subsequently extracted using the R package, flowCore~\citep{LeMeur:2007uo}.
-Extracted data were then analyzed using a custom Python script, and gates determined.
+Cells were then washed, fixed in 1\% formaldehyde\jdbcomment{are you sure it is formaldehyde and not paramformaldehyde?}, and washed further before a final resuspension and analysis. 
+To determine HA-expressing units, we then determined the fraction of cells that were HA positive and calculated the HA-expressing units from these fractions assuming a Poisson infection process.
+
+For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating for single-cells in FlowJo\jdbcomment{What do you mean gating for single cells?
+The gates select populations of cells}, data were subsequently extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
 
-\subsection{Generation of single cell infection data}
-All samples were processed using the 10x Chromium Single Cell 3' platform, both in initial, tagged, cDNA generation and in final library preparation.
-All time points, save for the 8 hour repeat, were performed on the same day as described below.
-A549 cells were seeded as described above in a 6-well plate, with two wells per time point. 
-A single well of cells was trypsonized and counted prior to initiation of the experiment for the purposes of calculating MOI.
+\subsection{Infections for single-cell mRNA sequencing}
+All samples were processed using the 10x Chromium Single Cell 3' platform~\cite{zheng2017massively}, both in initial, tagged, cDNA generation and in final library preparation \jdbcomment{say the version of the reagents used -- v1 or v2?
+Also, in the preceding sentence, what do you mean by both the initial cDNA generation and in final library preparation?
+Can't you just say that you used 10X without specifying both?}.
+All time points except for the 8-hour repeat, were performed on the same day.
+
+For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
+A single well of cells was trypsinized and counted prior to initiation of the experiment for the purposes of calculating MOI.
 Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
-At the initiation of our experiment, all cells were changed from D10 to influenza growth media, regardless of the desired time point.
-Cells were then infected with an intended, flow cytometry calculated, MOI of 0.3, in order of time point -- 10 hour, then 8 hour, then 6 hour time points, spaced two hours apart.
-This MOI was calculated in the absence of an additional washing step, and so potentially represents an overestimate of our final infectious dose.
-One hour post infection, media was changed to fresh influenza growth media.
-All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media and normalizing potential batch effects across samples.
-For our replicate 8 hour time point, cells were infected as above, save for no wash step was included, the flow cytometry informed MOI was instead 0.1, and the sample was prepared on a different day.
-After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the guidelines provided by 10x genomics.
-All subsequent steps followed 10x genomics guidelines through library preparation.
+At the initiation of our experiment, the wells for all time points were changed from D10 to influenza growth media.
+Cells were then infected with 0.3 HA-expressing units of virus per cell (a determined by flow cytometry).
+The infections were performed in order of time point: first the 10-hour time point, then the 8-hour, and then the 6-hour time point.
+At one hour after infection, the media for each time point was changed to fresh influenza growth media.
+Note that the HA-expressing units were calculated without this additional washing step, and so likely represent an overestimate of our final infectious dose (consistent with the fact that fewer than 30\% of cells appear infected in the single-cell sequencing data).
+All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
+For our replicate 8-hour time point, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
+After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
+All subsequent steps through library preparation followed the manufacturer's protocol.
 Samples were sequenced on an Illumina HiSeq. 
+The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Jesse will upload and add this.}
 
-\subsection{Computational analysis of single cell transcriptomes}
+\subsection{Computational analysis of single-cell mRNA sequencing data}
 Sequencing data were aligned and a gene-cell-barcode matrix determined using the 10x Genomics software package, CellRanger v. 2.0.0. 

### 2017-09-01 paper/paper.tex
finished revision to Methods

@@
 \usepackage{natbib}
+\usepackage{float}
+
+\newfloat{suppfile}{thp}{lofsupfile}
+%\renewcommand{\thesuppfile}{Supplementary file \arabic{suppfile}}
+\floatname{suppfile}{Supplementary file}
 
@@
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Sequencing data were aligned and a gene-cell-barcode matrix determined using the 10x Genomics software package, CellRanger v. 2.0.0. 
-Wild-type and synonymous viral transcripts were identified using a custom python script, and these annotations appended to the gene-cell-barcode matrix for further analysis.
-These matrices were then evaluated using the R package Monocle, version 2.4.0~\citep{Qiu:2017dw,Trapnell:2014kg}.
-Cell barcodes associated with 2.5-fold fewer or more UMI counts than the sample median were considered to be potential doublets or inefficient/empty reactions, and were excluded from further downstream analyses.
-In order to empirically determine a cutoff for potential lysis-derived influenza reads we used the 10 hour time point, which had the highest influenza transcript count.
-The relative purity of barcodes was calculated out to the 35th percentile of influenza-associated reads.
-These values were plotted against the number of UMIs attributed to influenza, and a curve was fitted using the loess function in R. 
-The derivative of the resultant curve was calculated, and the threshold set at the minimum value wherein the derivative is less than 0 -- indicating the point below which there is a sacrifice in sample purity, indicating an increase in the relative proportion of extracellular-derived influenza reads.
-Cells meeting this threshold were defined as infected, and those failing to do so were excluded from further analysis.
-For each time point an additional number of cells, either 50 or the number of infected cells determined by our threshold above, whichever was larger, with no detectable influenza transcripts were also included for further analysis.
+Jupyter notebooks that perform all of the computational analyses are available at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
+If you are a reviewer who needs access, please contact the editor.}
 
-Using Monocle'?s cell hierarchy function, cells were binned into those with no influenza, and those with reads derived from influenza. 
-Influenza infected cells were further described to possess <2\%, >2\%, and >20\% of reads derived from influenza using nested cell classes. 
-Candidate marker genes for t-distributed stochastic neighbor embedding (tSNE) dimensionality reduction that would allow us to separate these populations were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
+Briefly, the raw deep sequencing data were processed using the 10X Genomics software package CellRanger (version 2.0.0).
+The reads were aligned to a concatenation of the human and influenza virus transcriptomes.
+The human transcriptome was generated by filtering genome assembly GRCh38 for protein coding genes defined in the GTF file GRCh38.87.
+The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}).
+CellRanger calls cells based on the number of observed cell barcodes, and creates a cell-gene matrix.
+We used custom Python code to annotate the cells in this matrix by the number of viral reads that could be assigned to the wildtype and synonymously barcoded virus.
+This annotated cell-gene matrix is in Supplementary~file~\ref{suppfile:cellgenematrix}.
+The Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/flu_single_cell/blob/master/align_and_annotate.ipynb} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
+If you are a reviewer who needs access, please contact the editor.}, and is also available as Supplementary~file~\ref{suppfile:alignandannotate}.
+
+The annotated cell-gene matrix was then further analyzed in R, primarily using the Monocle package (version 2.4.0)~\citep{Qiu:2017dw,Trapnell:2014kg}.
+The Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/flu_single_cell/blob/master/monocle_analysis.ipynb} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
+If you are a reviewer who needs access, please contact the editor.}, and is also available as Supplementary~file~\ref{suppfile:monocle}.
+
+Briefly, this analysis was performed as follows.
+For each sample, cell barcodes with 2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were considered to be potential doublets or inefficient/empty reactions, and were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}C).
+In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C).
+We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
+This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
+The 10-hour sample has the most viral mRNA, and is expected to have the most lysis-associated influenza reads.
+We therefore determined the threshold using just this sample (Figure~\ref{fig:viralbarcodes}C), and then applied this threshold to all samples.
+This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
+For subsequent analyses, we retained all infected cells and a subsample of uninfected cells (the greater of 50 or the number of infected cells for that sample).
+The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the t-SNE clustering and other downstream analyses.
+
+For the semi-supervised t-SNE clustering, we used Monocle's cell hierarchy function to bin cells into those with no viral mRNA, <2\% viral mRNA, between 2\% and 20\% viral mRNA, and >20\%. 
+Candidate marker genes t-SNE dimensionality reduction were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
 The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
 
-Analyses of host expression required alteration to the standard Monocle pipeline and subsetting of our dataset.
-First, we only considered host genes in our analyses, as the massive contribution of viral variation could swamp out more nuanced effects.
-Second, as influenza has the capacity to suppress the host transcriptome ~\citep{BercovichKinori:2016iw}, and moreover contributes significantly to the total number UMIs that are attributed to a cell, we decided to assign the cell's size factor, a scalar value representing efficiency of UMI capture, based on host transcripts alone.
-We excluded uninfected cells from these analyses, as well as those influenza-infected cells that lacked a gene from any of the eight segments. 
-Lastly, cells which contained a fraction of influenza-derived reads of greater than 25\% were assessed at a value of 25\% such that they did not contribute an undue weight to our analysis. 
-Genes whose expression co-varied with the fraction of mRNAs in a cell derived from influenza were then identified, using the Monocle function differentialGeneTest, after removing variance explained by sample to sample variation.
-Retaining those meeting a q-value threshold of 0.1, the expression of these genes was visualized on a per-cell basis, arranging cells in order of increasing percentage of reads derived from influenza.
-
-As an additional analysis, we ranked cells within each sample according to the relative percentage of reads derived from influenza.
-As above, we only considered cells expressing genes from all eight viral genomic segments, and determined size factors based on the host transcriptome alone.
-Genes whose expression correlated with influenza ranking were then determined using the Monocle function differentialGeneTest, again removing variance explained by sample to sample variation and using a q-value threshold of 0.1.
-Cells were then arranged my ranking within each sample, in order of increasing percentage of reads derived from influenza within a sample, and differentially-regulated genes were visualized.
+For the analyses of cellular genes that differed in expression as a function of the amount of viral mRNA, we only considered cells that expressed all 8 viral mRNAs to avoid effects driven simply by viral gene absence.
+We also only considered cellular genes in the differential gene analysis, since viral gene expression will tautologically co-vary with the amount of viral mRNA.
+Additionally, because influenza virus has the capacity to degrade or prevent the synthesis of host mRNAs~\citep{BercovichKinori:2016iw} and contributes significantly to the total number UMIs in some cells, we calculate size factors (a scalar value representing efficiency of UMI capture) based on cellular transcripts alone.
+Finally, we assigned all cells a ceiling fraction of mRNA from virus of 25\% so that a few extremely high-expressing cells did not dominate. 
+Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation.
+\label{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
 
@@
 
+\clearpage
+
+\begin{suppfile}
+\caption{\label{suppfile:cellgenematrix}
+The annotated cell-gene matrix in Matrix Market Format.}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{\label{suppfile:alignandannotate}
+Jupyter notebook that runs CellRanger to align and annotate the reads.
+The ZIP file also includes associated custom scripts run by this notebook.}
+\end{suppfile}
+
+\begin{suppfile}
+\caption{\label{suppfile:monocle}
+Jupyter notebook that analyzes the cell-gene matrix, primarily using Monocle.}
+\end{suppfile}
+
 \end{document}

### 2017-09-07 paper/paper.tex
Addressed comments and added rough Discussion section.

@@
 Infected cells were further annotated by whether their viral mRNAs derived from wild-type virus, synonymously barcoded virus, or both.
-\jdbcomment{Make the matrix more in line with the data that we actually see.
-Make the counts of the different host genes reasonably similar between cells.
-We think most host genes have pretty similar expression across cells, but your example shows all having at least 2-fold difference and some much larger.
-So adjust the numbers to be more similar among cells.
-Also, in A and the matrix in C, just write out ``wildtype'', ``synonymous'', and ``not detected'' (or ``none``) -- there is plenty of space, so no need to use acronyms.}
 }
@@
 Error bars $\pm$ S.D., n=6 (qPCR replicates). 
-\jdbcomment{Can you invert the y-axis on this plot so that it shows TCID50 / HA.}
 {\bf (B)} 
@@
 Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
-This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells \jdbcomment{What is the reference here?}.
+This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells ~\citep{Heldt:2015iz}.
 
@@
 {\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
-}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
+{\bf(F)} Gaussian kernel density estimate on HA positive cells from {\bf(C)}
+}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D03.png}}
 \figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
@@
 \end{figure} 
-
 Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
@@
 \section{Discussion}
-Still needs to be written.
+We have generated total transcriptome measurements of individual influenza-infected cells -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
+Many of our observations closely match those made by other methods, validating our approach and granting us confidence in exploring more nuanced features of our data.
+Overall we find that influenza replication is defined broadly by both significant stochasticity and significant robustness, and that a view of viral infection as either strictly deterministic or incredibly chaotic fails to accurately describe the infectious process.
+Moreover we found that stress, likely induced by oxidative damage, appears to define the cellular response to infection.
+
+One of the most striking features of stochasticity with respect to influenza infection is the absence of genes corresponding to a given genomic segment, observed both by ourselves and others~\citep{Brooke:2013kb,Dou:2017cp}
+While there is significant speculation, and some indirect evidence, regarding the step of viral replication at which these genes are lost, there currently exists no strong consensus as to its identity -- indeed there may be no single event wholly responsible the stochastic loss of viral gene expression ~\citep{Brooke:2017gm, Heldt:2015iz}
+Regardless it does not seem unreasonable to surmise that this is a consequence of possessing a segmented genome.
+If so, it can be reasonably concluded that at some phase of viral replication, whether packaging, delivery to the nucleus, production of replicative intermediates, or some as-yet unsurmised step, the components of the viral genome must act in at least a semi-independent fashion.
+A consequence of this stochasticity is a subpopulation of infected cells that express significantly lower levels of viral transcripts. 
+Consistent with prior population-level studies of transcription, we find that absence of any of the four viral gene products necessary to form the viral polymerase leads to a dramatic drop in the levels of mRNA transcripts detected~\citep{Vreede:2004ip,eisfeld2015centre}.
+Our results are in slight contrast to a prior flow cytometric analysis of protein expression that described a relatively minor drop in expression associated with absence of viral ribonucleoprotein~\citep{Brooke:2013kb}.
+This discrepancy could be explained by two possibilities. 
+The first is that mRNA levels are not wholly predictive of viral protein abundance. 
+The second is that flow cytometry depends on a user-defined dynamic range of sensitivity as well as positive and negative gates, and that the potential 2-3 log difference in population means was either not captured in the sensitivity of the instrument or that gates on positive populations excluded bone-fide positives.
+Indeed, gating on HA staining using a strict cutoff based on an uninfected population alone produces a curve that is consistent with a bimodal distribution of HA expression (Figure ~\ref{fig:coexpression} Figure Supplement ~\ref{figsupp:flowcyto}F).
+
+However, despite extensive heterogeneity, there seems to be robustness to the infectious lifecycle.
+For example, influenza, regardless of the viral burden, exhibits relatively constrained ratios between the viral transcripts.
+This suggests that the processes that are driving heterogeneity of influenza mRNA abundance act equivalently throughout the viral genome.
+Potential mechanisms include components such as binding and entry kinetics that act on a virion and thus on all segments packaged therein, limitations of transcription rate that are dependent upon the same pool of cellular resources, or potentially even feedback-dependent mechanisms controlling viral transcription.
+
+Perhaps even more important to influenza?s success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events.
+While non-infectious events might be considered metabolic ?waste?, they would, theoretically, not impact the growth of successful viral infections. 
+However, recognition of any virus, infectious or not, and production of interferon, would impact the entire viral population. 
+While we cannot extrapolate much from our single interferon-positive cell, it can be stated that interferon induction by influenza is rare in our dataset. 
+This is consistent with prior experiments using reporter cell lines,  which places the induction at ~1\% of cells with detectable influenza protein expression -- albeit this value is likely sensitive to differences in viral propagation and strain utilized ~\citep{Killip:2017ef, Chen:2010cr}.
+If the major influenza interferon antagonist, NS1, is absent in ~10-20\% of infections, and lowly expressed in those infections engaging in only primary transcription, we must posit that NS1 alone is insufficient to explain influenza evasion of innate immunity.
+Interestingly, if other segmented viruses also exhibit a pattern of stochastic segment loss, they too would need to overcome this hurdle and escape innate immunity through mechanisms relying on more than a gene product generated by a single segment.
+
+In addition to viral mRNA kinetics, our analysis also allows us to observe those features of the cellular transcriptome that covary with influenza transcription.
+Critically, our analysis does not require synchronization or infection with a high MOI, and thus avoids potential artifacts produced by either approach. 
+We can therefore see extreme changes in a small subset of cells within a larger population.
+Moreover, we can also observe how transcriptional changes partition between cells, and exclude dramatic changes such as interferon induction that, while important, can mask other interesting biologically-relevant responses.
+With this in mind, the predominant response associated with increased influenza mRNA transcript abundance appears to be the oxidative stress pathway. 
+Generally characterized as antiviral, it is possible that cellular stress pathways represent an additional arm of innate immunity -- detecting a cellular state consistent with a loss of homeostasis as resources are diverted to viral propagation.
+Intriguingly, one element of the response, cyclophilin A (\emph{CypA}), is both thought to suppress influenza replication through degradation of matrix protein as well as act as a proinflammatory signal ~\citep{Liu:2009er, Jin:2004ii}.
+It will be interesting to follow up on these observations and determine the broader mechanisms through which cellular stress pathways are involved in pro- or anti-viral responses.
+
+The success garnered here underscores that single-cell approaches seem distinctly suited to study of viral pathogens, as not only are many host responses governed by bimodality and stochasticity ~\citep{Shalek:2014ey,Shalek:2013ej}, but also viral populations themselves exhibit a staggering array of diversity.
+Our results with influenza have provided significant support for earlier analysis of viral stochasticity, further defined the sources and impacts of that stochasticity, and generated potentially fertile avenues of research.
+While it is early days yet, it is our hope that our efforts will contribute to a deeper understanding of viral populations, treating population-level measurements as emergent phenomena defined by underlying individual stochasticity rather than analyzing viral replication in a strictly deterministic framework.
 
@@
 To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
-
-The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by \jdbcomment{Alistair, add this.}
+The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours, when obvious cytopathic effect was noted.
 
@@
 cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
-Transcript abundance was measured using SYBR\textsuperscript{TM} green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the primers listed in \jdbcomment{just list the primers here, as there aren't that many of them}.
+Transcript abundance was measured using SYBR\textsuperscript{TM} green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the following primers: \emph{HA}: 5'-GGCCCAACCACACATTCAAC-3', 5'-GCTCATCACTGCTAGACGGG-3', \emph{IFNB1}: 5'-AAACTCATGAGCAGTCTGCA-3', 5'-AGGAGATCTTCAGTTTCGGAGG-3', \emph{L32}: 5'-AGCTCCCAAAAATAGACGCAC-3', 5'-TTCATAGCAGTAGGCACAAAGG-3'.
 Biological triplicates were performed for all samples.
-For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA \jdbcomment{what are these primers?
-If they are the same ones given above, indicate that.
-If there are from one of the references that are at the end of this sentence, then indicate that.} using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl,Hoffmann:2001vj}.
+For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA adapted from \citep{Hoffmann:2001vj}  using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl}.
 qPCR was then performed as for mRNA measurements.
@@
 After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
-Cells were then washed, fixed in 1\% formaldehyde\jdbcomment{are you sure it is formaldehyde and not paramformaldehyde?}, and washed further before a final resuspension and analysis. 
+Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
 To determine HA-expressing units, we then determined the fraction of cells that were HA positive and calculated the HA-expressing units from these fractions assuming a Poisson infection process.
 
-For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating for single-cells in FlowJo\jdbcomment{What do you mean gating for single cells?
-The gates select populations of cells}, data were subsequently extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
+For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating to exclude multiplets in FlowJo, data were subsequently extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
@@
 \subsection{Infections for single-cell mRNA sequencing}
-All samples were processed using the 10x Chromium Single Cell 3' platform~\cite{zheng2017massively}, both in initial, tagged, cDNA generation and in final library preparation \jdbcomment{say the version of the reagents used -- v1 or v2?
-Also, in the preceding sentence, what do you mean by both the initial cDNA generation and in final library preparation?
-Can't you just say that you used 10X without specifying both?}.
+Single cell sequencing libraries were generated using the 10x Chromium Single Cell 3' platform, V1~\cite{zheng2017massively}.
 All time points except for the 8-hour repeat, were performed on the same day.

### 2017-09-08 paper/paper.tex
Added rough introduction. Changed one part of the methods that was not quite right. Have made some changes to Discussion, but still need to add paragraph on flu variation

@@
 \section{Introduction}
+Recent advances in sequencing and visualization technologies have revealed that cell-to-cell heterogeneity is a fundamental feature of biological processes -- impacting events from bacterial entrance into quiescent states to organogenesis in multicellular organisms~\citep{JohnstonJr:2010jn}.
+Due to this heterogeneity, biological pathways generated from bulk population data may not accurately describe the timing, order, or even the identity of events within single cells ~\citep{Trapnell:2014kg}.
+Nearly no other feature of biological life is as defined by heterogeneity as viral infection, wherein events can be highly stochastic due to the incredible combination of both genetic and non-genetic sources of variation.
 
-Heterogeneity is important in a lot of cellular processes even when isogenic~\citep{Shalek:2014ey,Shalek:2013ej}.
+The contribution of heterogeneity to infectious processes has been well appreciated in the study of viral quasispecies, wherein pathogenesis is best described as an emergent process dependent upon underlying genetic diversity ~\citep{Lauring:2010if}.
+This concept has revolutionized the study of RNA viruses: including polio, hepatitis C, and influenza \citep{Echeverria:2015ew, Kuroda:2010gt}. 
+Direct evidence of the contribution of the quasispecies to pathogenesis has been observed -- for instance mutant strains of polio that produce less variation exhibit significantly less fitness in animal models of infection despite displaying wild-type replication kinetics \emph{in vitro}~\citep{Vignuzzi:2005ju}.
+However, these studies have largely been limited to population-level analyses, which can link standing diversity to complex biological phenomena but are unable to deconvolute precise events in individual infected cells.
 
-Population (genetic) heterogeneity is also important. 
-Viral quasispecies, cancer single-cell, etc.
-Salmonella paper (PhoP).
+Non-genetic variation has also been described to contribute to the host response to infection. 
+Single-cell transcriptomic analyses have revealed incredibly coordinated responses driven by stochasticity, even in tissue culture models using genetically homogeneous cells.
+This heterogeneity in response has been observed both with respect to pure ligands such as lipopolysaccharide, wherein differences must be host-cell intrinsic ~\citep{Shalek:2014ey,Shalek:2013ej,Rand:2012kl}, and with respect to pathogens, wherein differences are likely a complex mixture of host-cell intrinsic and pathogen-specific heterogeneity ~\citep{Avraham:2015cj}.
+Moreover, stochasticity is observed both in interferon production, which drives the anti-pathogen state, and the expression of the downstream effector molecules of innate immunity.
+As inflammation is a potentially deleterious process, restricting a response to a small subset of cells upon initial recognition of a threat might be particularly advantageous.
 
-Literature on viral burst-size heterogeneity.
-This goes back to Delbruck, Andino polio paper~\citep{Schulte:2014br}, the MDCK / flu paper.
+Influenza, a segmented negative sense RNA virus, places a significant infectious burden on the global population and yet, despite many decades of study, there is still significant conjecture regarding those features of viral populations that drive observed phenotypes.
+Cells infected with this virus exhibit a wide degree of heterogeneity, including a number of semi-infectious events wherein the full complement of viral gene products is absent~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}.
+This has broadly been observed to impact rates of recombination, however its contribution to other viral behaviors remains unclear~\citep{Fonville:2015cg}. 
+Influenza-infected cells are also observed to exhibit a wide-breadth of productivity -- exhibiting a 3-log range in infectious progeny even in high-MOI infections~\citep{Heldt:2015iz}. 
+This appears to be a general trend of many RNA viral infections, from phage to polio, and the features underlying these observations appear multitudinous~\citep{DELBRUCK:1945wo, Schulte:2014br}.
 
-Discuss segmented nature of influenza.
-Maybe in the context of how this could further increase heterogeneity because there is a lot of potential for entire genes to be missing.
-Includes Yewdell and Lowen papers.
+We now hope to apply new single-cell transcriptomic technologies to the study of influenza replication in a tissue culture model of infection.
+Early single-cell transcriptomic approaches were laborious -- requiring the generation of sequencing libraries on individual cells~\citep{Tang:2009kj,ld:2012gj}.
+Breakthroughs in microfluidics helped to automate this process, but retained strictly parallel processing of sequencing libraries.
+More recent technologies have introduced more complex means of barcoding and partitioning cells, allowing for the generation of libraries consisting of thousands of cells with ?single pot?
+downstream processing.
+These include the use of reverse-phase emulsions, gravity capture, and even \emph{in situ} reverse transcription after cellular partitioning~\citep{JunSSi,Gierahn:2017bc,Klein:2015ki,Macosko:2015ht}. 
+We hope to leverage these technological advances in the study of influenza, using transcriptomic analysis to understand the phenotypic heterogeneity that abounds during infection with this virus.
 
@@
 {\bf(C)},{\bf(D)},{\bf(E)} Gating controls, single infection with eGFP virus, single infection with wild-type virus, and uninfected cells, respectively. 
-{\bf(F)} Gaussian kernel density estimate on HA positive cells from {\bf(C)}
-}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D03.png}}
+}{\includegraphics[width=\linewidth]{figures/Pseudovirus_flow_cytometry/coinfFlow_D02.png}}
 \figdata{The raw data plotted in this figure are in \texttt{p\_co-infection.csv}.}
@@
 We have generated total transcriptome measurements of individual influenza-infected cells -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
-Many of our observations closely match those made by other methods, validating our approach and granting us confidence in exploring more nuanced features of our data.
+The dataset we have generated is largely consistent with prior flow cytometry-based analyses, allowing us to build upon those prior observations with a more holistic view, beginning to understand the consequences of heterogeneity in influenza.
 Overall we find that influenza replication is defined broadly by both significant stochasticity and significant robustness, and that a view of viral infection as either strictly deterministic or incredibly chaotic fails to accurately describe the infectious process.
@@
 If so, it can be reasonably concluded that at some phase of viral replication, whether packaging, delivery to the nucleus, production of replicative intermediates, or some as-yet unsurmised step, the components of the viral genome must act in at least a semi-independent fashion.
+
 A consequence of this stochasticity is a subpopulation of infected cells that express significantly lower levels of viral transcripts. 
@@
 Our results are in slight contrast to a prior flow cytometric analysis of protein expression that described a relatively minor drop in expression associated with absence of viral ribonucleoprotein~\citep{Brooke:2013kb}.
-This discrepancy could be explained by two possibilities. 
-The first is that mRNA levels are not wholly predictive of viral protein abundance. 
-The second is that flow cytometry depends on a user-defined dynamic range of sensitivity as well as positive and negative gates, and that the potential 2-3 log difference in population means was either not captured in the sensitivity of the instrument or that gates on positive populations excluded bone-fide positives.
-Indeed, gating on HA staining using a strict cutoff based on an uninfected population alone produces a curve that is consistent with a bimodal distribution of HA expression (Figure ~\ref{fig:coexpression} Figure Supplement ~\ref{figsupp:flowcyto}F).
+This discrepancy could be explained by two possibilities: the first is that mRNA measurements may not be wholly reflective of protein levels, and the second is that flow cytometry requires decisions regarding sensor sensitivity and positive gates and the potential biases that those decisions incur.
+While sequencing-based analysis comes with its own set of caveats, the detection threshold is 'baked-in' and so potentially less subject to bias.
 
@@
 
-Perhaps even more important to influenza?s success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events.
-While non-infectious events might be considered metabolic ?waste?, they would, theoretically, not impact the growth of successful viral infections. 
+Perhaps even more important to influenza's success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events.
+While non-infectious events might be considered metabolic 'waste', they would, theoretically, not impact the growth of successful viral infections. 
 However, recognition of any virus, infectious or not, and production of interferon, would impact the entire viral population. 
@@
 Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
-The synonymous mutations were chosen to introduce nucleotides that are common in naturally occurring H1N1 viruses to avoid unwanted attenuation.
 The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}

### 2017-09-09 paper/paper.tex
Additional paragraph devoted to diversity in discussion, final paragraph edited slightly

@@
 
+It is immediately apparent that a model of influenza infection wherein there is equivalent accumulation of transcripts across successfully infected cells is far to facile to describe the true situation.
+While infection certainly proceeds over time, with gradual increase in influenza transcript abundance, the level of transcripts expressed in cells varies wildly.
+While some of these differences are likely due to temporal variation in binding and internalization, it will be interesting in follow-up experiments to determine whether there is an ultimate level of influenza transcript that is invariably reached or whether the variation we observe is indicative of a final setpoint.
+Indeed, our analyses of host transcriptomes suggests that those cells expressing higher levels of influenza are distinguishable, even at 6 hours, from later time points expressing similar influenza burden, giving an indication that there are features above-and-beyond temporal variation driving differences in viral mRNA expression\jdbcomment{If we make this point we need to add the ranked graph to the results and address it} .
+
 One of the most striking features of stochasticity with respect to influenza infection is the absence of genes corresponding to a given genomic segment, observed both by ourselves and others~\citep{Brooke:2013kb,Dou:2017cp}
@@
 Our results with influenza have provided significant support for earlier analysis of viral stochasticity, further defined the sources and impacts of that stochasticity, and generated potentially fertile avenues of research.
-While it is early days yet, it is our hope that our efforts will contribute to a deeper understanding of viral populations, treating population-level measurements as emergent phenomena defined by underlying individual stochasticity rather than analyzing viral replication in a strictly deterministic framework.
+We stand now at an exciting moment in virology, as technology allows us to observe population-level measurements as emergent phenomena defined by underlying individual stochasticity rather than analyzing viral replication in a strictly deterministic framework.
+It is our hope that this work will serve not only as a window into influenza replication, but as a part of a grander design in re-examining and understanding the fundamental features of viral infection.
 

### 2017-09-12 paper/paper.tex
some tweaks to Results

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Transcriptional dynamics of influenza virus infection in single cells}
+\title{Extreme heterogeneity of influenza virus transcription in single cells}
 
@@
 We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{zheng2017massively,Macosko:2015ht,Klein:2015ki}.
-Each droplet is associated with a unique \emph{cell barcode} that tags all mRNAs from that cell during reverse-transcription.
-A random \emph{unique molecular identifier (UMI)} is additionally appended to each mRNA molecule during reverse transcription.
+Each droplet contains primers with a unique \emph{cell barcode} that tags all mRNAs from that droplet during reverse-transcription.
+Each primer also contains a \emph{unique molecular identifier (UMI)} that is appended to each mRNA molecule during reverse transcription.
 The 3' ends of the mRNAs are sequenced and mapped to the human and influenza virus transcriptomes to determine transcript identities.
@@
 {\bf (C)}  
-A549 cells were infected with an equal mixture of mutant and wild-type virus. 
+For the single-cell mRNA sequencing, A549 cells were infected with an equal mixture of mutant and wild-type virus. 
 Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
@@
 Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
-These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotype.
+These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes.
 To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
-We then validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to both virion RNA (Figure~\ref{fig:viruspopulations}A) and particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
+We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (Figure~\ref{fig:viruspopulations}A) and to particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
 In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
@@
 We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
+For all samples except for the replicate 8-hour timepoint, we replaced the infection inoculum with fresh media after one hour, thereby ensuring that most infection was initiated during a narrow time window.
 We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
@@
 But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
-While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted close to half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}C).
-%Therefore, there is extraordinary variability in the number of viral transcripts per infected cell.
+While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}C).
 
@@
 However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
-Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection with both variants.
+Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection.
 
@@
 However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
-Because the bulk of cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
+Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
 
@@
 The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
-In fact, a bare few infected cells contribute the majority of influenza transcripts.
-Specifically, at 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
+At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
 
@@
 We wondered if the absence of specific viral genes might be associated with reduced amounts of viral mRNA within single infected cells.
-In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\cite{huang1990determination}.
-Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes for each of the four constituent RNP proteins so that the cell can produce more RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
+In particular, transcription of influenza virus mRNAs is performed by the viral ribonucleoprotein (RNP) complex, which consists of the three proteins that encode the tripartite polymerase (PB2, PB1, and PA) as well as nucleoprotein (NP)~\citep{huang1990determination}.
+Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes encoding each of the four RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
 This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).
@@
 Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
-Cells with high viral burden always express all the RNP genes, but some cells with high viral burden lack each of the other viral genes.
+Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
 {\bf (B)}
-Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each viral gene.
+Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
 A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
@@
 {\bf (C)}
-Among cells that express all four RNP genes, this panel shows the fraction of that express each of the four other genes, as well as the fraction that express \emph{all} four of the other genes.
-The fraction that express all four other genes is well predicted by simply multiplying the frequencies of cells that express each of these genes individually, indicating that gene absence is approximately independent across these four genes.
+The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
+The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
 }
@@
 
-We also sought to quantify the fraction of infected cells that completely failed to express a given gene.
+We also quantified the fraction of infected cells that completely failed to express a given gene.
 We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
@@
 The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C).
-Our measurements of the frequency at which infected cells fail to express individual viral genes are roughly consistent with estimates made by others using different means~\citep{Brooke:2013kb,Dou:2017cp}.
+If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 40-50\% of infected cells express mRNAs from all eight genes.
+This estimate of the frequency at which infected cells express mRNAs from all eight gene segments is slightly higher than previous estimates of 13\%~\citep{Brooke:2013kb} and 20\%~\citep{Dou:2017cp}.
+At least one difference is that \citet{Brooke:2013kb} stained for proteins whereas we examined the expression of mRNAs -- it is likely that some cells contain mutated viral genes that fail to produce stable protein even when mRNA is expressed.
 
-\subsection{The relative amounts of different viral mRNAs are fairly consistent across cells.}
-The results above show that the total amount of viral mRNA in infected cells varies over orders of magnitude.
+\subsection{The relative amounts of different viral mRNAs are more consistent across cells.}
+The results above show that the amount of viral mRNA in infected cells varies over several orders of magnitude.
 Does the relative expression of viral genes exhibit similar cell-to-cell variability?
@@
 The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
-The cell-to-cell consistency in the relative expression of different viral genes generally becomes even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
-Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA appear to have roughly similar effects on all viral genes in a given cell.
-This finding is consistent with prior work showing positive correlations among the abundance of several viral gene segments in individual cells ~\citep{Heldt:2015iz}.
+The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
+Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
+This finding is consistent with prior work showing positive correlations among the abundance of several viral genes in individual cells~\citep{Heldt:2015iz}.
 
@@
 Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
-In particular, influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
+Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
 In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
@@
 \caption{\label{fig:tsne}
-A t-SNE plot created by semi-supervised clustering of the cells using genes that co-vary with viral infection status.
+A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
 Each point is a single cell, and each panel shows an identical layout but colors the cells according to a different property.
 {\bf (A), (B)}
-Cells colored by the fraction of their mRNA that is derived from virus.
+Cells colored by the fraction of their mRNA derived from virus.
 {\bf (C)}
-Cells colored by the experimental sample from which they are derived.
+Cells colored by the experimental sample.
 {\bf (D), (E)}
@@
 
-We posited that the paucity of interferon induction might be due to the activity of influenza virus's major interferon antagonist, the NS1 protein.
+We posited that the paucity of interferon induction might be due to the activity of influenza virus's major interferon antagonist, the NS1 protein~\citep{garcia1998influenza,hale2008multifunctional}.
 We therefore identified cells that expressed substantial amounts of viral mRNA but lacked the NS gene (Figure~\ref{fig:tsne}F).
@@
 But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having $>$10\% of their mRNA derived from virus (Figure~\ref{fig:tsne}).
-Therefore, most infected cells fail to express interferon even when the virus lacks its major known interferon antagonist.
+Therefore, most infected cells fail to express interferon even when the virus lacks its major interferon antagonist.
 Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
-Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus clearly activate innate-immune responses much less frequently than might be presumed from bulk studies that flood cells with viral stocks full of defective particles.
+Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus only rarely activate innate-immune responses.
 
@@
 We examined whether any host genes were differentially expressed in cells with more viral mRNA.
-We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of any effects due to the stochastic presence or absence of particular viral transcripts.
+We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
 We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).
@@
 
-Many of the genes that exhibit increased expression in cells with large amounts of viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
+Many of the genes with increased expression in cells with more viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
 These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}).
@@
 Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation.
-\label{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
+\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
 

### 2017-09-12 paper/paper.tex
Tweaks to Methods

@@
 \figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
-The number of viral barcodes called for each sample and influenza gene segment. 
+The number of viral barcodes called for each sample and gene segment. 
 Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
@@
 \figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
-A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cels in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
+A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
 The cyclohexamide prevents translation of new PB2, PB1, PA, and NP protein, and so prevents the formation of the new RNPs needed for secondary transcription.
@@
 Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
-The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}
+The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
+
 To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
@@
 To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
-The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours, when obvious cytopathic effect was noted.
+The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.
 
-For the experiments in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, we created a virus that carried an HA gene segment in which the carried GFP in place of most of the HA coding sequence by following a scheme first described by \citet{marsh2007specific}.
-Briefly, we created a plasmid encoding a viral RNA with GFP in place of the HA coding sequence in the context of the pHH21~\citep{Neumann:1999ws} reverse-genetics plasmid, removing potential start codons upstream of the GFP (see Figure~\ref{fig:coexpression}-source~data~\ref{figdata:HAflank} for the sequence of the viral RNA)
-We then generated GFP-carrying virus by reverse-genetics in previously described cells constitutively expressing HA~\citep{Doud:2016gm}.
-A unidirectional vector for generating viral vRNA encoding eGFP flanked by HA packaging sequences was generated from a previously described vector, pHH21~\citep{Neumann:1999ws}. 
-To obtain sufficient titers, this HA-eGFP virus was expanded for 44 than than 36 hours after initiating infection at an MOI of 0.01.
+For the experiments in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, we created a virus that carried an HA gene segment in which GFP replaced most of the HA coding sequence, following a scheme first described by \citet{marsh2007specific}.
+Briefly, we created a plasmid encoding a viral RNA with GFP in place of the HA coding sequence in the context of the pHH21~\citep{Neumann:1999ws} reverse-genetics plasmid, removing potential start codons upstream of the GFP (see Figure~\ref{fig:coexpression}-source~data~\ref{figdata:HAflank} for the sequence of the viral RNA).
+We then generated GFP-carrying virus by reverse-genetics in cells constitutively expressing HA~\citep{Doud:2016gm}.
+To obtain sufficient titers, this HA-eGFP virus was expanded for 44 rather than 36 hours after initiating infection at an MOI of 0.01.
 
@@
 Biological triplicates were performed for all samples.
-For measurements of viral genomic HA content, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN Rneasy plus miniprep kit protocol, and cDNA was generated using  SuperScript\textsuperscript{TM} III first-strand synthesis supermix with primers specific for vRNA adapted from \citep{Hoffmann:2001vj}  using the manufacturer's protocol for random-hexamer priming~\citep{Xue:2017dl}.
-qPCR was then performed as for mRNA measurements.
-In order to obtain molecular resolution on our measurements of HA vRNA, a standard curve was generated from three independent dilutions of the HA-encoding reverse genetics plasmid. 
+
+For the measurements of viral genomic HA content in Figure~\ref{fig:viruspopulations}A, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol.
+The cDNA was generated using SuperScript\textsuperscript{TM} III first-strand synthesis supermix using the manufacturer's protocol, and using the universal vRNA primers of \citet{Hoffmann:2001vj} with the modifications described in  \citet{Xue:2017dl}.
+The qPCR was then performed as for mRNA measurements.
+A standard curve was generated from three independent dilutions of the HA-encoding reverse genetics plasmid. 
 All vRNA values represent three independent RNA extractions with two replicate qPCR measurements.
@@
 Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
-After subsequent washes in supplemented PBS, cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
+After washing in PBS plus 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
 Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
-To determine HA-expressing units, we then determined the fraction of cells that were HA positive and calculated the HA-expressing units from these fractions assuming a Poisson infection process.
+We then determined the fraction of cells that were HA positive and calculated the HA-expressing units assuming a Poisson infection process.
 
-For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating to exclude multiplets in FlowJo, data were subsequently extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
+For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
@@
 \subsection{Infections for single-cell mRNA sequencing}
-Single cell sequencing libraries were generated using the 10x Chromium Single Cell 3' platform, V1~\cite{zheng2017massively}.
-All time points except for the 8-hour repeat, were performed on the same day.
+Single-cell sequencing libraries were generated using the 10x Chromium Single Cell 3' platform~\citep{zheng2017massively} using the V1 reagents.
 
+All time points except for the 8-hour replicate were prepared on the same day.
 For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
@@
 Samples were sequenced on an Illumina HiSeq. 
-The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Jesse will upload and add this.}
+The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{needs to be added.}
 
@@
 Jupyter notebooks that perform all of the computational analyses are available at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}
+If you are a reviewer who needs access, please contact the editor.}, and are also available in Supplementary~file~\ref{suppfile:code}.
 
@@
 This annotated cell-gene matrix is in Supplementary~file~\ref{suppfile:cellgenematrix}.
-The Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/flu_single_cell/blob/master/align_and_annotate.ipynb} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}, and is also available as Supplementary~file~\ref{suppfile:alignandannotate}.
+A Jupyter notebook that performs these analyses is in Supplementary~file~\ref{suppfile:code}.
 
-The annotated cell-gene matrix was then further analyzed in R, primarily using the Monocle package (version 2.4.0)~\citep{Qiu:2017dw,Trapnell:2014kg}.
-The Jupyter notebook that performs these analyses is at \url{https://github.com/jbloomlab/flu_single_cell/blob/master/monocle_analysis.ipynb} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}, and is also available as Supplementary~file~\ref{suppfile:monocle}.
-
-Briefly, this analysis was performed as follows.
-For each sample, cell barcodes with 2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were considered to be potential doublets or inefficient/empty reactions, and were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}C).
+The annotated cell-gene matrix was analyzed in R, primarily using the Monocle package (version 2.4.0)~\citep{Qiu:2017dw,Trapnell:2014kg}.
+A Jupyter notebook that performs these analyses is in Supplementary~file~\ref{suppfile:code}.
+For each sample, cell barcodes that had $>$2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}C).
 In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C).
@@
 This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
-The 10-hour sample has the most viral mRNA, and is expected to have the most lysis-associated influenza reads.
+The 10-hour sample has the most viral mRNA, and so is expected to have the most lysis-associated influenza reads.
 We therefore determined the threshold using just this sample (Figure~\ref{fig:viralbarcodes}C), and then applied this threshold to all samples.
@@
 For subsequent analyses, we retained all infected cells and a subsample of uninfected cells (the greater of 50 or the number of infected cells for that sample).
-The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the t-SNE clustering and other downstream analyses.
+The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the downstream analyses.
 
 For the semi-supervised t-SNE clustering, we used Monocle's cell hierarchy function to bin cells into those with no viral mRNA, <2\% viral mRNA, between 2\% and 20\% viral mRNA, and >20\%. 
-Candidate marker genes t-SNE dimensionality reduction were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
+Candidate marker genes for t-SNE dimensionality reduction were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
 The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
@@
 Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation.
-\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
+Figure~\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
 
 \section{Acknowledgments}
-
-Additional information can be given in the template, such as to not include funder information in the acknowledgments section.
+We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
 
@@
 \begin{suppfile}
-\caption{\label{suppfile:cellgenematrix}
-The annotated cell-gene matrix in Matrix Market Format.}
+\caption{\label{suppfile:code}
+Computer code for the analyses.
+This ZIP file contains a Jupyter notebook that runs CellRanger to align and annotate the reads, and a Jupyter notebook that uses Monocle to analyze the cell-gene matrix.
+The ZIP file also includes associated custom scripts.}
 \end{suppfile}
@@
 \begin{suppfile}
-\caption{\label{suppfile:alignandannotate}
-Jupyter notebook that runs CellRanger to align and annotate the reads.
-The ZIP file also includes associated custom scripts run by this notebook.}
-\end{suppfile}
-
-\begin{suppfile}
-\caption{\label{suppfile:monocle}
-Jupyter notebook that analyzes the cell-gene matrix, primarily using Monocle.}
+\caption{\label{suppfile:cellgenematrix}
+The annotated cell-gene matrix in Matrix Market Format.}
 \end{suppfile}

### 2017-09-13 paper/paper.tex
re-draft of Intro

@@
 \DeclareSIUnit\Molar{M}
-\usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document
+%\usepackage{indentfirst} %I prefer to have indents following section headings, I find it to be more consistent in style throughout document
 \usepackage{natbib}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Extreme heterogeneity of influenza virus transcription in single cells}
+\title{Extreme heterogeneity of influenza virus transcriptional load in single cells}
 
@@
 \section{Introduction}
-Recent advances in sequencing and visualization technologies have revealed that cell-to-cell heterogeneity is a fundamental feature of biological processes -- impacting events from bacterial entrance into quiescent states to organogenesis in multicellular organisms~\citep{JohnstonJr:2010jn}.
-Due to this heterogeneity, biological pathways generated from bulk population data may not accurately describe the timing, order, or even the identity of events within single cells ~\citep{Trapnell:2014kg}.
-Nearly no other feature of biological life is as defined by heterogeneity as viral infection, wherein events can be highly stochastic due to the incredible combination of both genetic and non-genetic sources of variation.
+Viruses can cause massive and rapid changes in a cell's transcriptome as they churn out viral mRNAs and hijack cellular machinery.
+For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{Hatada:1989vz}.
+In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
+This anti-viral response is another prominent transcriptional signature of high-MOI influenza virus infection in bulk cells~\citep{Geiss:2002a}.
 
-The contribution of heterogeneity to infectious processes has been well appreciated in the study of viral quasispecies, wherein pathogenesis is best described as an emergent process dependent upon underlying genetic diversity ~\citep{Lauring:2010if}.
-This concept has revolutionized the study of RNA viruses: including polio, hepatitis C, and influenza \citep{Echeverria:2015ew, Kuroda:2010gt}. 
-Direct evidence of the contribution of the quasispecies to pathogenesis has been observed -- for instance mutant strains of polio that produce less variation exhibit significantly less fitness in animal models of infection despite displaying wild-type replication kinetics \emph{in vitro}~\citep{Vignuzzi:2005ju}.
-However, these studies have largely been limited to population-level analyses, which can link standing diversity to complex biological phenomena but are unable to deconvolute precise events in individual infected cells.
+However, initiation of an actual influenza infection typically involves just a few virions infecting a few cells~\citep{varble2014influenza,mccrone2017evolutionary}. 
+The dynamics of viral infection in these individual cells may not mirror bulk measurements made on many cells infected at high MOI.
+Over 70 years ago, \citet{Delbruck:1945wo} showed that there was a $\sim$100-fold range in the number of progeny virions produced per cell by clonal bacteria infected with clonal bacteriophage.
+Subsequent work has shown similar heterogeneity in the number of progeny virions per cell during infection with other viruses~\citep{Zhu:2009ea,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.
 
-Non-genetic variation has also been described to contribute to the host response to infection. 
-Single-cell transcriptomic analyses have revealed incredibly coordinated responses driven by stochasticity, even in tissue culture models using genetically homogeneous cells.
-This heterogeneity in response has been observed both with respect to pure ligands such as lipopolysaccharide, wherein differences must be host-cell intrinsic ~\citep{Shalek:2014ey,Shalek:2013ej,Rand:2012kl}, and with respect to pathogens, wherein differences are likely a complex mixture of host-cell intrinsic and pathogen-specific heterogeneity ~\citep{Avraham:2015cj}.
-Moreover, stochasticity is observed both in interferon production, which drives the anti-pathogen state, and the expression of the downstream effector molecules of innate immunity.
-As inflammation is a potentially deleterious process, restricting a response to a small subset of cells upon initial recognition of a threat might be particularly advantageous.
+In the case of influenza virus infection, targeted single-cell measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
+The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
+In addition, the activation of innate-immune responses is inherently stochastic~\citep{Shalek:2013ej,Shalek:2014ey,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
+However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.
 
-Influenza, a segmented negative sense RNA virus, places a significant infectious burden on the global population and yet, despite many decades of study, there is still significant conjecture regarding those features of viral populations that drive observed phenotypes.
-Cells infected with this virus exhibit a wide degree of heterogeneity, including a number of semi-infectious events wherein the full complement of viral gene products is absent~\citep{Brooke:2013kb, Heldt:2015iz,Dou:2017cp}.
-This has broadly been observed to impact rates of recombination, however its contribution to other viral behaviors remains unclear~\citep{Fonville:2015cg}. 
-Influenza-infected cells are also observed to exhibit a wide-breadth of productivity -- exhibiting a 3-log range in infectious progeny even in high-MOI infections~\citep{Heldt:2015iz}. 
-This appears to be a general trend of many RNA viral infections, from phage to polio, and the features underlying these observations appear multitudinous~\citep{DELBRUCK:1945wo, Schulte:2014br}.
-
-We now hope to apply new single-cell transcriptomic technologies to the study of influenza replication in a tissue culture model of infection.
-Early single-cell transcriptomic approaches were laborious -- requiring the generation of sequencing libraries on individual cells~\citep{Tang:2009kj,ld:2012gj}.
-Breakthroughs in microfluidics helped to automate this process, but retained strictly parallel processing of sequencing libraries.
-More recent technologies have introduced more complex means of barcoding and partitioning cells, allowing for the generation of libraries consisting of thousands of cells with ?single pot?
-downstream processing.
-These include the use of reverse-phase emulsions, gravity capture, and even \emph{in situ} reverse transcription after cellular partitioning~\citep{JunSSi,Gierahn:2017bc,Klein:2015ki,Macosko:2015ht}. 
-We hope to leverage these technological advances in the study of influenza, using transcriptomic analysis to understand the phenotypic heterogeneity that abounds during infection with this virus.
+Here we use single-cell mRNA sequencing to quantify the levels of all cellular and viral mRNAs in cells infected with influenza virus at low MOI.
+We find extremely large variation in the amount of viral mRNA expressed in individual cells.
+Both co-infection and activation of innate-immune pathways are rare in our low-MOI infections, and do not appear to be the major drivers of cell-to-cell heterogeneity in viral transcriptional load.
+Individual infected cells often fail to express specific viral genes, and such gene absence explains some but certainly not all of the cell-to-cell heterogeneity.
+A variety of cellular genes, including ones involved in the oxidative-stress response, co-vary with viral transcriptional load.
+Overall, our work demonstrates remarkable heterogeneity in the transcriptional outcome of influenza virus infection among nominally identical cells infected with a relatively pure population of virions under controlled conditions.
 
@@
 Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
-These defective particles become prevalent when a virus is grown at high multiplicity of infection (MOI), where complementation permits the growth of otherwise deleterious genotypes.
+These defective particles become prevalent when a virus is grown at high MOI, where complementation permits the growth of otherwise deleterious genotypes.
 To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
@@
 At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
+Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation at least in our low-MOI infections.
 
@@
 For instance, the cell with 11.2\% of its mRNA from virus in the upper right of Figure~\ref{fig:coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
-Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,Fonville:2015cg,aguilera2017plaques}.
+Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,brooke2014influenza,Fonville:2015cg,aguilera2017plaques}.
 
@@
 Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
-In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
-However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2014ey,Shalek:2013ej,PerezCidoncha:2014jr,bhushal2017cell}.
+%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
+However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2013ej,Shalek:2014ey,PerezCidoncha:2014jr,bhushal2017cell}.
 We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.
@@
 
-\subsection{Genes involved in the oxidative-stress response co-vary with viral gene expression.}
+\subsection{Some host genes co-vary with viral gene expression.}
 We examined whether any host genes were differentially expressed in cells with more viral mRNA.
@@
 It is our hope that this work will serve not only as a window into influenza replication, but as a part of a grander design in re-examining and understanding the fundamental features of viral infection.
+
+MATERIAL FROM INTRODUCTION THAT MIGHT BE USEFUL HERE:
+
+Recent advances in sequencing and visualization technologies have revealed that cell-to-cell heterogeneity is a fundamental feature of biological processes -- impacting events from bacterial entrance into quiescent states to organogenesis in multicellular organisms~\citep{JohnstonJr:2010jn}.
+Due to this heterogeneity, biological pathways generated from bulk population data may not accurately describe the timing, order, or even the identity of events within single cells ~\citep{Trapnell:2014kg}.
+Nearly no other feature of biological life is as defined by heterogeneity as viral infection, wherein events can be highly stochastic due to the incredible combination of both genetic and non-genetic sources of variation.
+
+The contribution of heterogeneity to infectious processes has been well appreciated in the study of viral quasispecies, wherein pathogenesis is best described as an emergent process dependent upon underlying genetic diversity ~\citep{Lauring:2010if}.
+This concept has revolutionized the study of RNA viruses: including polio, hepatitis C, and influenza \citep{Echeverria:2015ew, Kuroda:2010gt}. 
+Direct evidence of the contribution of the quasispecies to pathogenesis has been observed -- for instance mutant strains of polio that produce less variation exhibit significantly less fitness in animal models of infection despite displaying wild-type replication kinetics \emph{in vitro}~\citep{Vignuzzi:2005ju}.
+However, these studies have largely been limited to population-level analyses, which can link standing diversity to complex biological phenomena but are unable to deconvolute precise events in individual infected cells.
+
+Non-genetic variation has also been described to contribute to the host response to infection. 
+Single-cell transcriptomic analyses have revealed incredibly coordinated responses driven by stochasticity, even in tissue culture models using genetically homogeneous cells.
+This heterogeneity in response has been observed both with respect to pure ligands such as lipopolysaccharide, wherein differences must be host-cell intrinsic ~\citep{Shalek:2014ey,Shalek:2013ej,Rand:2012kl}, and with respect to pathogens, wherein differences are likely a complex mixture of host-cell intrinsic and pathogen-specific heterogeneity ~\citep{Avraham:2015cj}.
+Moreover, stochasticity is observed both in interferon production, which drives the anti-pathogen state, and the expression of the downstream effector molecules of innate immunity.
+As inflammation is a potentially deleterious process, restricting a response to a small subset of cells upon initial recognition of a threat might be particularly advantageous.
+
+We now hope to apply new single-cell transcriptomic technologies to the study of influenza replication in a tissue culture model of infection.
+Early single-cell transcriptomic approaches were laborious -- requiring the generation of sequencing libraries on individual cells~\citep{Tang:2009kj,ld:2012gj}.
+Breakthroughs in microfluidics helped to automate this process, but retained strictly parallel processing of sequencing libraries.
+More recent technologies have introduced more complex means of barcoding and partitioning cells, allowing for the generation of libraries consisting of thousands of cells with ?single pot?
+downstream processing.
+These include the use of reverse-phase emulsions, gravity capture, and even \emph{in situ} reverse transcription after cellular partitioning~\citep{JunSSi,Gierahn:2017bc,Klein:2015ki,Macosko:2015ht}. 
+We hope to leverage these technological advances in the study of influenza, using transcriptomic analysis to understand the phenotypic heterogeneity that abounds during infection with this virus.
 

### 2017-09-13 paper/paper.tex
re-draft abstract and work on beginning of Discussion

@@
 \begin{abstract}
-Influenza virus infection induces large changes in cellular transcription.
-Previously this has mostly been looked at using bulk measurements
-Here we examine the process at the level of single cells.
-We find extremely wide variation in the extent of viral gene transcription across infected cells.
-IFN induction is very rare.
-Some cellular pathways may be consistently altered in cells with high burden of viral transcripts.
-Overall, highlights remarkable heterogeneity in the outcome of infection.
+Viral infection can cause large changes in a cell's transcriptome, but these changes have mostly been studied by bulk measurements on many cells.
+We use single-cell mRNA sequencing to examine the consequences of influenza virus infection in individual cells.
+We find extremely wide cell-to-cell variation in viral gene expression -- viral transcripts compose no more than a few percent of total mRNA in most infected cells, but a few cells derive over half their mRNA from virus.
+Many infected cells fail to express at least one viral gene, and this gene absence explains some of the variation in viral transcriptional load.
+Activation of innate immune pathways is rare, even in cells that fail to express influenza virus's major interferon antagonist.
+We identify some cellular genes that co-vary in abundance with the amount of viral mRNA.
+Overall, our results highlight the extensive cell-to-cell heterogeneity in the transcriptional consequences of viral infection.
 \end{abstract}
@@
 \section{Discussion}
-We have generated total transcriptome measurements of individual influenza-infected cells -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
+We have quantified the total transcriptome composition of single cells infected with influenza virus.
+While we certainly observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.
+
+One of the most striking forms of heterogeneity is the complete failure of some infected cells to express one or more viral genes, which we estimate occurs in about half the infected cells in our experiments.
+The absence of viral genes in infected cells has been noted previously using more targeted approaches~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the the total viral transcriptional load as a function of the level of each mRNA.
+We find that absence of any of the four genes that encode the viral RNP leads to a dramatic drop in the total amount of viral mRNA in an infected cell, consistent with prior bulk studies of viral transcription~\citep{Vreede:2004ip,eisfeld2015centre}.
+Interestingly, the \emph{reason} some cells fail to express some viral genes remains unclear.
+The prototypical influenza virion packages one copy of each of the eight gene segments~\citep{noda2006architecture,hutchinson2010genome}, but some virions surely package fewer~\citep{brooke2014influenza}.
+However, it is also possible that much of the viral gene absence is due to stochastic loss of viral RNPs after infection but prior to the initiation of viral transcription in the nucleus.
+Further work will be needed to distinguish between these possibilities.
+
+ -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
 The dataset we have generated is largely consistent with prior flow cytometry-based analyses, allowing us to build upon those prior observations with a more holistic view, beginning to understand the consequences of heterogeneity in influenza.
@@
 Indeed, our analyses of host transcriptomes suggests that those cells expressing higher levels of influenza are distinguishable, even at 6 hours, from later time points expressing similar influenza burden, giving an indication that there are features above-and-beyond temporal variation driving differences in viral mRNA expression\jdbcomment{If we make this point we need to add the ranked graph to the results and address it} .
-
-One of the most striking features of stochasticity with respect to influenza infection is the absence of genes corresponding to a given genomic segment, observed both by ourselves and others~\citep{Brooke:2013kb,Dou:2017cp}
-While there is significant speculation, and some indirect evidence, regarding the step of viral replication at which these genes are lost, there currently exists no strong consensus as to its identity -- indeed there may be no single event wholly responsible the stochastic loss of viral gene expression ~\citep{Brooke:2017gm, Heldt:2015iz}
-Regardless it does not seem unreasonable to surmise that this is a consequence of possessing a segmented genome.
-If so, it can be reasonably concluded that at some phase of viral replication, whether packaging, delivery to the nucleus, production of replicative intermediates, or some as-yet unsurmised step, the components of the viral genome must act in at least a semi-independent fashion.
-
-A consequence of this stochasticity is a subpopulation of infected cells that express significantly lower levels of viral transcripts. 
-Consistent with prior population-level studies of transcription, we find that absence of any of the four viral gene products necessary to form the viral polymerase leads to a dramatic drop in the levels of mRNA transcripts detected~\citep{Vreede:2004ip,eisfeld2015centre}.
-Our results are in slight contrast to a prior flow cytometric analysis of protein expression that described a relatively minor drop in expression associated with absence of viral ribonucleoprotein~\citep{Brooke:2013kb}.
-This discrepancy could be explained by two possibilities: the first is that mRNA measurements may not be wholly reflective of protein levels, and the second is that flow cytometry requires decisions regarding sensor sensitivity and positive gates and the potential biases that those decisions incur.
-While sequencing-based analysis comes with its own set of caveats, the detection threshold is 'baked-in' and so potentially less subject to bias.
 

### 2017-09-14 paper/paper.tex
added version of Figure 5A for each sample individually

@@
 We find extremely wide cell-to-cell variation in viral gene expression -- viral transcripts compose no more than a few percent of total mRNA in most infected cells, but a few cells derive over half their mRNA from virus.
-Many infected cells fail to express at least one viral gene, and this gene absence explains some of the variation in viral transcriptional load.
+Some infected cells fail to express at least one viral gene, and this gene absence partially explains the variation in viral transcriptional load.
 Activation of innate immune pathways is rare, even in cells that fail to express influenza virus's major interferon antagonist.
@@
 }{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
+\figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
+The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
+Points are colored as in Figure~\ref{fig:fluburdenbyflugene}.
+}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
 \figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
@@
 The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
+We do not show statistical analyses for other samples, as the number of infected cells is too low for meaningful statistical analyses.
 }{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
@@
 
-We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A). 
+We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
 Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
@@
 These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA.
-However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
+However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express(Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
 The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.
@@
 We have quantified the total transcriptome composition of single cells infected with influenza virus.
-While we certainly observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.
+While we observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.
 
@@
 Further work will be needed to distinguish between these possibilities.
+
+However, the absence of specific viral genes in some cells explains only a fraction of the variation in amount of viral mRNA, which still varies by over ...
 
@@
 Indeed, our analyses of host transcriptomes suggests that those cells expressing higher levels of influenza are distinguishable, even at 6 hours, from later time points expressing similar influenza burden, giving an indication that there are features above-and-beyond temporal variation driving differences in viral mRNA expression\jdbcomment{If we make this point we need to add the ranked graph to the results and address it} .
-
-However, despite extensive heterogeneity, there seems to be robustness to the infectious lifecycle.
-For example, influenza, regardless of the viral burden, exhibits relatively constrained ratios between the viral transcripts.
-This suggests that the processes that are driving heterogeneity of influenza mRNA abundance act equivalently throughout the viral genome.
-Potential mechanisms include components such as binding and entry kinetics that act on a virion and thus on all segments packaged therein, limitations of transcription rate that are dependent upon the same pool of cellular resources, or potentially even feedback-dependent mechanisms controlling viral transcription.
 

### 2017-09-14 paper/paper.tex
revisions to Discussion -- still need another pass over this before sending back to Alistair.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Extreme heterogeneity of influenza virus transcriptional load in single cells}
+\title{Extreme heterogeneity of influenza virus transcription in single cells}
 
@@
 Viruses can cause massive and rapid changes in a cell's transcriptome as they churn out viral mRNAs and hijack cellular machinery.
-For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{Hatada:1989vz}.
+For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{hatada1989control}.
 In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
@@
 The dynamics of viral infection in these individual cells may not mirror bulk measurements made on many cells infected at high MOI.
-Over 70 years ago, \citet{Delbruck:1945wo} showed that there was a $\sim$100-fold range in the number of progeny virions produced per cell by clonal bacteria infected with clonal bacteriophage.
-Subsequent work has shown similar heterogeneity in the number of progeny virions per cell during infection with other viruses~\citep{Zhu:2009ea,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.
+Over 70 years ago, Max Delbruck showed that there was a $\sim$100-fold range in the number of progeny virions produced per cell by clonal bacteria infected with clonal bacteriophage~\citep{Delbruck:1945wo}.
+Subsequent work has shown similar heterogeneity during infection with other viruses~\citep{Zhu:2009ea,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.
 
@@
 The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
-In addition, the activation of innate-immune responses is inherently stochastic~\citep{Shalek:2013ej,Shalek:2014ey,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
+In addition, the activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
 However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.
@@
 We engineered a virus that carried two synonymous mutations near the 3' end of each mRNA.
-At top are the specific mutations for PB2.
+At top are the mutations for PB2.
 At bottom are locations of the synonymous mutations relative to the typical distribution of read depth for our 3'-end sequencing.
@@
 The wild-type and synonymously barcoded viruses transcribe their genes with similar kinetics. 
-Shown is the abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 as assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
+The abundance of the viral hemagglutinin (HA) transcript relative to the cellular housekeeping gene L32 was assessed by qPCR in A549 cells infected at an MOI of 0.5 (as determined on MDCK-SIAT1 cells).
 Error bars $\pm$ S.D., n=3.
@@
 {\bf (B)} 
-Our viral stocks have a higher ratio of infectious particles to particles capable of expressing the viral HA protein.
-A549 cells were infected with all three viral stocks at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
+Our viral stocks have a higher ratio of infectious particles to particles capable of expressing HA protein.
+A549 cells were infected at an MOI of 0.1, and the percentage of cells expressing HA protein at 9 hours post-infection was quantified by antibody staining and flow cytometry.
 {\bf (C)} 
@@
 Error bars $\pm$ S.D., n=3.
-Note that the MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
+Note that MOIs were calculated by TCID50 on MDCK-SIAT1 cells, whereas the experiments in this figure involved infection of A549 cells.
 }
@@
 
-\subsection{Absence of viral genes partially explains cell-to-cell variability in viral transcript abundance.}
+\subsection{Absence of viral genes partially explains cell-to-cell variability in viral load.}
 The influenza genome is segmented, and cells can fail to express a viral mRNA if the encoding gene segment is not packaged in the infecting virion or fails to initiate transcription after infection.
@@
 }
-\figsupp[Secondary transcription from newly synthesized RNPs is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
+\figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
 A549 cells were infected at an MOI of 0.2 as calculated on MDCK-SIAT1 cells in either the presence or absence of the protein-translation inhibitor cyclohexamide, and viral mRNA was quantified at 8 hours post-infection by qPCR.
@@
 These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA.
-However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express(Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
+However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
+Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
 The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.
@@
 Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
-The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{Hatada:1989vz}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
+The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
 The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
@@
 %In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
-However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,Shalek:2013ej,Shalek:2014ey,PerezCidoncha:2014jr,bhushal2017cell}.
+However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell}.
 We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.
@@
 \figdata{The results of a gene-set analysis are in \texttt{p\_sig\_cellular\_genes.csv}.}
-\figsupp[Many genes that co-vary with the amount of viral mRNA are involved in the response to oxidative stress. \label{figsupp:ROStable}]{
-Table delineating genes in (Figure~\ref{fig:cellulargenes}) that are associated with the response to oxidative stress ~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
+\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
+Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
 }{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
@@
 Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
-Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.
+%Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.
 
@@
 We have quantified the total transcriptome composition of single cells infected with influenza virus.
-While we observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{Hatada:1989vz,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.
+While we observe a general increase in the amount of viral mRNA over time as expected from bulk measurements~\citep{hatada1989control,Shapiro:1987ur}, there is wide variation in viral gene expression among individual infected cells.
 
-One of the most striking forms of heterogeneity is the complete failure of some infected cells to express one or more viral genes, which we estimate occurs in about half the infected cells in our experiments.
-The absence of viral genes in infected cells has been noted previously using more targeted approaches~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the the total viral transcriptional load as a function of the level of each mRNA.
-We find that absence of any of the four genes that encode the viral RNP leads to a dramatic drop in the total amount of viral mRNA in an infected cell, consistent with prior bulk studies of viral transcription~\citep{Vreede:2004ip,eisfeld2015centre}.
+The most obvious form of heterogeneity is the complete failure of some infected cells to express one or more viral genes, which we estimate occurs in about half the infected cells in our experiments.
+The absence of some viral genes in some infected cells has been noted previously~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the the total viral transcriptional load as a function of the level of each mRNA.
+We find that absence of any of the four genes that encode the viral RNP leads to a dramatic drop in the total amount of viral mRNA, consistent with prior bulk studies~\citep{Vreede:2004ip,eisfeld2015centre}.
 Interestingly, the \emph{reason} some cells fail to express some viral genes remains unclear.
@@
 However, it is also possible that much of the viral gene absence is due to stochastic loss of viral RNPs after infection but prior to the initiation of viral transcription in the nucleus.
-Further work will be needed to distinguish between these possibilities.
+%Further work is needed to distinguish between these possibilities.
 
-However, the absence of specific viral genes in some cells explains only a fraction of the variation in amount of viral mRNA, which still varies by over ...
+The absence of specific viral genes in some cells only partially explains the cell-to-cell variation in amount of viral mRNA, which still varies from $<$1\% to $>$50\% among cells expressing all the viral genes.
+It is likely that other viral genetic factors explain some of this remaining heterogeneity.
+The 3'-end sequencing strategy used in our experiments detects the presence of a viral gene, but does not identify whether that gene contains a mutation that might hinder viral replication.
+However, viral mutations are also unlikely to explain all the observed heterogeneity, since the current consensus estimates of influenza virus's mutation rate suggest that the typical virion in a stock such as the one used in our experiment should contain less than one mutation per genome~\citep{parvin1986measurement,suarez1992heterogeneity,suarez1994estimation,nobusawa2006comparison,bloom2014experimentally,pauly2017novel}.
 
- -- providing additional insight into the processes that drive influenza replication and its impact on the host at the cellular level.
-The dataset we have generated is largely consistent with prior flow cytometry-based analyses, allowing us to build upon those prior observations with a more holistic view, beginning to understand the consequences of heterogeneity in influenza.
-Overall we find that influenza replication is defined broadly by both significant stochasticity and significant robustness, and that a view of viral infection as either strictly deterministic or incredibly chaotic fails to accurately describe the infectious process.
-Moreover we found that stress, likely induced by oxidative damage, appears to define the cellular response to infection.
+The rest of the heterogeneity must be due to some combination of cellular factors and inherent stochasticity.
+We do find some features of the cellular transcriptome that co-vary with the amount of influenza mRNA in individual cells. 
+In particular, the viral load in individual cells is positively associated with the expression of genes involved in response to cellular stresses, including oxidative stress.
+It will be interesting to determine if these cellular transcriptional signatures are purely a consequence of the stress imposed by a large burden of viral replication, or if their stronger activation in some cells is a causative factor that promotes viral transcription.
+However, it would not be surprising if a substantial amount of the cell-to-cell heterogeneity can never be ascribed to pre-existing features of the viral genome or cellular state. 
+Apparently stochastic heterogeneity is a common feature of many processes that have been examined at a single-cell level~\citep{cai2006stochastic,raj2006stochastic,buganim2012single,shalek2013single,avraham2015pathogen} -- especially when those processes are initiated by very small numbers of initial molecules~\citep{elowitz2002stochastic}, as is the case for low-MOI viral infection.
 
-It is immediately apparent that a model of influenza infection wherein there is equivalent accumulation of transcripts across successfully infected cells is far to facile to describe the true situation.
-While infection certainly proceeds over time, with gradual increase in influenza transcript abundance, the level of transcripts expressed in cells varies wildly.
-While some of these differences are likely due to temporal variation in binding and internalization, it will be interesting in follow-up experiments to determine whether there is an ultimate level of influenza transcript that is invariably reached or whether the variation we observe is indicative of a final setpoint.
-Indeed, our analyses of host transcriptomes suggests that those cells expressing higher levels of influenza are distinguishable, even at 6 hours, from later time points expressing similar influenza burden, giving an indication that there are features above-and-beyond temporal variation driving differences in viral mRNA expression\jdbcomment{If we make this point we need to add the ranked graph to the results and address it} .
+One factor that rather surprisingly does \emph{not} appreciably contribute to the heterogeneity in our experiments is activation of innate-immune interferon pathways.
+Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of these cells fail to express the influenza-virus interferon antagonist NS1.
+It is known that interferon activation is stochastic at the level of single cells in response to both synthetic ligands~\citep{shalek2013single,shalek2014single,bhushal2017cell} and actual infection~\citep{Rand:2012kl,PerezCidoncha:2014jr,avraham2015pathogen,Killip:2017ef}.
+But interferon expression is a prominent transcriptional signature of high-MOI influenza virus infection of bulk cells, including in the epithelial cell line and at the time-points used in our experiments~\citep{Geiss:2002a,sutejo2012activation}.
+So it is notable how rarely single cells express interferon.
+Interferon expression would surely be more common at later times or with a viral stock passaged at higher MOI, since paracrine interferon signaling~\citep{crotta2013type} and accumulation of defective viral particles enhance innate-immune detection~\citep{Tapia:2013kf,Lopez:2014en}.
+However, the early events of physiological influenza infection involve just a few infecting virions~\citep{varble2014influenza,mccrone2017evolutionary}, and so it is interesting to speculate whether rare stochastic interferon activation during the first few cycles of viral replication could contribute to heterogeneity in the eventual outcome of infection.
 
-Perhaps even more important to influenza's success in the face of stochasticity is the capacity of the virus to control the host innate immune response despite a large number of non-infectious events.
-While non-infectious events might be considered metabolic 'waste', they would, theoretically, not impact the growth of successful viral infections. 
-However, recognition of any virus, infectious or not, and production of interferon, would impact the entire viral population. 
-While we cannot extrapolate much from our single interferon-positive cell, it can be stated that interferon induction by influenza is rare in our dataset. 
-This is consistent with prior experiments using reporter cell lines,  which places the induction at ~1\% of cells with detectable influenza protein expression -- albeit this value is likely sensitive to differences in viral propagation and strain utilized ~\citep{Killip:2017ef, Chen:2010cr}.
-If the major influenza interferon antagonist, NS1, is absent in ~10-20\% of infections, and lowly expressed in those infections engaging in only primary transcription, we must posit that NS1 alone is insufficient to explain influenza evasion of innate immunity.
-Interestingly, if other segmented viruses also exhibit a pattern of stochastic segment loss, they too would need to overcome this hurdle and escape innate immunity through mechanisms relying on more than a gene product generated by a single segment.
+Interestingly, despite the extreme heterogeneity in the amount of total viral mRNA per cell, the relative levels of the various viral mRNAs are more consistent across cells.
+If we ignore the variation due to the fact that some cells completely fail to express specific viral genes, we find that the relative levels of the different viral mRNAs are similar across cells and generally reflective of classical bulk measurements~\citep{hatada1989control}.
+This fact suggests that despite the stochasticity that would seem inherent in initiating transcription and replication of each gene from a single copy carried by the incoming virion, if that gene is not absent or completely lost, then the virus possesses mechanisms to control relative gene expression~\citep{Shapiro:1987ur,hatada1989control,perez2010influenza,heldt2012modeling,chua2013influenza}.
+This fact also suggests that the sources of cell-to-cell heterogeneity in viral transcriptional load must exert relatively concordant effects on all viral gene segments.
 
-In addition to viral mRNA kinetics, our analysis also allows us to observe those features of the cellular transcriptome that covary with influenza transcription.
-Critically, our analysis does not require synchronization or infection with a high MOI, and thus avoids potential artifacts produced by either approach. 
-We can therefore see extreme changes in a small subset of cells within a larger population.
-Moreover, we can also observe how transcriptional changes partition between cells, and exclude dramatic changes such as interferon induction that, while important, can mask other interesting biologically-relevant responses.
-With this in mind, the predominant response associated with increased influenza mRNA transcript abundance appears to be the oxidative stress pathway. 
-Generally characterized as antiviral, it is possible that cellular stress pathways represent an additional arm of innate immunity -- detecting a cellular state consistent with a loss of homeostasis as resources are diverted to viral propagation.
-Intriguingly, one element of the response, cyclophilin A (\emph{CypA}), is both thought to suppress influenza replication through degradation of matrix protein as well as act as a proinflammatory signal ~\citep{Liu:2009er, Jin:2004ii}.
-It will be interesting to follow up on these observations and determine the broader mechanisms through which cellular stress pathways are involved in pro- or anti-viral responses.
-
-The success garnered here underscores that single-cell approaches seem distinctly suited to study of viral pathogens, as not only are many host responses governed by bimodality and stochasticity ~\citep{Shalek:2014ey,Shalek:2013ej}, but also viral populations themselves exhibit a staggering array of diversity.
-Our results with influenza have provided significant support for earlier analysis of viral stochasticity, further defined the sources and impacts of that stochasticity, and generated potentially fertile avenues of research.
-We stand now at an exciting moment in virology, as technology allows us to observe population-level measurements as emergent phenomena defined by underlying individual stochasticity rather than analyzing viral replication in a strictly deterministic framework.
-It is our hope that this work will serve not only as a window into influenza replication, but as a part of a grander design in re-examining and understanding the fundamental features of viral infection.
-
-MATERIAL FROM INTRODUCTION THAT MIGHT BE USEFUL HERE:
-
-Recent advances in sequencing and visualization technologies have revealed that cell-to-cell heterogeneity is a fundamental feature of biological processes -- impacting events from bacterial entrance into quiescent states to organogenesis in multicellular organisms~\citep{JohnstonJr:2010jn}.
-Due to this heterogeneity, biological pathways generated from bulk population data may not accurately describe the timing, order, or even the identity of events within single cells ~\citep{Trapnell:2014kg}.
-Nearly no other feature of biological life is as defined by heterogeneity as viral infection, wherein events can be highly stochastic due to the incredible combination of both genetic and non-genetic sources of variation.
-
-The contribution of heterogeneity to infectious processes has been well appreciated in the study of viral quasispecies, wherein pathogenesis is best described as an emergent process dependent upon underlying genetic diversity ~\citep{Lauring:2010if}.
-This concept has revolutionized the study of RNA viruses: including polio, hepatitis C, and influenza \citep{Echeverria:2015ew, Kuroda:2010gt}. 
-Direct evidence of the contribution of the quasispecies to pathogenesis has been observed -- for instance mutant strains of polio that produce less variation exhibit significantly less fitness in animal models of infection despite displaying wild-type replication kinetics \emph{in vitro}~\citep{Vignuzzi:2005ju}.
-However, these studies have largely been limited to population-level analyses, which can link standing diversity to complex biological phenomena but are unable to deconvolute precise events in individual infected cells.
-
-Non-genetic variation has also been described to contribute to the host response to infection. 
-Single-cell transcriptomic analyses have revealed incredibly coordinated responses driven by stochasticity, even in tissue culture models using genetically homogeneous cells.
-This heterogeneity in response has been observed both with respect to pure ligands such as lipopolysaccharide, wherein differences must be host-cell intrinsic ~\citep{Shalek:2014ey,Shalek:2013ej,Rand:2012kl}, and with respect to pathogens, wherein differences are likely a complex mixture of host-cell intrinsic and pathogen-specific heterogeneity ~\citep{Avraham:2015cj}.
-Moreover, stochasticity is observed both in interferon production, which drives the anti-pathogen state, and the expression of the downstream effector molecules of innate immunity.
-As inflammation is a potentially deleterious process, restricting a response to a small subset of cells upon initial recognition of a threat might be particularly advantageous.
-
-We now hope to apply new single-cell transcriptomic technologies to the study of influenza replication in a tissue culture model of infection.
-Early single-cell transcriptomic approaches were laborious -- requiring the generation of sequencing libraries on individual cells~\citep{Tang:2009kj,ld:2012gj}.
-Breakthroughs in microfluidics helped to automate this process, but retained strictly parallel processing of sequencing libraries.
-More recent technologies have introduced more complex means of barcoding and partitioning cells, allowing for the generation of libraries consisting of thousands of cells with ?single pot?
-downstream processing.
-These include the use of reverse-phase emulsions, gravity capture, and even \emph{in situ} reverse transcription after cellular partitioning~\citep{JunSSi,Gierahn:2017bc,Klein:2015ki,Macosko:2015ht}. 
-We hope to leverage these technological advances in the study of influenza, using transcriptomic analysis to understand the phenotypic heterogeneity that abounds during infection with this virus.
+Overall, our work shows the power and importance of charactering cellular infection at the level of single cells~\citep{avraham2015pathogen}.
+The dynamics of viral infection in any given cell is shaped by the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes that are initiated by a single copy of the viral genome.
+Here we have shown how the confluence of these factors leads to extreme cell-to-cell heterogeneity in the transcriptional outcome of influenza virus infection.
+We anticipate that further deconstruction of the contributions of each factor will enable a deeper understanding of how the bulk features of infection emerge from the heterogeneous processes occurring within individual virus-infected cells.
 

### 2017-09-14 paper/paper.tex
Jesse revisions

@@
 \documentclass[9pt,lineno]{elife}
+\sloppy
 % Use the onehalfspacing option for 1.5 line spacing
@@
 \usepackage{float}
+\usepackage{tikz}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Extreme heterogeneity of influenza virus transcription in single cells}
+\title{Extreme heterogeneity of influenza virus infection in single cells}
 
@@
 We identify some cellular genes that co-vary in abundance with the amount of viral mRNA.
-Overall, our results highlight the extensive cell-to-cell heterogeneity in the transcriptional consequences of viral infection.
+Overall, our results highlight extensive cell-to-cell heterogeneity in the transcriptional consequences of viral infection.
 \end{abstract}
@@
 For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{hatada1989control}.
-In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
+In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
 This anti-viral response is another prominent transcriptional signature of high-MOI influenza virus infection in bulk cells~\citep{Geiss:2002a}.
@@
 Over 70 years ago, Max Delbruck showed that there was a $\sim$100-fold range in the number of progeny virions produced per cell by clonal bacteria infected with clonal bacteriophage~\citep{Delbruck:1945wo}.
-Subsequent work has shown similar heterogeneity during infection with other viruses~\citep{Zhu:2009ea,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.
+Subsequent work has shown similar heterogeneity during infection with other viruses~\citep{zhu2009growth,Schulte:2014br,combe2015single,Akpinar:2016gt}, including influenza virus~\citep{Heldt:2015iz}.
 
-In the case of influenza virus infection, targeted single-cell measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
+In the case of influenza virus infection, targeted measurements of specific proteins or RNAs have shed light on some factors that contribute to cell-to-cell heterogeneity. 
 The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
-In addition, the activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
+In addition, activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
 However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.
@@
 A variety of cellular genes, including ones involved in the oxidative-stress response, co-vary with viral transcriptional load.
-Overall, our work demonstrates remarkable heterogeneity in the transcriptional outcome of influenza virus infection among nominally identical cells infected with a relatively pure population of virions under controlled conditions.
+Overall, our work demonstrates remarkable heterogeneity in the transcriptional outcome of influenza virus infection among nominally identical cells infected with a relatively pure population of virions.
 
 \section{Results}
-
-\subsection{Strategy to measure mRNA in single virus-infected cells.}
-We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{zheng2017massively,Macosko:2015ht,Klein:2015ki}.
-Each droplet contains primers with a unique \emph{cell barcode} that tags all mRNAs from that droplet during reverse-transcription.
-Each primer also contains a \emph{unique molecular identifier (UMI)} that is appended to each mRNA molecule during reverse transcription.
-The 3' ends of the mRNAs are sequenced and mapped to the human and influenza virus transcriptomes to determine transcript identities.
-This information is combined with that provided by the UMIs and cell barcodes to quantify the number of molecules of each mRNA species that have been captured for each cell.
-
-Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
-We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
-Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
-We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
-Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
 \begin{figure}
@@
 
+\subsection{Strategy to measure mRNA in single virus-infected cells.}
+We performed single-cell mRNA sequencing using a droplet-based system that physically isolates individual cells prior to reverse transcription~\citep{zheng2017massively,macosko2015highly,Klein:2015ki}.
+Each droplet contains primers with a unique \emph{cell barcode} that tags all mRNAs from that droplet during reverse-transcription.
+Each primer also contains a \emph{unique molecular identifier (UMI)} that is appended to each mRNA molecule during reverse transcription.
+The 3' ends of the mRNAs are sequenced and mapped to the human and influenza virus transcriptomes to determine transcript identities.
+This information is combined with that provided by the UMIs and cell barcodes to quantify the number of molecules of each mRNA species that have been captured for each cell.
+
+Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
+We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
+Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
+We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
+Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
+
 We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
@@
 In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
-We confirmed that our viral stocks induced much less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
+We confirmed that our viral stocks induced less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
 
@@
 
-We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
-We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
-Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
-These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
-However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
-Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
-
-Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
-The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
-At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
-Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation at least in our low-MOI infections.
-
 \begin{figure}[t!]
@@
 
+We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
+We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
+Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
+These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
+However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
+Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
+
+Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
+The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
+At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
+Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.
+
 \subsection{Absence of viral genes partially explains cell-to-cell variability in viral load.}
@@
 {\bf (B)}
-Box plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
+Box and whisker plots showing the per-cell viral burden among cells with $>$0.5\% of their mRNA from virus, binned by whether or not the cells express each gene.
 A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
-Similar results are obtained if we examine only the 10-hour timepoint (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}).
-Top and bottom of box bound the first and third quartiles; the line in the box is at the median.
-Whiskers extend to the highest or lowest point within1.5$\times$ the interquartile range, and outliers are shown as circles.
 {\bf (C)}
-The fraction of that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
+The fraction of cells that express each of the four other genes among cells that express all RNP genes, as well as the fraction that express \emph{all} four of the other genes.
 The fraction that express all four genes is well predicted by simply multiplying the frequencies of cells that express each gene individually, indicating that gene absence is approximately independent across these genes.
@@
 The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
-Points are colored as in Figure~\ref{fig:fluburdenbyflugene}.
+Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
 }{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
@@
 The difference for NP is no longer statistically significant due to low counts of infected cells lacking NP, but the trend remains.
-We do not show statistical analyses for other samples, as the number of infected cells is too low for meaningful statistical analyses.
+We do not show statistical analyses for other samples, as the number of infected cells is too low.
 }{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
@@
 Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
-The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.
+%The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.
 
@@
 Because our sequencing captured all polyadenylated transcripts, we can examine whether there are prominent changes in the host-cell transcriptome in sub-populations of infected cells.
-Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,Crotta:2013ef}.
+Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
 %In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.8\linewidth]{figures/p_small_tsne_merge.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge.pdf}}
 \caption{\label{fig:tsne}
@@
 Cells colored by the number of detected transcripts from type I and III interferons (IFN).
-Only one cell has detectable interferon expression.
+Only one cell has detectable interferon expression (in red near the lower left of the plot).
 {\bf (E)}
@@
 Cells colored by whether they express the viral NS gene.
-The one interferon-positive cell is lacking NS, but many interferon-negative cells also lack NS.
+The one interferon-positive cell is lacking NS, but so are many interferon-negative cells.
 }
-
 \end{figure}
@@
 Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA. 
-\jdbcomment{scale shows?} }
+}
 \figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
@@
 
-The absence of specific viral genes in some cells only partially explains the cell-to-cell variation in amount of viral mRNA, which still varies from $<$1\% to $>$50\% among cells expressing all the viral genes.
+The absence of viral genes only partially explains the cell-to-cell variation in amount of viral mRNA, which still varies from $<$1\% to $>$50\% among cells expressing all the viral genes.
 It is likely that other viral genetic factors explain some of this remaining heterogeneity.
@@
 In particular, the viral load in individual cells is positively associated with the expression of genes involved in response to cellular stresses, including oxidative stress.
-It will be interesting to determine if these cellular transcriptional signatures are purely a consequence of the stress imposed by a large burden of viral replication, or if their stronger activation in some cells is a causative factor that promotes viral transcription.
-However, it would not be surprising if a substantial amount of the cell-to-cell heterogeneity can never be ascribed to pre-existing features of the viral genome or cellular state. 
+It will be interesting to determine if these cellular transcriptional signatures are simply a consequence of the stress imposed by a large burden of viral replication, or if their stronger activation in some cells is a causative factor that promotes viral transcription.
+However, it would not be surprising if a substantial amount of the cell-to-cell heterogeneity cannot be ascribed to pre-existing features of the viral genome or cellular state. 
 Apparently stochastic heterogeneity is a common feature of many processes that have been examined at a single-cell level~\citep{cai2006stochastic,raj2006stochastic,buganim2012single,shalek2013single,avraham2015pathogen} -- especially when those processes are initiated by very small numbers of initial molecules~\citep{elowitz2002stochastic}, as is the case for low-MOI viral infection.
@@
 One factor that rather surprisingly does \emph{not} appreciably contribute to the heterogeneity in our experiments is activation of innate-immune interferon pathways.
-Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of these cells fail to express the influenza-virus interferon antagonist NS1.
+Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of cells fail to express the influenza-virus interferon antagonist NS1.
 It is known that interferon activation is stochastic at the level of single cells in response to both synthetic ligands~\citep{shalek2013single,shalek2014single,bhushal2017cell} and actual infection~\citep{Rand:2012kl,PerezCidoncha:2014jr,avraham2015pathogen,Killip:2017ef}.
@@
 Interferon expression would surely be more common at later times or with a viral stock passaged at higher MOI, since paracrine interferon signaling~\citep{crotta2013type} and accumulation of defective viral particles enhance innate-immune detection~\citep{Tapia:2013kf,Lopez:2014en}.
-However, the early events of physiological influenza infection involve just a few infecting virions~\citep{varble2014influenza,mccrone2017evolutionary}, and so it is interesting to speculate whether rare stochastic interferon activation during the first few cycles of viral replication could contribute to heterogeneity in the eventual outcome of infection.
+However, the early events of physiological influenza infection involve just a few virions~\citep{varble2014influenza,mccrone2017evolutionary}, and so it is interesting to speculate whether rare events such as interferon activation during the first few cycles of viral replication could contribute to heterogeneity in the eventual outcome of infection.
 
 Interestingly, despite the extreme heterogeneity in the amount of total viral mRNA per cell, the relative levels of the various viral mRNAs are more consistent across cells.
-If we ignore the variation due to the fact that some cells completely fail to express specific viral genes, we find that the relative levels of the different viral mRNAs are similar across cells and generally reflective of classical bulk measurements~\citep{hatada1989control}.
-This fact suggests that despite the stochasticity that would seem inherent in initiating transcription and replication of each gene from a single copy carried by the incoming virion, if that gene is not absent or completely lost, then the virus possesses mechanisms to control relative gene expression~\citep{Shapiro:1987ur,hatada1989control,perez2010influenza,heldt2012modeling,chua2013influenza}.
-This fact also suggests that the sources of cell-to-cell heterogeneity in viral transcriptional load must exert relatively concordant effects on all viral gene segments.
+If we ignore the fact that some cells completely fail to express specific viral genes, we find that the relative levels of the different viral mRNAs are similar across cells and generally reflective of classical bulk measurements~\citep{hatada1989control}.
+This finding suggests that despite the stochasticity inherent in initiating transcription and replication of each gene from a single copy carried by the incoming virion, the virus possesses mechanisms to control relative gene expression~\citep{Shapiro:1987ur,hatada1989control,perez2010influenza,heldt2012modeling,chua2013influenza}.
+It also suggests that the sources of cell-to-cell heterogeneity in viral transcriptional load must exert relatively concordant effects on all viral genes.
 
 Overall, our work shows the power and importance of charactering cellular infection at the level of single cells~\citep{avraham2015pathogen}.
-The dynamics of viral infection in any given cell is shaped by the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes that are initiated by a single copy of the viral genome.
-Here we have shown how the confluence of these factors leads to extreme cell-to-cell heterogeneity in the transcriptional outcome of influenza virus infection.
+The dynamics of viral infection in any given cell is shaped by the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes initiated by a single copy of each viral gene.
+We have shown how the confluence of these factors leads to extreme cell-to-cell heterogeneity in the transcriptional outcome of influenza virus infection.
 We anticipate that further deconstruction of the contributions of each factor will enable a deeper understanding of how the bulk features of infection emerge from the heterogeneous processes occurring within individual virus-infected cells.
@@
 At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
-At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots at of the clarified viral supernatant -80 \si{\degreeCelsius }.
+At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots of the clarified viral supernatant at -80 \si{\degreeCelsius }.
 We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
@@
 \subsection{qPCR}
-For the qPCR in Figure~\ref{fig:viruspopulations} and Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells / well in a 6-well tissue culture plate in D10 the day prior to infection.
+For the qPCR in Figure~\ref{fig:viruspopulations} and Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells per well in a 6-well tissue culture plate in D10 the day prior to infection.
 On the day of infection, a single well was trypsinized and the cells were counted in order to determine the appropriate amount of virus to use to achieve the intended MOI.
@@
 RNA was purified using the QIAGEN RNeasy plus mini kit following manufacturer's instructions. 
-cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using manufacturer's directions. 
+cDNA was synthesized using an oligoDT primer and the SuperScript\textsuperscript{TM} III first-strand synthesis supermix from ThermoFisher using the manufacturer's protocol. 
 Transcript abundance was measured using SYBR\textsuperscript{TM} green PCR master mix, using a combined anneal/extension step of 60 \si{\degreeCelsius } for one minute with the following primers: \emph{HA}: 5'-GGCCCAACCACACATTCAAC-3', 5'-GCTCATCACTGCTAGACGGG-3', \emph{IFNB1}: 5'-AAACTCATGAGCAGTCTGCA-3', 5'-AGGAGATCTTCAGTTTCGGAGG-3', \emph{L32}: 5'-AGCTCCCAAAAATAGACGCAC-3', 5'-TTCATAGCAGTAGGCACAAAGG-3'.
@@
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Jupyter notebooks that perform all of the computational analyses are available at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}, and are also available in Supplementary~file~\ref{suppfile:code}.
+Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
+If you are a reviewer who needs access, please contact the editor.}.
 
@@
 For each sample, cell barcodes that had $>$2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}C).
+
 In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C).
@@
 The 10-hour sample has the most viral mRNA, and so is expected to have the most lysis-associated influenza reads.
-We therefore determined the threshold using just this sample (Figure~\ref{fig:viralbarcodes}C), and then applied this threshold to all samples.
+We therefore determined the threshold using just this sample, and then applied this threshold to all samples.
+The threshold is shown in Figure~\ref{fig:viralbarcodes}C.
 This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
@@
 The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the downstream analyses.
+Cells were classified as co-infected if both viral variants had an RNA level that exceeded the threshold, and if the minor variant contributed at least 5\% of the viral mRNA.
 
@@
 
-\nocite{*} % This command displays all refs in the bib file
+%\nocite{*} % This command displays all refs in the bib file
 \bibliography{references}

### 2017-09-15 paper/paper.tex
a few more revisions

@@
 \begin{abstract}
-Viral infection can cause large changes in a cell's transcriptome, but these changes have mostly been studied by bulk measurements on many cells.
-We use single-cell mRNA sequencing to examine the consequences of influenza virus infection in individual cells.
-We find extremely wide cell-to-cell variation in viral gene expression -- viral transcripts compose no more than a few percent of total mRNA in most infected cells, but a few cells derive over half their mRNA from virus.
-Some infected cells fail to express at least one viral gene, and this gene absence partially explains the variation in viral transcriptional load.
-Activation of innate immune pathways is rare, even in cells that fail to express influenza virus's major interferon antagonist.
-We identify some cellular genes that co-vary in abundance with the amount of viral mRNA.
-Overall, our results highlight extensive cell-to-cell heterogeneity in the transcriptional consequences of viral infection.
+Viral infection can dramatically alter a cell's transcriptome.
+However, these changes have mostly been studied by bulk measurements on many cells.
+Here we use single-cell mRNA sequencing to examine the transcriptional consequences of influenza virus infection.
+We find extremely wide cell-to-cell variation in production of viral gene transcripts -- viral transcripts compose less than a percent of total mRNA in many infected cells, but a few cells derive over half their mRNA from virus.
+Some infected cells fail to express at least one viral gene, and this gene absence partially explains variation in viral transcriptional load.
+Despite variation in total viral load, the relative abundances of viral mRNAs are fairly consistent across infected cells.
+Activation of innate immune pathways is rare, but some cellular genes co-vary in abundance with the amount of viral mRNA.
+Overall, our results highlight the complexity of viral infection at the level of single cells.
 \end{abstract}
@@
 For instance, cells infected with influenza virus at high multiplicity of infection (MOI) express an \emph{average} of 50,000 to 100,000 viral mRNAs per cell, corresponding to 5 to 25\% of all cellular mRNA~\citep{hatada1989control}.
-In addition, infection can trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
+Infection can also trigger innate-immune sensors that induce the expression of cellular anti-viral genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
 This anti-viral response is another prominent transcriptional signature of high-MOI influenza virus infection in bulk cells~\citep{Geiss:2002a}.
@@
 {\bf (C)}  
-For the single-cell mRNA sequencing, A549 cells were infected with an equal mixture of mutant and wild-type virus. 
+For the single-cell mRNA sequencing, A549 cells were infected with an equal mixture of wild-type and synonymously barcoded virus. 
 Immediately prior to collection, cells were physically separated into droplets and cDNA libraries were generated containing the indicated barcodes. 
@@
 Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
-This finding is consistent with prior work showing positive correlations among the abundance of several viral genes in individual cells~\citep{Heldt:2015iz}.
+This finding is consistent with prior work showing positive correlations among the abundance of several viral genome segments in individual cells~\citep{Heldt:2015iz}.
 
@@
 Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
-We then determined the fraction of cells that were HA positive and calculated the HA-expressing units assuming a Poisson infection process.
+We then determined the fraction of cells that were HA positive and calculated the HA-expressing units.
 

### 2017-09-22 paper/paper.tex
recolored tSNE plots

@@
 Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
+The scale bar indicates the fold-increase or decrease over the mean, truncated at 3-fold on both sides. 
 }
@@
 \section{Acknowledgments}
+We thank Xiaojie Qiu for helpful advice about use of the Monocle software package.
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.

### 2017-09-24 paper/paper.tex
revisions based on David Bacsik comments

@@
 We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
-For all samples except for the replicate 8-hour timepoint, we replaced the infection inoculum with fresh media after one hour, thereby ensuring that most infection was initiated during a narrow time window.
+We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
+The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
 We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
@@
 We determined a cutoff for calling ``true'' infections by fitting a curve to the mean barcode purity of all cells with greater than a given fraction of their mRNA derived from virus.
+The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
 We called the cutoff at the point at which purity stops increasing with the fraction of viral mRNA.
@@
 Viral transcripts are classified as \emph{syn} if they mapped to a synonymously barcoded influenza transcript, \emph{wt} if they mapped to a wild-type influenza transcript, \emph{invalid} if multiple reads for the same UMI differed on the status of the viral barcode, and as \emph{uncalled} if none of the reads for that UMI overlapped the region of the viral transcript containing the viral barcode.
+For calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
 }{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
@@
 Consistent with the idea that NS1 is important for suppressing interferon, the one interferon-positive cell lacked detectable expression of the NS gene.
-But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having $>$10\% of their mRNA derived from virus (Figure~\ref{fig:tsne}).
-Therefore, most infected cells fail to express interferon even when the virus lacks its major interferon antagonist.
-Our results are in line with other work showing that NS1-deficient influenza virus fails to deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
-Therefore, single cells infected at low MOI with a relatively ``pure'' stock of influenza virus only rarely activate innate-immune responses.
+But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having a substantial amount of their mRNA derived from virus (Figure~\ref{fig:tsne}).
+This result is in line with other work showing that NS1-deficient influenza virus does not deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
+Therefore, many individual infected cells fail to activate innate-immune responses even when the virus lacks its major interferon antagonist.
 
@@
 The absence of some viral genes in some infected cells has been noted previously~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the the total viral transcriptional load as a function of the level of each mRNA.
-We find that absence of any of the four genes that encode the viral RNP leads to a dramatic drop in the total amount of viral mRNA, consistent with prior bulk studies~\citep{Vreede:2004ip,eisfeld2015centre}.
+We find that cells lacking expression of any of the four genes that encode the viral RNP express much less total viral mRNA, consistent with prior bulk studies~\citep{Vreede:2004ip,eisfeld2015centre}.
 Interestingly, the \emph{reason} some cells fail to express some viral genes remains unclear.
@@
 However, it is also possible that much of the viral gene absence is due to stochastic loss of viral RNPs after infection but prior to the initiation of viral transcription in the nucleus.
-%Further work is needed to distinguish between these possibilities.
 
@@
 The 3'-end sequencing strategy used in our experiments detects the presence of a viral gene, but does not identify whether that gene contains a mutation that might hinder viral replication.
-However, viral mutations are also unlikely to explain all the observed heterogeneity, since the current consensus estimates of influenza virus's mutation rate suggest that the typical virion in a stock such as the one used in our experiment should contain less than one mutation per genome~\citep{parvin1986measurement,suarez1992heterogeneity,suarez1994estimation,nobusawa2006comparison,bloom2014experimentally,pauly2017novel}.
+However, viral mutations are also unlikely to explain all the observed heterogeneity, since current consensus estimates of influenza virus's mutation rate suggest that the typical virion in a stock such as the one used in our experiment should contain less than one mutation per genome~\citep{parvin1986measurement,suarez1992heterogeneity,suarez1994estimation,nobusawa2006comparison,bloom2014experimentally,pauly2017novel}.
 
 The rest of the heterogeneity must be due to some combination of cellular factors and inherent stochasticity.
-We do find some features of the cellular transcriptome that co-vary with the amount of influenza mRNA in individual cells. 
-In particular, the viral load in individual cells is positively associated with the expression of genes involved in response to cellular stresses, including oxidative stress.
-It will be interesting to determine if these cellular transcriptional signatures are simply a consequence of the stress imposed by a large burden of viral replication, or if their stronger activation in some cells is a causative factor that promotes viral transcription.
-However, it would not be surprising if a substantial amount of the cell-to-cell heterogeneity cannot be ascribed to pre-existing features of the viral genome or cellular state. 
-Apparently stochastic heterogeneity is a common feature of many processes that have been examined at a single-cell level~\citep{cai2006stochastic,raj2006stochastic,buganim2012single,shalek2013single,avraham2015pathogen} -- especially when those processes are initiated by very small numbers of initial molecules~\citep{elowitz2002stochastic}, as is the case for low-MOI viral infection.
+Some features of the cellular transcriptome co-vary with the amount of influenza mRNA. 
+In particular, the viral load in individual cells is associated with the expression of genes involved in response to cellular stresses, including oxidative stress.
+It will be interesting to determine if these cellular transcriptional signatures are simply a consequence of the stress imposed by viral replication, or if their stronger activation in some cells is a causative factor that promotes viral transcription.
+However, it also would not be surprising if a substantial amount of the cell-to-cell heterogeneity cannot be ascribed to pre-existing features of either the viral genome or cellular state. 
+Apparently stochastic heterogeneity is a common feature of many processes at a single-cell level~\citep{cai2006stochastic,raj2006stochastic,buganim2012single,shalek2013single,avraham2015pathogen} -- especially when those processes are initiated by very small numbers of initial molecules~\citep{elowitz2002stochastic}, as is the case for low-MOI viral infection.
 
-One factor that rather surprisingly does \emph{not} appreciably contribute to the heterogeneity in our experiments is activation of innate-immune interferon pathways.
+Our data do suggest that the factors driving the heterogeneity in viral transcriptional load exert relatively concordant effects on all viral genes in a given cell.
+Specifically, despite the extreme heterogeneity in total viral mRNA per cell, the relative levels of the viral mRNAs are reasonably consistent across cells, and generally reflective of classical bulk measurements~\citep{hatada1989control}.
+Therefore, despite the stochasticity inherent in initiating transcription and replication of each gene from a single copy carried by the incoming virion, as long as a gene is not completely lost then the virus possesses mechanisms to control its relative expression~\citep{Shapiro:1987ur,hatada1989control,perez2010influenza,heldt2012modeling,chua2013influenza}.
+
+One factor that surprisingly does \emph{not} appreciably contribute to the heterogeneity in our experiments is activation of innate-immune interferon pathways.
 Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of cells fail to express the influenza-virus interferon antagonist NS1.
@@
 
-Interestingly, despite the extreme heterogeneity in the amount of total viral mRNA per cell, the relative levels of the various viral mRNAs are more consistent across cells.
-If we ignore the fact that some cells completely fail to express specific viral genes, we find that the relative levels of the different viral mRNAs are similar across cells and generally reflective of classical bulk measurements~\citep{hatada1989control}.
-This finding suggests that despite the stochasticity inherent in initiating transcription and replication of each gene from a single copy carried by the incoming virion, the virus possesses mechanisms to control relative gene expression~\citep{Shapiro:1987ur,hatada1989control,perez2010influenza,heldt2012modeling,chua2013influenza}.
-It also suggests that the sources of cell-to-cell heterogeneity in viral transcriptional load must exert relatively concordant effects on all viral genes.
-
 Overall, our work shows the power and importance of charactering cellular infection at the level of single cells~\citep{avraham2015pathogen}.
@@
 We have shown how the confluence of these factors leads to extreme cell-to-cell heterogeneity in the transcriptional outcome of influenza virus infection.
-We anticipate that further deconstruction of the contributions of each factor will enable a deeper understanding of how the bulk features of infection emerge from the heterogeneous processes occurring within individual virus-infected cells.
+Further deconstruction of the contributions of each factor will enable a deeper understanding of how the bulk features of infection emerge from the processes occurring within individual virus-infected cells.
 
@@
 We used custom Python code to annotate the cells in this matrix by the number of viral reads that could be assigned to the wildtype and synonymously barcoded virus.
+Only about half of the viral reads overlapped the barcoded regions of the genes (Figure~\ref{fig:workflow}A) and could therefore be assigned to a viral barcode (Figure~\ref{fig:viralbarcodes}-Figures~supplement~\ref{figsupp:barcodescalled}).
+So for calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
 This annotated cell-gene matrix is in Supplementary~file~\ref{suppfile:cellgenematrix}.
@@
 \section{Acknowledgments}
-We thank Xiaojie Qiu for helpful advice about use of the Monocle software package.
-We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
+We thank Xiaojie Qiu for advice about use of the Monocle software package, David Bacsik and Robert Bradley for comments on the manuscript, and the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
 

### 2017-09-24 paper/paper.tex
updated fig 3 and eliminated cumul frac plot per Cole suggestion

@@
 \subsection{Single cells show an extremely wide range of expression of viral mRNA.}
-We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate of the 8-hour timepoint.
+We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
 We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
@@
 But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
-While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}C).
+While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B).
 
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_cell_mRNA_summary.pdf}}
+\centerline{\includegraphics[width=\linewidth]{figures/p_cell_mRNA_summary.pdf}}
 \caption{\label{fig:cells}
-There is a very wide distribution in the number of viral mRNAs detected per cell.
+There is a very wide distribution in the amount of viral mRNA per cell.
 {\bf (A)} 
@@
 {\bf (B)} 
-Cumulative distribution of the fraction of mRNA derived from influenza virus for each sample.
-For each point on the x-axis, the y-axis indicates the fraction of cells with at least that much of their mRNA from virus.
-{\bf (C)} 
-The number of cellular and viral mRNAs for each cell is plotted as a point.
+The number of cellular and viral mRNAs detected for each cell is plotted as a point.
 The blue lines show the overall distribution of the number of cellular mRNAs per cell.
-Cells outside the red lines were considered outliers (possibly derived from two cells per droplet), and were excluded from subsequent analyses.
+The orange rug plot at the left of each panel shows the distribution of the number of viral mRNAs per cell.
+Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
 }
@@
 A Jupyter notebook that performs these analyses is in Supplementary~file~\ref{suppfile:code}.
-For each sample, cell barcodes that had $>$2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}C).
+For each sample, cell barcodes that had $>$2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}B).
 

### 2017-09-24 paper/paper.tex
changes to figure 4 in response to Cole comments

@@
 At low levels of viral mRNA there is often a roughly equal mix, suggesting contamination with viral mRNAs leaked from lysed cells.
-At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single particle.
+At higher levels of viral mRNA, cells generally have only one viral variant, suggesting infection initiated by a single virion.
 A few cells are also obviously co-infected with both viral variants.
 {\bf (C)}
-We determined a cutoff for calling ``true'' infections by fitting a curve to the mean barcode purity of all cells with greater than a given fraction of their mRNA derived from virus.
+We determined a cutoff for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
 The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
-We called the cutoff at the point at which purity stops increasing with the fraction of viral mRNA.
+We fit a curve to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
+See the Methods for details.
 {\bf (D)}

### 2017-09-25 paper/paper.tex
other changes in response to Cole comments

@@
 Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
-%The rest of this variability must be due to additional mechanisms such as differences in host-cell state, viral mutations not detectable by our 3'-end sequencing approach, or inherent stochasticity.
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_aledit.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr.pdf}}
 \caption{\label{fig:fluexpr}
@@
 This finding is consistent with prior work showing positive correlations among the abundance of several viral genome segments in individual cells~\citep{Heldt:2015iz}.
+It is also reminiscent of the observation that total cellular mRNA varies among single cells as a function of factors such as volume that affect all mRNA species concordantly~\citep{padovan2015single}.
 
@@
 \caption{\label{fig:cellulargenes}
-Cellular genes that co-vary in expression along with the the amount of influenza virus mRNA in individual cells expressing all eight viral genes.
-Shown are all genes that are differentially expressed at a false discovery rate of 0.1. 
+Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
+The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
+Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
 Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
-The scale bar indicates the fold-increase or decrease over the mean, truncated at 3-fold on both sides. 
+The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
 }
@@
 The most obvious form of heterogeneity is the complete failure of some infected cells to express one or more viral genes, which we estimate occurs in about half the infected cells in our experiments.
-The absence of some viral genes in some infected cells has been noted previously~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the the total viral transcriptional load as a function of the level of each mRNA.
+The absence of some viral genes in some infected cells has been noted previously~\citep{Brooke:2013kb,Heldt:2015iz,Dou:2017cp}, and our work provides a holistic view by quantifying the total viral transcriptional load as a function of the level of each mRNA.
 We find that cells lacking expression of any of the four genes that encode the viral RNP express much less total viral mRNA, consistent with prior bulk studies~\citep{Vreede:2004ip,eisfeld2015centre}.
@@
 Samples were sequenced on an Illumina HiSeq. 
-The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{needs to be added.}
+The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Upload in progress.}
 

### 2017-09-25 paper/paper.tex
added Sarah Teichmann citation

@@
 The influenza virus genome consists of eight negative-sense RNA segments, and many infected cells fail to express one more of these RNAs~\citep{Heldt:2015iz,Dou:2017cp} or their encoded proteins~\citep{Brooke:2013kb}.
-In addition, activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
+In addition, activation of innate-immune responses is inherently stochastic~\citep{shalek2013single,shalek2014single,bhushal2017cell,hagai2017gene}, and only some influenza-infected cells express anti-viral interferon genes~\citep{PerezCidoncha:2014jr,Killip:2017ef}.
 However, the extent of cell-to-cell variation in these and other host and viral factors remains unclear, as does the association among them in individual infected cells.
@@
 %In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
-However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell}.
+However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell,hagai2017gene}.
 We therefore hypothesized that we might see two sub-populations of infected cells: one in which the interferon response inhibited viral transcription, and another in which the virus was able to express high levels of its mRNA by evading or blocking this response.
@@
 Only one of the hundreds of virus-infected cells expresses any detectable interferon, despite the fact that a number of cells fail to express the influenza-virus interferon antagonist NS1.
-It is known that interferon activation is stochastic at the level of single cells in response to both synthetic ligands~\citep{shalek2013single,shalek2014single,bhushal2017cell} and actual infection~\citep{Rand:2012kl,PerezCidoncha:2014jr,avraham2015pathogen,Killip:2017ef}.
+It is known that interferon activation is stochastic at the level of single cells in response to both synthetic ligands~\citep{shalek2013single,shalek2014single,bhushal2017cell,hagai2017gene} and actual infection~\citep{Rand:2012kl,PerezCidoncha:2014jr,avraham2015pathogen,Killip:2017ef}.
 But interferon expression is a prominent transcriptional signature of high-MOI influenza virus infection of bulk cells, including in the epithelial cell line and at the time-points used in our experiments~\citep{Geiss:2002a,sutejo2012activation}.

### 2017-09-26 paper/paper.tex
give manually edited figures separate names

@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_manualedits.pdf}}
 \caption{\label{fig:fluexpr}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge_manualarrow.pdf}}
 \caption{\label{fig:tsne}
@@
 Cells colored by the number of detected transcripts from type I and III interferons (IFN).
-Only one cell has detectable interferon expression (in red near the lower left of the plot).
+Only one cell has detectable interferon expression (in orange, indicated with arrow).
 {\bf (E)}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
+\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap_manualedit.pdf}}
 \caption{\label{fig:cellulargenes}

### 2017-09-26 paper/paper.tex
added citations

@@
 
-However, initiation of an actual influenza infection typically involves just a few virions infecting a few cells~\citep{varble2014influenza,mccrone2017evolutionary}. 
+However, initiation of an actual influenza infection typically involves just a few virions infecting a few cells~\citep{varble2014influenza,poon2016quantifying,leonard2017transmission,mccrone2017evolutionary}.
 The dynamics of viral infection in these individual cells may not mirror bulk measurements made on many cells infected at high MOI.
@@
 We took care to generate stocks of virus that were relatively ``pure'' of defective particles.
-Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk}.
+Stocks of viruses typically contain an array of biologically active viral particles, some of which are defective for replication owing to mutations or deletions in essential viral genes~\citep{von1954incomplete,huang1970defective,Brooke:2014ch,Fonville:2015cg,Lauring:2010if,Dimmock:2014dk,saira2013sequence}.
 These defective particles become prevalent when a virus is grown at high MOI, where complementation permits the growth of otherwise deleterious genotypes.

### 2017-09-26 paper/paper-biorxiv.tex
version for bioRxiv: remove eLife banner, added funding acknowledgments

--- 2017-09-26 paper/paper.tex
@@
 %%% PREAMBLE 
-\documentclass[9pt,lineno]{elife}
+\documentclass[9pt,lineno]{elife_biorxiv}
 \sloppy
@@
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}.
+Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code}.
 
@@
 We thank Xiaojie Qiu for advice about use of the Monocle software package, David Bacsik and Robert Bradley for comments on the manuscript, and the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
+This work was supported by NIH grant R01GM102198 (to JDB), NIH grant R01AI127893 (to JDB), NIH grant DP2HD088158 (to CT), a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant (to JDB), and a W.
+M.
+Keck Foundation grant (to CT).
+ABR is supported by a Damon Runyon Fellowship.
+JDB is partly supported by a Faculty Scholars Award from the Howard Hughes Medical Institute and the Simons Foundation.
+CT is partly supported by an Alfred P. Sloan Foundation Research Fellowship. 
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
 

### 2017-10-17 paper/paper.tex
eLife full submission

@@
 Here we use single-cell mRNA sequencing to examine the transcriptional consequences of influenza virus infection.
-We find extremely wide cell-to-cell variation in production of viral gene transcripts -- viral transcripts compose less than a percent of total mRNA in many infected cells, but a few cells derive over half their mRNA from virus.
-Some infected cells fail to express at least one viral gene, and this gene absence partially explains variation in viral transcriptional load.
-Despite variation in total viral load, the relative abundances of viral mRNAs are fairly consistent across infected cells.
+We find extremely wide cell-to-cell variation in the productivity of viral transcription -- viral transcripts comprise less than a percent of total mRNA in many infected cells, but a few cells derive over half their mRNA from virus.
+Some infected cells fail to express at least one viral gene, but this gene absence only partially explains variation in viral transcriptional load.
+Despite variation in viral load, the relative abundances of viral mRNAs are fairly consistent across infected cells.
 Activation of innate immune pathways is rare, but some cellular genes co-vary in abundance with the amount of viral mRNA.
@@
 }
-\figdata{The raw data are in \texttt{p\_flu\_expr.csv}.}
+\figdata{The raw data for all cells are in \texttt{p\_flu\_expr\_all.csv}.}
+\figdata{The raw data for fully infected cells are in \texttt{p\_flu\_expr\_fullyinfected.csv}.}
 \end{figure}
@@
 \figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
-\figdata{The results of a gene-set analysis are in \texttt{p\_sig\_cellular\_genes.csv}.}
+\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
 \figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
@@
 \caption{\label{suppfile:cellgenematrix}
-The annotated cell-gene matrix in Matrix Market Format.}
+The annotated cell-gene matrix in Matrix Market Format.
+\jdbcomment{This file is too large for the \textit{eLife} submission system.
+We will communicate with the editors to get it uploaded for a final accepted version, or we will post it on DataDryad.
+If you are a reviewer and need access, please contact the editor.}
+}
 \end{suppfile}

### 2018-01-03 paper/ResponseToReviews.tex
added document with reviews

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

\title{Response to reviews of ``Extreme heterogeneity of influenza virus infection in single cells'' for \textit{eLife}}
\author{Alistair B. Russell, Cole Trapnell, and Jesse D. Bloom}

\begin{document}
\maketitle

\emph{Below, the original comments {\color{blue} are in blue}, and our responses are in black.}

\color{blue}

\section*{SUMMARY}

Your manuscript describes the use of a novel single cell RNAseq method to examine heterogeneity in viral mRNA production within single cells infected with single influenza A virions. You report that viral mRNA levels, as a proportion of total cellular mRNA, can vary significantly between infected cells, and this variability cannot be simply explained by the activity of innate anti-viral defenses. Further, correlating the levels of specific host transcripts with viral transcript levels allows you to identify pathways that tend to correlate with viral levels.

The strengths of the paper include the description of a novel method for leveraging single cell cRNAseq to quantify heterogeneity and stochasticity in viral gene expression during infection. This is a really important question for the field, and the approaches detailed here will likely by adopted by other groups studying other viruses. The authors do a good job of accounting for many of the pitfalls inherent in the experimental execution and data analysis. The manuscript is well written, and does an especially effective job of visually presenting complex datasets.

The main shortcoming of the paper is that it provides very little in the way of new information. As the authors clearly and honestly point out, most of the findings in this paper are simply confirming observations made in older papers. Additionally, the findings are purely descriptive, and provide little insight into the mechanisms that may give rise to the observations.

We believe that addressing the points below would help ameliorate some of the shortcomings of the paper.

\section*{MAJOR COMMENTS}

1. There are some obvious sources of variability that haven't been suitably discussed.

- The first is variation in the timing of the early stages of infection. No steps appear to have been taken to synchronize infection, or to limit secondary spread of the virus within the cultures. Could heterogeneity in the timing of binding/entry/fusion/trafficking explain a lot of the variation observed?

- Another potential contributor is the cell cycle status of the infected cells. Did the host transcript data shed light on whether cell cycle status influenced viral transcription levels? This is especially important to address given that some of the genes showing association with the viral burden are cell-cycle related.

- In assessing to what extent lack of RNP expression accounts for the viral mRNA expression variability (page 6, paragraph 2), we think it is important to take into account potential extracellular contamination. Specifically, it would be more convincing if the analysis omitted cells with mixed wildtype/synonymous clones.

- As the variability of the viral RNA load appears to be the central result of the manuscript, it would be useful to:

    a. Employ a statistic to quantify the variability (e.g. entropy or Gini index)

    b. Use simple models to illustrate how much variability can be accounted for by simple effects, such as expected Poisson co-infection frequency, or the likelihood of attaining full complement of RNP genes.

2. While the silent tagging method used to address co-infection is clever and appreciated, the issue is not fully settled. The dismissal of co-infection as a factor influencing cell-to-cell variation in lines 127-9 is based on too few cells to draw any conclusions, and thus needs to be tempered. Also, there is likely to be a significant amount of cryptic co-infection with identical barcode viruses (expected to be similar to that of mixed barcodes, ~10\%) that could influence measured heterogeneity. These points should be made in the text.

3. The analysis of host cell transcripts positively or negatively associated with high viral mRNA expression is pretty minimal. Do the host genes identified here match up with the results of previous studies that screened for host pro- and anti-viral factors (reviewed in Watanabe et al. DOI: 10.1016/j.chom.2010.05.008)? Also, targeted gene knockdown or over-expression experiments could help establish the causality underlying these relationships.

4. All analysis of viral gene expression seems to be at the segment level. How do different transcripts expressed from the same gene segment compare (i.e. NS1 and NEP)?

5. The determination of the minimum required influenza fraction (Figure 4) is based on sound logic; however, we are concerned that the observed results do not fully align with this model. Specifically, while the 10hr experiment looks reasonable, the distributions of wildtype/synonymous mixed cells in other experiments do not appear to show the same trend (most notable for 6hr and 8hr samples, where almost no mixed cells are observed at low fractions). In that regard, using the 10hr dataset to estimate thresholds for all of the other samples does not seem appropriate.

\section*{MINOR COMMENTS}

- Fig 2: Normalizing high versus low DI stocks based on equal TCID50 to compare IFN induction is not appropriate since these stocks will differ in total particle number. In this experiment, the cells would receive a much higher overall dose of viral RNAs with the high DI stock compared to the others.

- Fig. 3B: While I appreciate the density of information presented by this figure, the authors' main point concerning heterogeneity in the viral proportion of total mRNA reads might be made more clearly by simply plotting the relative proportion of mRNAs that are viral per cell.

- Line 110: Should say "likely reflecting co-infection"

- Lines 186-8: This statement doesn't make much sense since you are talking about variation in mRNA proportion. Global effects such as cell size should not affect proportion.

- Lines 193-8: Is this really remarkable? As the authors point out, there is an extensive body of work demonstrating complementation during IAV infection.

- Line 329: "characterizing" misspelled.

- Lines 330-4: This statement about a confluence of factors leading to extreme cell-to-cell heterogeneity should be reworded since the data here actually doesn't show a clear role for innate immunity in promoting heterogeneity and there is no significant evaluation of the effect of host cell state on heterogeneity.

- The "replicate" 8-hr sample isn't really a replicate since a different MOI was used. Also, the purpose of the differences in treatment for this sample compared with the 8-hr sample are not well explained.

-Figure 3 is not effective in supporting the observations made in the text (that most cells have little viral RNA, and that there's a wide variability in the amount of viral RNA). Separate distributions for the viral RNA would be more telling. Log scale would be more appropriate for showing counts per cell.

-Figure 4C: the threshold line appears to be a bit to the right of the global maximum - going through the description, it was not clear why that should be the case.

-Why was media change skipped in the replicate experiment?

-Consider using different color schemes for Fig 4A and B, as different variables are being shown.

-Figure 5A legend: axis descriptions appear to be swapped

\end{document}

### 2018-01-03 paper/ResponseToReviews.tex
sketched outline to response

@@
 
-The strengths of the paper include the description of a novel method for leveraging single cell cRNAseq to quantify heterogeneity and stochasticity in viral gene expression during infection.
+The strengths of the paper include the description of a novel method for leveraging single-cell RNAseq to quantify heterogeneity and stochasticity in viral gene expression during infection.
 This is a really important question for the field, and the approaches detailed here will likely by adopted by other groups studying other viruses.
@@
 Additionally, the findings are purely descriptive, and provide little insight into the mechanisms that may give rise to the observations.
+
+{\color{black}
+Thank you for the excellent and fair summary of our work.
+We agree that the approach is powerful, and thank you for recognizing our efforts to fully describe pitfalls and effectively present the complex datasets.
+As you note, we agree that our paper does not provide any truly new or unexpected biological result.
+We do our best to honestly acknowledge this fact in the paper.
+However, we do believe that the paper provides before the most explicit and direct observations of phenomenon that have previously been mostly inferred from circumstantial evidence.
+These include the extreme variability in transcripts per cell, the rarity of interferon induction even in the absence of NS, the high rate of missing segments, and the role of co-infection in... 
+}
 
@@
 
+{\color{black}
+Our argument here would be that washed cells in 8-2 essentially do this.
+Alistair is generating data by flow with heterogeneity with washing +/- synchronization of infection.
+The thought is this will show that synchronization doesn't matter with respect to HA protein heterogeneity.
+This would be a new supporting figure.
+
+The co-infection barcodes provide good evidence against much secondary spread. We can explain that.
+}
+
 - Another potential contributor is the cell cycle status of the infected cells.
@@
 
+{\color{black}
+Cole had some suggestions of how we could do a few analyses that address this:
+
+This is a super tricky thing to do - I have not seen computational methods that I trust described in the literature for detecting genes that co-vary with cell cycle.
+Aviv has a method that seems reasonable at first blush but I have some concerns with in terms of the details.
+I have had better luck testing for genes that are differentially expressed between post-mitotic/quiescent and those that are actively cycling
+
+What would first do is add up the transcript counts for well known markers of actively proliferating cells (CCNB2 is my favorite, but I you could sum a bunch of cyclins and CDKs) to get a ?proliferation index?
+for each cell.
+Then look at the distribution of this - I would expect it to be lognormal for a population of cells that are robustly growing.
+For cells that are mostly quiescent (e.g.
+a confluent epithelial plate), I would guess the transcript levels for the cells would be low, maybe with a long tail to the right.
+Then you would add the proliferation index values as a new column in the pData table in Monocle.
+Then you can use the index as a variable in the model formula for differentialGeneTest.
+This will give you genes that positively correlate with the proliferation index.
+The trouble is establishing what values of the index really correspond to actively cycling cells.
+Ideally, you?d have a bimodal distribution and you?d be able to set some clear threshold, but in practice this doesn?t usually happen.
+In general, this is pretty tricky to do without built-in controls (e.g.
+sets of cells that are sorted on DNA content, arrested with some drug, or what have you).
+So I am generally incline to take these sort of analyses with a huge grain of salt.
+}
+
 - In assessing to what extent lack of RNP expression accounts for the viral mRNA expression variability (page 6, paragraph 2), we think it is important to take into account potential extracellular contamination.
 Specifically, it would be more convincing if the analysis omitted cells with mixed wildtype/synonymous clones.
+
+{\color{black}
+If we didn't already do this, we can do it easily --- it shouldn't have much of an effect because mixed cells are rare.
+
+We can potentially add a supplemental figure based on wildtype / HA-GFP co-infection.
+}
 
@@
 
+{\color{black}
+Gini index is something we could definitely calculate. Not clear what we could compare it to, but we can at least calculate it and report it.
+We then calculate Gini index under some simple model where everything is due to decreased transcription when missing segments.
+Any such model is probably pretty bogus and oversimplified, so we'd want to do this in a way where everything ends up in a heavily caveated supplement that doesn't bog down text.
+}
+
 2.
@@
 These points should be made in the text.
+
+{\color{black}
+We should make sure that we don't say co-infection is unimportant. We're just saying that low MOI, it doesn't explain everything. Probably this needs to be re-worded.
+}
 
@@
 
+{\color{black}
+In response, gently point out that this isn't actually the point.
+
+It's not that analysis is minimal, it's that there isn't anything.
+
+Then we can somehow work in the Watanabe reference.
+
+We have the oxidative stress thing to add as a figure supplement.
+}
+
 4. All analysis of viral gene expression seems to be at the segment level. How do different transcripts expressed from the same gene segment compare (i.e. NS1 and NEP)?
+
+{\color{black}
+We have to explain that we can't reliably distinguish NEP/NS1 and M1/M2 for technical reasons that also affect essentially all other single-cell RNAseq.
+}
 
@@
 
+{\color{black}
+Valid point, maybe we should look at sample-specific cutoffs.
+}
+
 \section*{MINOR COMMENTS}
@@
 
+{\color{black}
+Gently explain why this is not relevant for the point we're trying to make.
+}
+
 - Fig.
 3B: While I appreciate the density of information presented by this figure, the authors' main point concerning heterogeneity in the viral proportion of total mRNA reads might be made more clearly by simply plotting the relative proportion of mRNAs that are viral per cell.
+
+{\color{black}
+We can add suggested plot if it isn't already there, it might already be 4E.
+}
 
@@
 
+{\color{black}
+Good point that we'll address already in first comment.}
+
 -Figure 3 is not effective in supporting the observations made in the text (that most cells have little viral RNA, and that there's a wide variability in the amount of viral RNA).

### 2018-01-10 paper/ResponseToReviews.tex
finished setting up response letter and addressed a few of the most minor reviewer comments

@@
 
+We believe that addressing the points below would help ameliorate some of the shortcomings of the paper.
+
 {\color{black}
-Thank you for the excellent and fair summary of our work.
-We agree that the approach is powerful, and thank you for recognizing our efforts to fully describe pitfalls and effectively present the complex datasets.
-As you note, we agree that our paper does not provide any truly new or unexpected biological result.
-We do our best to honestly acknowledge this fact in the paper.
-However, we do believe that the paper provides before the most explicit and direct observations of phenomenon that have previously been mostly inferred from circumstantial evidence.
-These include the extreme variability in transcripts per cell, the rarity of interferon induction even in the absence of NS, the high rate of missing segments, and the role of co-infection in... 
+This is a fair summary of our work.
+We agree that our approach is powerful, and thank you for recognizing our efforts to fully describe pitfalls and effectively present the complex datasets.
+
+As you note, our approach does not uncover any truly new biological phenomenon, and (as you also note) we honestly describe how our results end up mostly confirming older observations made using more indirect methods.
+However, there is substantial value to explicitly observing these processes in single cells, even if the findings are mostly in line with existing circumstantial evidence from bulk studies or less comprehensive single-cell approaches (e.g., flow cytometry).
+Our explicit measurements of transcript counts in single cells provide a more definitive and quantitative description of influenza virus infection than prior inferences from other approaches.
+Also, the extremeness of the heterogeneity across cells is surprising even in light of all that was previously known.
+
+We have revised our manuscript in response to the many helpful suggestions made by the reviewers below.
+Below we detail these revisions.
+As you will see, we have been able to address all of the major questions and suggestions.
 }
-
-We believe that addressing the points below would help ameliorate some of the shortcomings of the paper.
 
@@
 
-{\color{black}
+{\color{red}
 Our argument here would be that washed cells in 8-2 essentially do this.
@@
 
-{\color{black}
+{\color{red}
 Cole had some suggestions of how we could do a few analyses that address this:
@@
 
-{\color{black}
+{\color{red}
 If we didn't already do this, we can do it easily --- it shouldn't have much of an effect because mixed cells are rare.
@@
 
-{\color{black}
+{\color{red}
 Gini index is something we could definitely calculate. Not clear what we could compare it to, but we can at least calculate it and report it.
@@
 
-{\color{black}
+{\color{red}
 We should make sure that we don't say co-infection is unimportant. We're just saying that low MOI, it doesn't explain everything. Probably this needs to be re-worded.
@@
 
-{\color{black}
+{\color{red}
 In response, gently point out that this isn't actually the point.
@@
 
-{\color{black}
+{\color{red}
 We have to explain that we can't reliably distinguish NEP/NS1 and M1/M2 for technical reasons that also affect essentially all other single-cell RNAseq.
@@
 
-{\color{black}
+{\color{red}
 Valid point, maybe we should look at sample-specific cutoffs.
@@
 
-{\color{black}
+{\color{red}
 Gently explain why this is not relevant for the point we're trying to make.
@@
 
-{\color{black}
+{\color{red}
 We can add suggested plot if it isn't already there, it might already be 4E.
@@
 
+{\color{black}
+Thank you for pointing this out, we have made this fix.}
+
 - Lines 186-8: This statement doesn't make much sense since you are talking about variation in mRNA proportion. Global effects such as cell size should not affect proportion.
+
+{\color{black}
+This is a good point. 
+Since the statement in question is tangential to our central conclusions, we have just removed the lines in question.
+}
 
@@
 - Line 329: "characterizing" misspelled.
+
+{\color{black}
+This typo has been fixed.
+}
 
@@
 
-{\color{black}
+{\color{red}
 Good point that we'll address already in first comment.}
@@
 
+{\color{black}
+Thank you for pointing out this error in the legend, it has been fixed.
+}
+
 \end{document}

### 2018-01-10 paper/paper.tex
finished setting up response letter and addressed a few of the most minor reviewer comments

@@
 However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
-Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection.
+Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, likely reflecting co-infection.
 
@@
 {\bf (A)} 
-Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
+The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
 Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
@@
 This finding is consistent with prior work showing positive correlations among the abundance of several viral genome segments in individual cells~\citep{Heldt:2015iz}.
-It is also reminiscent of the observation that total cellular mRNA varies among single cells as a function of factors such as volume that affect all mRNA species concordantly~\citep{padovan2015single}.
 
@@
 
-Overall, our work shows the power and importance of charactering cellular infection at the level of single cells~\citep{avraham2015pathogen}.
+Overall, our work shows the power and importance of characterizing cellular infection at the level of single cells~\citep{avraham2015pathogen}.
 The dynamics of viral infection in any given cell is shaped by the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes initiated by a single copy of each viral gene.

### 2018-01-11 paper/ResponseToReviews.tex
added stub responses to all points
changed Figure 4 color schemes per reviewer suggestion
explain about inability to distinguish spliced flu genes

@@
 
-{\color{red}
-We have to explain that we can't reliably distinguish NEP/NS1 and M1/M2 for technical reasons that also affect essentially all other single-cell RNAseq.
+{\color{black}
+This is an important point that we failed to adequately discuss in the original manuscript.
+As the reviews note, two of the influenza transcripts (M1 / M2  and NS1 / NS2) are generated from alternative splicing of gene segments.
+However, each pair of these transcripts share the same 3' end.
+Most current single-cell mRNA sequencing strategies, including the 10X platform that we use, can only sequence the 3' end of the transcript.
+The reason is that the cell barcode is appended to this end, and so Illumina sequencing can only extend to regions proximal to the 3' end if they are still to capture the barcode (as is required for single-cell mRNA sequencing).
+Therefore, these techniques cannot accurately identify alternative splicing -- this is simply an inherent limitation of the current techniques.
+We now clearly explain this fact in the following added text:
+\begin{quote}
+\textsl{
+Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively (Dubois et al, 2014).
+However, an inherent limitation of current established single-cell mRNA sequencing techniques is that they only sequence the 3' end of the transcript (Zheng et al., 2017; Macosko et al., 2015; Klein et al., 2015; Cao et al, 2017).
+Since the alternative spliceoforms M1 / M2 and NS1 / NEP share the same 3' ends, we cannot distinguish them and therefore will refer simply to the combined counts of transcripts from each of these alternatively spliced segments as the M and NS genes.}
+\end{quote}
 }
@@
 
+{\color{red} needs to be re-worded}
+
 - Line 329: "characterizing" misspelled.
@@
 - Lines 330-4: This statement about a confluence of factors leading to extreme cell-to-cell heterogeneity should be reworded since the data here actually doesn't show a clear role for innate immunity in promoting heterogeneity and there is no significant evaluation of the effect of host cell state on heterogeneity.
+
+{\color{red} needs to be re-worded}
 
@@
 
+{\color{red} Needs to be explaiend}
+
 -Figure 4C: the threshold line appears to be a bit to the right of the global maximum - going through the description, it was not clear why that should be the case.
+
+{\color{red} Needs to be explained}
 
@@
 
+{\color{red} Needs to be explained}
+
 -Consider using different color schemes for Fig 4A and B, as different variables are being shown.
+
+{\color{black} 
+This is a good suggestion.
+We now use different color schemes for Figures 4A and 4B.
+We also changed the color scheme for Figure 4E to be different than those for 4A and 4B since it also shows a different variable.}
 

### 2018-01-11 paper/paper.tex
explain about inability to distinguish spliced flu genes

@@
 
-We examined the total amount of viral mRNA versus the expression of each viral gene (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
-Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
+We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
+Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively~\citep{dubois2014influenza}.
+However, an inherent limitation of current established single-cell mRNA sequencing techniques is that they only sequence the 3' end of the transcript~\citep{zheng2017massively,macosko2015highly,Klein:2015ki,cao2017comprehensive}.
+Since the alternative spliceoforms M1 / M2 and NS1 / NEP share the same 3' ends, we cannot distinguish them and therefore will refer simply to the combined counts of transcripts from each of these alternatively spliced segments as the M and NS genes.
+
+Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
 However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.

### 2018-01-11 paper/ResponseToReviews.tex
added requested supplement to Fig 3 to show viral mRNA per cell distribution in another way

@@
 
-{\color{red}
-We can add suggested plot if it isn't already there, it might already be 4E.
+{\color{black}
+The reviewer points out that Figure 3B is a complicated way to show the heterogeneity in total mRNA.
+In response to this comment, we have also added a more straightforward presentation in Figure 3 - Figure supplement 1, which shows the distribution of the viral proportion as a cumulative fraction plot.
+This new plot should address the reviewer's point.
+
+Incidentally, the most obvious way to plot these data would initially simply seem to be a histogram or kernel density plot with fraction of mRNA from virus on the x-axis, and proportion of cells on the y-axis.
+However, such a plot turns out to be visually un-interpretable, because so much of the density is at zero (un-infected cells) that it is impossible to see any information for other parts of the plot for any reasonable y-axis scaling.
+This problem can be remedied by \emph{only} showing infected cells (getting rid of the huge bump in density at zero due to un-infected cells) -- but that can only be done after we have introduced a criterion to define which cells are infected, which doesn't happen until later in the text (and we want to show all the data before this censoring).
+Therefore, we also plot a more straightforward histogram display in Figure 4E for only the infected cells after censoring un-infected cells.
+We believe the combination of Figure 3B, the new Figure 3 - Figure - supplement 1, and Figure 4E effectively capture all the effects given these complexities of visual display for data distributed in such an asymmetric way.
 }

### 2018-01-11 paper/paper.tex
added requested supplement to Fig 3 to show viral mRNA per cell distribution in another way

@@
 {\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
-}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}
+}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}
+\end{figure}
 
@@
 We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
-As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B).
-Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B).
+As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
+Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
 But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
@@
 Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
+Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
 }
+\figsupp[Cumulative fraction plot of proportion of total mRNA from virus. \label{figsupp:viral_mRNA_cumulfrac}]{
+For each sample, this plot shows the fraction of all cells that derive at least the indication fraction of their mRNA from influenza virus.
+}{\includegraphics[width=0.5\linewidth]{figures/p_mRNA_per_cell_cumul.pdf}}
 \end{figure}

### 2018-01-11 paper/ResponseToReviews.tex
set up to add Figure 4 - Figure supplement 3, which shows HA / NS1 expression at high or low MOI

@@
 {\color{red}
-We should make sure that we don't say co-infection is unimportant. We're just saying that low MOI, it doesn't explain everything. Probably this needs to be re-worded.
+The text will be revised to ensure that we don't say co-infection is unimportant. We're just saying that low MOI, it doesn't explain everything.
+
+In addition, we will add the new Figure 4 - figure supplement 3 which shows HA or NS1 expression in cells infected at low or high MOI.
 }

### 2018-01-11 paper/paper.tex
set up to add Figure 4 - Figure supplement 3, which shows HA / NS1 expression at high or low MOI

@@
 }{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}
+\figsupp[Flow cytometry analysis of viral protein production of cells infected at low and high infectious dose. \label{fig:flowdose}]
+{
+Expression of a viral protein in cells infected at high or low initial infectious dose (MOIs of 0.1 or 5 \jdbcomment{based on what titering method, TCID50?}).
+While a higher dose leads to more cells expressing high amounts of viral protein, it does not substantially increase the amount of viral protein in either the low-expressing or high-expressing cells. 
+Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the distribution of where cells fall in the highly heterogeneous distribution of viral protein production.
+Cells dual-stained for NS1 and HA \jdbcomment{using what antibodies}.
+The panels at left show the distribution of expression of NS1 in cells identified as virally infected based on the HA staining, and the panels at right show the distribution of HA expression in cells identified as virally infected based on the NS1 staining.
+Data are shown for three independent replicates.
+}{\jdbcomment{needs to be added}}
 \end{figure}

### 2018-01-12 paper/paper.tex
Have added flow cytometry data for Figure 4 supplement 3. I decided to include coinfection but can remove if need be. Modified methods for NS1 staining and analysis

--- 2018-01-11 paper/paper.tex
@@
 \figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
-Full flow cytometry date for  Figure~\ref{fig:viruspopulations}B. 
+Full flow cytometry data for  Figure~\ref{fig:viruspopulations}B. 
 A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
@@
 }{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}
+
+\figsupp[Flow cytometry analysis of viral protein production of cells contingent on infectious dose or coinfection state. \label{figsupp:flowdose}]
+{
+(Top) Expression of a viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
+Cells were concurrently stained for HA and NS1 10 hours post infection.
+HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
+While a higher dose leads to more cells expressing high amounts of viral protein, it does not substantially increase the amount of viral protein in either the low-expressing or high-expressing cells. 
+Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the distribution of where cells fall in the highly heterogeneous distribution of viral protein production.
+(Bottom) Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
+At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry.
+Cells could be annotated as infected by one or more infectious particles (wild-type or pseudovirus alone) or at least two infectious particles (coinfection).
+Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
+Replicates shown.
+}{\includegraphics[width=\linewidth]{figures/Coinf.jpg}}
 \end{figure}
@@
 At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
-Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.
+Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.
 
@@
 \subsection{Flow cytometry titering and analyses}
-To determine viral titers in terms of HA-expressing units and for the flow cytometry in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
+To determine viral titers in terms of HA-expressing units and for the flow cytometry in Figure~\ref{fig:coexpression}, Figure~supplement~\ref{figsupp:flowcyto}, and Figure~supplement~\ref{figsupp:flowdose} A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
 Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
-After washing in PBS plus 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
+After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
 Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
 We then determined the fraction of cells that were HA positive and calculated the HA-expressing units.
-
-For Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
+For NS1 staining in Figure~supplement~\ref{figsupp:flowdose}, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
+washed, and analyzed.
+For Figure~\ref{fig:coexpression}, Figure~supplement~\ref{figsupp:flowcyto}, and  Figure~supplement~\ref{figsupp:flowdose}), after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
+For  Figure~supplement~\ref{figsupp:flowdose}) channels were additionally compensated in FLowJo.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
+For Figure~supplement~\ref{figsupp:flowdose}), the percentage of influenza-infected cells was determined by HA staining alone, and the top quantile of NS1-stained cells matching that percentage were taken as the NS1 positive population.
 

### 2018-01-12 paper/ResponseToReviews.tex
finished text updates about coinfection and viral production

--- 2018-01-11 paper/ResponseToReviews.tex
@@
 This is a fair summary of our work.
-We agree that our approach is powerful, and thank you for recognizing our efforts to fully describe pitfalls and effectively present the complex datasets.
+Thank you for recognizing the power of the approach and our efforts to fully describe pitfalls and effectively present the complex datasets.
 
@@
 
-{\color{red}
-We should make sure that we don't say co-infection is unimportant. We're just saying that low MOI, it doesn't explain everything. Probably this needs to be re-worded.
+{\color{black}
+These are very good points.
+We have added the following text that clearly draws out the fact that our approach will only identify half the co-infections:
+\begin{quote}
+\textsl{
+In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
+Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure 4D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
+}
+\end{quote}
+
+The reviewers are correct that our statement about the role of co-infection in influencing cell-to-cell variation was too strong.
+We have clarified this statement.
+We have also added a figure supplement (Figure 4-Figure supplement 3) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
+This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
+The new text reads:
+\begin{quote}
+\textsl{
+Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
+In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 3).
+This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 3).
+}
+\end{quote}
 }

### 2018-01-12 paper/paper.tex
finished text updates about coinfection and viral production

@@
 {
-(Top) Expression of a viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
-Cells were concurrently stained for HA and NS1 10 hours post infection.
+(Top) Expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
+Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
 HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
-While a higher dose leads to more cells expressing high amounts of viral protein, it does not substantially increase the amount of viral protein in either the low-expressing or high-expressing cells. 
-Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the distribution of where cells fall in the highly heterogeneous distribution of viral protein production.
+While a higher dose leads to more cells expressing high amounts of viral protein, it does not greatly increase the amount of viral protein in either the low-expressing or high-expressing cells. 
+Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the proportions of cells that fall in different parts of the highly heterogeneous distribution.
 (Bottom) Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
@@
 Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
-These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
-However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
+These annotation thresholds are conservative, and may miss some true low-level infections.
+However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
+In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
+Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure~\ref{fig:viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
 Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
@@
 At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
-Notably, co-infected cells do not express dramatically more viral mRNA than other cells (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.
+Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
+In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
+This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
 

### 2018-01-13 paper/paper.tex

@@
 {
-(Top) Expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
+{\bf (A)}
+Expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
 Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
@@
 Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the proportions of cells that fall in different parts of the highly heterogeneous distribution.
-(Bottom) Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment. 
+{\bf (B)}
+Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment at an MOI of 0.1 for each virus. 
 At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry.

### 2018-01-13 paper/paper.tex
added new threshold to call infected cells

@@
 Candidate marker genes for t-SNE dimensionality reduction were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
-The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
+The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 50 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
 

### 2018-01-13 paper/ResponseToReviews.tex
fixed threshold for infected cells

@@
 We have clarified this statement.
-We have also added a figure supplement (Figure 4-Figure supplement 3) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
+We have also added a figure supplement (Figure 4-Figure supplement 4) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
 This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
@@
 Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
-In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 3).
-This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 3).
+In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 4).
+This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 4).
 }
@@
 
-{\color{red}
-Valid point, maybe we should look at sample-specific cutoffs.
+{\color{black}
+This is a good point. 
+We now use the procedure in Figure 4C to call thresholds \emph{separately} for each sample as the reviewers suggest.
+The sample-specific calling is shown in Figure 4-Figure supplement 3.
+These new thresholds lead to a modest increase in the number of cells called as virally infected for the earlier timepoint samples.
+These newly called cells therefore slightly shift the quantitative values in all subsequent analyses, although the changes are very small and all results remain unchanged.
+
+The reason that the earlier timepoints have less mixed barcodes is because these earlier samples have much less total viral mRNA -- and when their is less viral mRNA in total, there is also less available to be acquired from lysed cells.
 }
@@
 
-{\color{red} Needs to be explained}
+{\color{black} 
+This is a good observation. 
+We were determining the threshold as the maximum of a loess fit to the data. 
+Previously, the loess fitting parameters caused the fit line to overshoot, putting the threshold a bit to the right.
+We have adjusted span parameter in the loess fitting, which fixes the overshoot.
+The revised Figure 4C shows the loess fit in orange as well as the threshold in dotted green -- and the threshold now goes right through the maximum.
+}
 

### 2018-01-13 paper/paper.tex
fixed threshold for infected cells

@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_frac_flu_summary.pdf}}
+\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}
 \caption{\label{fig:viralbarcodes}
@@
 {\bf (C)}
-We determined a cutoff for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
+We determined a threshold for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
 The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
-We fit a curve to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
+We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
+This plot illustrates the process for the 10-hour sample, see Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds} for similar plots for other samples.
 See the Methods for details.
@@
 }{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
+\figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
+Cell lysis can lead cells to the spurious association of small amounts of extraneous mRNA with individual cells.
+We wanted to avoid classifying as infected cells that had simply acquired such lysis-derived viral mRNA.
+The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
+As is shown in Figure~\ref{fig:viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
+For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
+We called the threshold at this point of maximum purity (dotted green line).
+For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as Figure~\ref{fig:viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
+Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (Figure~\ref{fig:cells}B).
+}{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}
 \figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
@@
 
-We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
+We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
 We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
@@
 The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
-At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
+At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
 Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
@@
 }{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
-\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
+\figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
 \end{figure}
@@
 We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
-At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C).
-The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C).
-If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 40-50\% of infected cells express mRNAs from all eight genes.
+At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C and Figure~\ref{fig:fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
+The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C and Figure~\ref{fig:fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
+If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 35-50\% of infected cells express mRNAs from all eight genes.
 This estimate of the frequency at which infected cells express mRNAs from all eight gene segments is slightly higher than previous estimates of 13\%~\citep{Brooke:2013kb} and 20\%~\citep{Dou:2017cp}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_manualedits.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr.pdf}}
 \caption{\label{fig:fluexpr}
+\jdbcomment{add expression wedges}
 Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge_manualarrow.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge.pdf}}
 \caption{\label{fig:tsne}
+\jdbcomment{add arrow}
 A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
@@
 We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
-We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).
+We identified 43 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).
 
 \begin{figure}
-\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap_manualedit.pdf}}
+\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
 \caption{\label{fig:cellulargenes}
+\jdbcomment{add gray arrow wedge}
 Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
@@
 
-In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C).
+In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}).
 We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
 This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
-The 10-hour sample has the most viral mRNA, and so is expected to have the most lysis-associated influenza reads.
-We therefore determined the threshold using just this sample, and then applied this threshold to all samples.
-The threshold is shown in Figure~\ref{fig:viralbarcodes}C.
+As is apparent from Figure~\ref{fig:viralbarcodes}B, only the 10-hour sample and the 8-hour-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
+The likely reason is that these samples have more total viral mRNA (and so there is more available mRNA to be acquired from lysed cells); in addition, there is always some experimental variability in the amount of cell lysis during the 10X sequencing process, and these samples may simply have the most.
+So the above threshold procedure is appropriate for those two samples.
+For the other samples, we simply set a minimum threshold of requiring at least a fraction 2$\times$$10^{-4}$ reads to come from viral mRNA as explained in the legend to Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
+The thresholds for each sample are shown in Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
 This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
@@
 
-%\nocite{*} % This command displays all refs in the bib file
 \bibliography{references}

### 2018-01-13 paper/ResponseToReviews.tex
responded to minor comment about Figure 3

@@
 
-{\color{red} Needs to be explaiend}
+{\color{black} 
+We have added Figure 3-Figure supplement 1, which uses a log scale to show the distribution of the amount of viral mRNA per cell.
+We still also retain Figure 3 to show the non-transformed data.
+Many cells have no viral mRNA, and it is not possible to show these non-infected cells on a logged scale, so the new combination of Figure 3 and Figure 3-supplement 1 provides both linear and log-scale views.
+Simple histograms or kernel density plots of the distribution of viral mRNA are not effective until we have removed non-infected cells (as is done in Figure 4E after setting the thresholds for calling infection) since the distribution is so strongly peaked at the low end.
+}
 

### 2018-01-13 paper/ResponseToReviews.tex
addressed minor comment about "Remarkably"

@@
 
-{\color{red} needs to be re-worded}
+{\color{black} 
+We have removed the word ``Remarkably'' and just pointed out that in some cells, co-infection is responsible for providing the full gene complement.
+This new text more clearly explains that it has long been inferred that co-infection can complement missing segments.
+Our contribution is to directly observe single cells where we can identify specific genes that come from different co-infecting viruses, thereby providing an explicit single-cell characterization of this phenomenon.
+}
 

### 2018-01-13 paper/paper.tex
addressed minor comment about "Remarkably"

@@
 Seven of these 10 cells expressed all eight viral genes.
-Remarkably, the majority (4 of 7) of these cells would \emph{not} have expressed all the viral genes in the absence of co-infection, since they have at least one gene exclusively derived from each viral variant.
+The majority (4 of 7) of these cells would \emph{not} have expressed all the viral genes in the absence of co-infection, since they have at least one gene exclusively derived from each viral variant.
 For instance, the cell with 11.2\% of its mRNA from virus in the upper right of Figure~\ref{fig:coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
-Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,brooke2014influenza,Fonville:2015cg,aguilera2017plaques}.
+Our data therefore provide the first direct single-cell observation of the fact that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,brooke2014influenza,Fonville:2015cg,aguilera2017plaques}.
 

### 2018-01-13 paper/ResponseToReviews.tex
clarified in response to minor comment about comparing IFN based equal TCID50 for different DI stocks

@@
 
-{\color{red}
-Gently explain why this is not relevant for the point we're trying to make.
+{\color{black}
+The reviewers are completely correct.
+High DI stocks have more physical virions per infectious unit (e.g., TCID50), as indeed our own data in Figure 2A,B show.
+Therefore, the fact that Figure 2C shows that the high DI stocks induce more IFN per TCID50 does \emph{not} imply that the high DI stocks are more IFN inducing per physical virion.
+
+We have clarified that we are \emph{only} using Figure 2C to show that the high DI stocks induce more IFN per TCID50, and that this could be due entirely to more physical virions.
+This is an important point for readers less versed in how viral passaging / DI content can influence IFN induction, since many studies simply report on the TCID50s used and provide no information on stock purity.
+This fact likely explains the discrepancy in IFN induction in our study and prior bulk studies (e.g., \textit{eLife}, 5:e18311) on A549 cells that used viruses passaged at high MOIs that almost certainly contained many DI particles.
+
+The revised text clearly explains how Figure 2C is comparing stocks with equal TCID50 but not equal physical virions. 
+Specifically, we have added the following statement to the legend for Figure 2C, and similar sentences in the manuscript main text:
+\begin{quote}
+\textsl{
+Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
+Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels (Figure 2A,B).}
+\end{quote}
 }

### 2018-01-13 paper/paper.tex
clarified in response to minor comment about comparing IFN based equal TCID50 for different DI stocks

@@
 We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (Figure~\ref{fig:viruspopulations}A) and to particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
-In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
-We confirmed that our viral stocks induced less interferon than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
+In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (Figure~\ref{fig:viruspopulations}A,B).
+We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
 
@@
 {\bf (C)} 
-Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
+Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
+Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels. 
 Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.

### 2018-01-13 paper/ResponseToReviews.tex
repeat Figure 5 analyses (RNP effect on viral expression) excluding co-infected cells

@@
 
-{\color{red}
-If we didn't already do this, we can do it easily --- it shouldn't have much of an effect because mixed cells are rare.
-
-We can potentially add a supplemental figure based on wildtype / HA-GFP co-infection.
+{\color{black}
+This is a good suggestion.
+We have re-performed the analyses omitting the co-infected cells with mixed viral barcodes.
+These new analyses are in Figure 5-Figure supplement 3 and Figure 5-Figure supplement 5.
+The results remain essentially unchanged from the ones analyzing the full set of cells that are shown in main panels of Figure 5A,B.
 }

### 2018-01-13 paper/paper.tex
repeat Figure 5 analyses (RNP effect on viral expression) excluding co-infected cells

@@
 Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
+Plots for individual samples are in Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}.
 {\bf (B)}
@@
 A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
+The trends are similar if we look only at the 10-hour sample (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}) or exclude known co-infected cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdenbyflugene_nocoinfection}).
 {\bf (C)}
@@
 \figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
-The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
+The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
 Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
 }{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
+\figsupp[Like panel (A), but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburden_nocoinfection}]{
+The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, excluding cells that were annotated as coinfected based on the presence of both viral barcodes.
+}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_no_coinfect.pdf}}
 \figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
@@
 }{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
+\figsupp[Like panel (B) but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburdenbyflugene_nocoinfection}]{
+All findings in Figure~\ref{fig:fluburdenbyflugene}B remain unchanged if we exclude cells called as coinfected based on the presence of mixed viral barcodes.
+}{\includegraphics[width=\linewidth]{figures/p_no_coinfect_flu_burden_flu_gene_test}}
 \figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
@@
 
-We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
+We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
 Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively~\citep{dubois2014influenza}.
@@
 
-Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
+Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
 However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
@@
 However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
-Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}).
+Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
 

### 2018-01-13 paper/ResponseToReviews.tex
clarified that 8hr-2 is not really a replicate, but done slightly differently

@@
 
-- The "replicate" 8-hr sample isn't really a replicate since a different MOI was used.
+- The ``replicate'' 8-hr sample isn't really a replicate since a different MOI was used.
 Also, the purpose of the differences in treatment for this sample compared with the 8-hr sample are not well explained.
 
-{\color{red}
-Good point that we'll address already in first comment.}
+{\color{black}
+This point is correct.
+The second 8-hour sample (8hr-2) was done slightly differently, since the cells were not washed and the infecting MOI was slightly less.
+We have clarified the text to emphasize that the 8hr-2 sample is \emph{not} simply a replicate.
+
+The purpose of the 8hr-2 sample was to assess whether timing of infection was a major contributor to heterogeneity.
+This sample is useful for looking at that question because the omission of the washing step means that infection for this sample can in principle occur at any time over the experimental window.
+This is now clearly explained in the text.
+For more details, see our response to MAJOR COMMENT \#1 above, which deals with the question of the role of infection timing / synchronization.
+}
 
@@
 
-{\color{red} Needs to be explained}
+{\color{black} 
+This has been clarified in the text.
+See the response to MINOR COMMENT three before this one, and to MAJOR COMMENT \#1.}
 

### 2018-01-13 paper/paper.tex
put pathway enrichment in sorted CSV rather than unsorted TSV
clarified that 8hr-2 is not really a replicate, but done slightly differently

@@
 \subsection{Single cells show an extremely wide range of expression of viral mRNA.}
-We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
-We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
-The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
+We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, performing two slightly different variants of the experiment for the 8-hour timepoint.
+For most of the samples, we replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
+However, for the second 8-hour sample (which we denote as ``8hr-2'' in the figures), we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
 We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
@@
 \figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
-\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
+\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.csv}.}
 \figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
@@
 We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
-To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
+To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus grow for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
 The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.
@@
 
-All time points except for the 8-hour replicate were prepared on the same day.
+All time points except for the second 8-hour sample (8hr-2) were prepared on the same day.
 For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
@@
 All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
-For our replicate 8-hour time point, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
+For 8hr-2 sample, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
 After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
@@
 This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
-As is apparent from Figure~\ref{fig:viralbarcodes}B, only the 10-hour sample and the 8-hour-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
+As is apparent from Figure~\ref{fig:viralbarcodes}B, only the 10-hour sample and the 8hr-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
 The likely reason is that these samples have more total viral mRNA (and so there is more available mRNA to be acquired from lysed cells); in addition, there is always some experimental variability in the amount of cell lysis during the 10X sequencing process, and these samples may simply have the most.

### 2018-01-13 paper/ResponseToReviews.tex
added Gini coefficients to quantify unevenness

@@
 
+{\color{black}
+The idea of quantifying the variability in viral mRNA across infected cells is a good one.
+We have done this using the Gini coefficient.
+The results are in Figure 4--Figure supplement 4, and are referred to in the text related to that figure.
+As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceeds the Gini coefficient for the unevenness of income distribution in the United States.
+}
+
 {\color{red}
-Gini index is something we could definitely calculate. Not clear what we could compare it to, but we can at least calculate it and report it.
-We then calculate Gini index under some simple model where everything is due to decreased transcription when missing segments.
-Any such model is probably pretty bogus and oversimplified, so we'd want to do this in a way where everything ends up in a heavily caveated supplement that doesn't bog down text.
+Still need to think about a simple model.
 }

### 2018-01-13 paper/paper.tex
added link to raw data on GEO
added Gini coefficients to quantify unevenness

@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}
+\centerline{\includegraphics[width=0.83\linewidth]{figures/p_frac_flu_summary.pdf}}
 \caption{\label{fig:viralbarcodes}
@@
 Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
+Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
 }
@@
 }{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}
-\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
+\figsupp[Fraction of total viral mRNA from a given fraction of infected cells, and associated Gini coefficients.\label{figsupp:cumulfracflu}]{
 The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
-For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.
+For instance, the plot for the 8hrs sample shows that $\sim$50\% of all viral mRNA is derived from $\sim$8\% of the infected cells.
+The facet titles above each plot also give the Gini coefficient~\citep{gini1921measurement} that calculates the heterogeneity in the distribution of viral mRNA among infected cells.
+Gini coefficients of 0 indicate a perfectly even distribution across cells, and Gini coefficients of 1 indicate a maximally skewed distribution.
 }{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}
@@
 At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
+One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\cite{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
+Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
+The Gini coefficients are $\ge$0.64 for all samples.
+As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.
+
 Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
@@
 Samples were sequenced on an Illumina HiSeq.
-The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Upload in progress.}
 
@@
 The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}).
+The aligned deep sequencing data are available on the GEO repository under accession GSE108041 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE108041}).
+
 CellRanger calls cells based on the number of observed cell barcodes, and creates a cell-gene matrix.

### 2018-01-14 paper/ResponseToReviews.tex
added new supp fig and response related to infection timing

@@
 
-{\color{red}
-Our argument here would be that washed cells in 8-2 essentially do this.
-Alistair is generating data by flow with heterogeneity with washing +/- synchronization of infection.
-The thought is this will show that synchronization doesn't matter with respect to HA protein heterogeneity.
-This would be a new supporting figure.
+{\color{black}
+This is an important point.
+We did not synchronize infection, although for our 8-hour samples, we washed the cells one hour after infection for one sample (8hr) but not the other (8hr-2) in order to help evaluate the importance of infection timing.
+The heterogeneity was similar between these two samples, as demonstrated by the new quantification method (Gini coefficients) suggested by the reviewers in a later comment.
 
-The co-infection barcodes provide good evidence against much secondary spread. We can explain that.
+We have also performed new experiments where we infected cells with synchronized infections by pre-binding virus on ice, washing it away, and then raising the temperature (this is a previously described procedure in the influenza virus literature).
+Analysis of this experiment by flow cytometry staining is shown in the new Figure 4-Figure supplement 4, and shows that these synchronized infections look very similar to ones that are neither synchronized nor washed in terms of viral protein expression.
+
+Finally, the barcoded viruses (Figure 4B) provide strong evidence against secondary infection, since secondary infection would tend to lead to mixed-barcode cells because during secondary infections there would be many of progeny viruses of both barcodes.
+
+In addition to adding Figure 4-Figure supplement 4, we have added the following text related to this point:
+
+\begin{quote}
+\textsl{
+One possible source of heterogeneity in the amount of viral mRNA per cell is variability in the timing of infection.
+If some cells are infected earlier in the experiment than others, then they might have substantially more viral mRNA.
+However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
+First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (Figure 4E, Figure 4-Figure~supplement~3), despite the fact that the potential time window for infection is much longer in the former sample.
+Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C.
+As shown in Figure 4-Figure~supplement 4, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
+Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since Figure 4B does not show the pervasive presence of mixed barcodes that would occur in this case.
+Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.
+}
+\end{quote}
 }
@@
 We have done this using the Gini coefficient.
-The results are in Figure 4--Figure supplement 4, and are referred to in the text related to that figure.
+The results are in Figure 4--Figure supplement 3, and are referred to in the text related to that figure.
 As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceeds the Gini coefficient for the unevenness of income distribution in the United States.
@@
 We have clarified this statement.
-We have also added a figure supplement (Figure 4-Figure supplement 4) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
+We have also added a figure supplement (Figure 4-Figure supplement 5) which uses an independent method (flow cytometry staining of individual viral proteins) to support the fact that even at high infectious dose both low- and high-producing cells remain, although their relative proportions change.
 This figure supplement also supports our clarified statement that viral mRNA (or protein) production is not a simple continuous function of infectious dose, but is also affected by other sources of heterogeneity.
@@
 Notably, Figure 4E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
-In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 4).
-This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 4).
+In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure 4-Figure supplement 5).
+This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure 4-Figure supplement 5).
 }

### 2018-01-14 paper/paper.tex
added new supp fig and response related to infection timing

@@
 However, for the second 8-hour sample (which we denote as ``8hr-2'' in the figures), we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
+The rationale for including a sample without a media change was to determine the importance of synchronicity of the timing of infection as discussed later in this subsection.
+
 We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
 As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
-Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
+Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B, Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
 But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
-While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B).
+While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B, Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
 
@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.83\linewidth]{figures/p_frac_flu_summary.pdf}}
+\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}
 \caption{\label{fig:viralbarcodes}
@@
 }
+
 \figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
@@
 }{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
+
 \figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
@@
 }{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}
-\figsupp[Fraction of total viral mRNA from a given fraction of infected cells, and associated Gini coefficients.\label{figsupp:cumulfracflu}]{
+
+\figsupp[Cumulative distributions of viral mRNA per cell and Gini coefficients.\label{figsupp:cumulfracflu}]{
 The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
@@
 
-\figsupp[Flow cytometry analysis of viral protein production of cells contingent on infectious dose or coinfection state. \label{figsupp:flowdose}]
+\figsupp[Synchronization of infection does not greatly affect heterogeneity. \label{figsupp:synchronize}]
+{\jdbcomment{Alistair, can you add caption}
+}{\jdbcomment{Alistair, can you add add figure}}
+
+\figsupp[Effects of infectious dose or coinfection state. \label{figsupp:flowdose}]
 {
 {\bf (A)}
-Expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
+Flow cytometry analysis of expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
 Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
@@
 {\bf (B)}
-Cells were co-infected with a mix of wild-type virus and virus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment at an MOI of 0.1 for each virus. 
-At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry.
-Cells could be annotated as infected by one or more infectious particles (wild-type or pseudovirus alone) or at least two infectious particles (coinfection).
+Cells were co-infected with a mix of wild-type virus and pseudovirus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment at an MOI of 0.1 for each virus. 
+At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry for these proteins and GFP.
+Cells could be annotated as infected by virions of the same type (wild-type infection indicated by presence of HA, or pseudovirus infection indicated by the presence of GFP) or both types of virions (indicated by presence of HA and GFP).
 Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
@@
 At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
-One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\cite{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
+One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\citep{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
 Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
@@
 As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.
+
+One possible source of heterogeneity in the amount of viral mRNA per cell is variability in the timing of infection.
+If some cells are infected earlier in the experiment than others, then they might have substantially more viral mRNA.
+However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
+First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (Figure~\ref{fig:viralbarcodes}E, Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
+Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\jdbcomment{Alistair, can you add relevant citation for this method?}.
+As shown in Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
+Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since Figure~\ref{fig:viralbarcodes}B does not show the pervasive presence of mixed barcodes that would occur in this case.
+Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.
 

### 2018-01-14 paper/ResponseToReviews.tex
added material about cell-cycle and more analysis of cellular genes

@@
 
-{\color{red}
-Cole had some suggestions of how we could do a few analyses that address this:
-
-This is a super tricky thing to do - I have not seen computational methods that I trust described in the literature for detecting genes that co-vary with cell cycle.
-Aviv has a method that seems reasonable at first blush but I have some concerns with in terms of the details.
-I have had better luck testing for genes that are differentially expressed between post-mitotic/quiescent and those that are actively cycling
-
-What would first do is add up the transcript counts for well known markers of actively proliferating cells (CCNB2 is my favorite, but I you could sum a bunch of cyclins and CDKs) to get a ?proliferation index?
-for each cell.
-Then look at the distribution of this - I would expect it to be lognormal for a population of cells that are robustly growing.
-For cells that are mostly quiescent (e.g.
-a confluent epithelial plate), I would guess the transcript levels for the cells would be low, maybe with a long tail to the right.
-Then you would add the proliferation index values as a new column in the pData table in Monocle.
-Then you can use the index as a variable in the model formula for differentialGeneTest.
-This will give you genes that positively correlate with the proliferation index.
-The trouble is establishing what values of the index really correspond to actively cycling cells.
-Ideally, you?d have a bimodal distribution and you?d be able to set some clear threshold, but in practice this doesn?t usually happen.
-In general, this is pretty tricky to do without built-in controls (e.g.
-sets of cells that are sorted on DNA content, arrested with some drug, or what have you).
-So I am generally incline to take these sort of analyses with a huge grain of salt.
+{\color{black}
+We have expanded the analysis to specifically look at cell-cycle related genes.
+A gene-set analysis (Figure 9-Figure source data 2) finds that the G2-M cell-cycle checkpoint is associated with viral mRNA, with the expression of genes in this pathway typically expressed at lower levels in cells with more viral mRNA.
+This analysis is now discussed in the second-to-last paragraph of the Results.
 }
@@
 
-{\color{red}
-In response, gently point out that this isn't actually the point.
+{\color{black}
+We have substantially expanded the last section of the Results that focuses on cellular gene expression.
+We call out the gene-set enrichments, which show that genes involved in oxidative stress responses are at higher levels in cells with more viral mRNA.
+The possible pro-viral role of transient oxidative stress has been reported before, and we have performed new experiments (Figure 9-Figure supplement 2) that directly support it is pro-viral, suggesting causality.
 
-It's not that analysis is minimal, it's that there isn't anything.
+We also compare the genes that we find significantly associated with viral mRNA to the 128 genes highlighted by Watanabe et al (2010) as having been found multiple times in genome-wide studies.
+There is no overlap.
+This fact is discussed in the last paragraph of the Results.
+We theorize that this is because most of the cell-to-cell heterogeneity during infection of ``normal'' (e.g., not genetically perturbed) cells is \emph{not} due to changes in expression of the type of large-effect genes that are typically identified in genome-wide screens.
+Rather, our results would seem to suggest that a substantial amount of the heterogeneity is due to viral / stochastic factors (e.g., failure to express certain viral genes), or perhaps more subtle changes in cellular state such as transient oxidative stress.
 
-Then we can somehow work in the Watanabe reference.
-
-We have the oxidative stress thing to add as a figure supplement.
+Importantly, we do \emph{not} believe that the main value of our approach is as a sort of single-cell gene knockdown screen to identify pro- or anti-viral genes.
+Our goal is to figure out the actual extent and sources of cell-to-cell heterogeneity, which may not be due to primarily alterations in the expression level of single host genes with large effects.
 }

### 2018-01-14 paper/paper.tex
added material about cell-cycle and more analysis of cellular genes

@@
 Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
-%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
 However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell,hagai2017gene}.
@@
 We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
-We identified 43 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}).
+We identified 43 cellular genes that co-varied with viral mRNA expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}, Figure~\ref{fig:cellulargenes}-source~data~\ref{figdata:cellulargenes}).
 
-\begin{figure}
-\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
+\begin{figure}[t!]
+\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap.pdf}}
 \caption{\label{fig:cellulargenes}
@@
 }
-\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
-\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.csv}.}
+
+\figdata{The full results of the differential expression test are in \texttt{p\_sig\_cellular\_genes.csv}. \label{figdata:cellulargenes}}
+
+\figdata{A gene-set analysis for pathways associated with the amount of viral mRNA is in \texttt{p\_pathway\_enrichment.csv}. \label{figdata:geneset}}
+
 \figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
@@
 }{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
+
+\figsupp[Pre-treating to induce oxidative stress increases the fraction of infected cells expressing detectable viral protein. \label{figsupp:oxidative}]
+{
+\jdbcomment{Alistair, can you add figure and caption showing the increase in the percent of flu positive cells after treatment.
+It should just be possible to put the methods for this experiment in the figure caption.}
+}
+{\jdbcomment{add figure}}
+
 \end{figure}
 
-Many of the genes with increased expression in cells with more viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
+A gene-set analysis shows that many cellular genes that are associated with the amount of viral mRNA are involved in the response to reactive oxygen species or hypoxia (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figdata:geneset}).
+Genes known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress are often expressed at higher levels in cells with more viral mRNA (Figure~\ref{fig:cellulargenes}).
 These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}).
@@
 The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
-Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
-%Overall, the expression of some of the host genes in Figure~\ref{fig:cellulargenes}, along with inherent stochasticity and specific mutations to viral genes in individual cells, probably explain most of the variation in viral mRNA levels observed across the infected cells in our dataset.
+To directly test the effect of transient oxidative stress, we compared the fraction of cells that expressed detectable viral protein when infected either with or without pre-treatment to induce oxidative stress.
+Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:oxidative} shows that the cells pre-treated to induce oxidative stress exhibited more frequent detectable expression of viral protein.
+These results, in conjunction with the differential expression test in Figure~\ref{fig:cellulargenes} and the prior work mentioned above, suggest that oxidative stress acts in a proviral fashion.
+
+The gene-set analysis also found that the amount of viral mRNA was associated with the expression of genes involved in the G2-M cell-cycle checkpoint (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figdata:geneset}).
+The cell-cycle associated genes CCND3, MKI67, UBE2S, and CENPF are all expressed at significantly lower levels in cells with more viral mRNA (Figure~\ref{fig:cellulargenes}).
+However, our data are not sufficient to determine whether the lower expression of these genes is a cause or effect of the reduction in viral mRNA.
+
+Interestingly, none of the cellular genes that are significantly associated with the amount of viral mRNA in our study are among the 128 genes that \citet{watanabe2010cellular} report as having been identified multiple times in genome-wide screens for factors affecting influenza virus replication.
+One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral factors, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.
 
@@
 Figure~\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
+We performed the gene set analysis using the P -alues from the Monocle differentialGeneTest with \texttt{piano}~\citep{varemo2013enriching} using the hallmark gene set from GSEA v6~\citep{subramanian2005gene} and Fisher's method.
 

### 2018-01-14 paper/ResponseToReviews.tex
some small tweaks to response wording

@@
 However, there is substantial value to explicitly observing these processes in single cells, even if the findings are mostly in line with existing circumstantial evidence from bulk studies or less comprehensive single-cell approaches (e.g., flow cytometry).
-Our explicit measurements of transcript counts in single cells provide a more definitive and quantitative description of influenza virus infection than prior inferences from other approaches.
 Also, the extremeness of the heterogeneity across cells is surprising even in light of all that was previously known.
 
-We have revised our manuscript in response to the many helpful suggestions made by the reviewers below.
-Below we detail these revisions.
-As you will see, we have been able to address all of the major questions and suggestions.
+We have revised our manuscript in response to the many helpful suggestions from the reviewers.
+As you will see, we have addressed all the major questions and suggestions.
 }
@@
 This is an important point.
-We did not synchronize infection, although for our 8-hour samples, we washed the cells one hour after infection for one sample (8hr) but not the other (8hr-2) in order to help evaluate the importance of infection timing.
+We did not synchronize infection, although for our 8-hour samples, we washed the cells one hour after infection for one sample (8hr) but not the other (8hr-2) in order to evaluate the importance of infection timing.
 The heterogeneity was similar between these two samples, as demonstrated by the new quantification method (Gini coefficients) suggested by the reviewers in a later comment.
 
-We have also performed new experiments where we infected cells with synchronized infections by pre-binding virus on ice, washing it away, and then raising the temperature (this is a previously described procedure in the influenza virus literature).
-Analysis of this experiment by flow cytometry staining is shown in the new Figure 4-Figure supplement 4, and shows that these synchronized infections look very similar to ones that are neither synchronized nor washed in terms of viral protein expression.
+We have also performed new experiments where we synchronized infection by pre-binding virus on ice, washing it away, and then raising the temperature.
+Analysis of this experiment by flow cytometry staining is in the new Figure 4-Figure supplement 4, and shows that these synchronized infections look similar to ones that are neither synchronized nor washed.
 
-Finally, the barcoded viruses (Figure 4B) provide strong evidence against secondary infection, since secondary infection would tend to lead to mixed-barcode cells because during secondary infections there would be many of progeny viruses of both barcodes.
+Finally, the barcoded viruses (Figure 4B) provide strong evidence against secondary infection, since secondary infection would tend to lead to mixed-barcode cells because there would be many of progeny viruses of both barcodes.
 
@@
 The results are in Figure 4--Figure supplement 3, and are referred to in the text related to that figure.
-As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceeds the Gini coefficient for the unevenness of income distribution in the United States.
+As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceed the Gini coefficient for the unevenness of income distribution in the United States.
 }
@@
 We call out the gene-set enrichments, which show that genes involved in oxidative stress responses are at higher levels in cells with more viral mRNA.
-The possible pro-viral role of transient oxidative stress has been reported before, and we have performed new experiments (Figure 9-Figure supplement 2) that directly support it is pro-viral, suggesting causality.
+The possible pro-viral role of transient oxidative stress has been reported before, and we have performed new experiments (Figure 9-Figure supplement 2) that directly support the idea that it is pro-viral.
+These new experiments therefore suggest causality for one of the host-cell factors that we identify.
 
@@
 
-Importantly, we do \emph{not} believe that the main value of our approach is as a sort of single-cell gene knockdown screen to identify pro- or anti-viral genes.
+Importantly, we do \emph{not} believe that the main value of our approach is as a method to identify pro- or anti-viral genes.
 Our goal is to figure out the actual extent and sources of cell-to-cell heterogeneity, which may not be due to primarily alterations in the expression level of single host genes with large effects.
@@
 The reviewer points out that Figure 3B is a complicated way to show the heterogeneity in total mRNA.
-In response to this comment, we have also added a more straightforward presentation in Figure 3 - Figure supplement 1, which shows the distribution of the viral proportion as a cumulative fraction plot.
-This new plot should address the reviewer's point.
+In response to this comment, we have added Figure 3 - Figure supplement 1, which shows the distribution of the viral proportion as a cumulative fraction plot.
 

### 2018-01-14 paper/paper.tex
updated to new figure referencing enabled by new _eLife_ LaTex class file, this cleans up the LaTex source a bit

@@
 Infected cells will express viral as well as cellular mRNAs -- however the cell barcodes and UMIs cannot distinguish whether a cell was initially infected by one or multiple viral particles.
-We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (Figure~\ref{fig:workflow}A).
-Critically, these synonymous mutations did not greatly impact viral growth kinetics (Figure~\ref{fig:workflow}B).
+We therefore engineered an influenza virus (strain A/WSN/1933) that additionally carried \emph{viral barcodes} consisting of synonymous mutations near the 3' end of each transcript (\FIG{workflow}A).
+Critically, these synonymous mutations did not greatly impact viral growth kinetics (\FIG{workflow}B).
 We infected A549 human lung carcinoma cells with an equal mix of the wild-type and synonymously barcoded viruses.
-Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (Figure~\ref{fig:workflow}C).
+Cells infected by a single virion will exclusively express mRNAs from either wild-type or synonymously barcoded virus, whereas cells that are co-infected with multiple virions will often express mRNAs from both the wild-type and synonymously barcoded viruses (\FIG{workflow}C).
 
@@
 To minimize the levels of defective particles, we propagated our viral stocks at low MOI for a relatively brief period of time~\citep{Xue:2016bw}.
-We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (Figure~\ref{fig:viruspopulations}A) and to particles capable of inducing expression of a single viral protein (Figure~\ref{fig:viruspopulations}B).
-In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (Figure~\ref{fig:viruspopulations}A,B).
-We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (Figure~\ref{fig:viruspopulations}C).
+We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (\FIG{viruspopulations}A) and to particles capable of inducing expression of a single viral protein (\FIG{viruspopulations}B).
+In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (\FIG{viruspopulations}A,B).
+We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (\FIG{viruspopulations}C).
 
@@
 \figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
-Full flow cytometry data for  Figure~\ref{fig:viruspopulations}B. 
+Full flow cytometry data for  \FIG{viruspopulations}B. 
 A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
@@
 
-We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
-As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
-Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B, Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
+We recovered between 3,000 and 4,000 cells for each sample (\FIG{cells}A). 
+As expected for a low-MOI infection, most cells expressed little or no viral mRNA (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
+Also as expected, the amount of viral mRNA per cell among infected cells increased over time (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
 But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
-While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B, Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
+While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
 
@@
 Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
-Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
+\FIGSUPP[cells]{viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
 }
@@
 However, most cells are infected by at most one virion, and so the reads from truly infected cells will usually derive almost entirely from one of the two viral variants. 
-Figure~\ref{fig:viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and Figure~\ref{fig:viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
+\FIG{viralbarcodes}A shows the fraction of viral reads in individual cells from each viral variant, and \FIG{viralbarcodes}B indicates the fraction of viral reads from the most abundant variant in that cell.
 Most cells with large amounts of viral mRNA have viral transcripts exclusively derived from one viral variant -- indicating non-random partitioning as expected from viral infection.
@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.85\linewidth]{figures/p_frac_flu_summary.pdf}}
+\centerline{\includegraphics[width=0.82\linewidth]{figures/p_frac_flu_summary.pdf}}
 \caption{\label{fig:viralbarcodes}
@@
 We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
-This plot illustrates the process for the 10-hour sample, see Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds} for similar plots for other samples.
+This plot illustrates the process for the 10-hour sample, see \FIGSUPP[viralbarcodes]{thresholds} for similar plots for other samples.
 See the Methods for details.
@@
 Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
-Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
+\FIGSUPP[viralbarcodes]{cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
 }
@@
 The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
-As is shown in Figure~\ref{fig:viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
+As is shown in \FIG{viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
 For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
 We called the threshold at this point of maximum purity (dotted green line).
-For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as Figure~\ref{fig:viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
-Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (Figure~\ref{fig:cells}B).
+For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as \FIG{viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
+Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (\FIG{cells}B).
 }{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}
@@
 
-We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
+We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (\FIG{viralbarcodes}C, \FIGSUPP[viralbarcodes]{thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
 We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
-Figure~\ref{fig:viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
+\FIG{viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
 These annotation thresholds are conservative, and may miss some true low-level infections.
@@
 In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
-Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (Figure~\ref{fig:viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
-Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in Figure~\ref{fig:viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
+Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (\FIG{viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
+Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in \FIG{viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
 
-Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (Figure~\ref{fig:viralbarcodes}E). 
+Strikingly, the extreme variation in the number of viral transcripts per cell remains even after we apply these rigorous criteria for annotating infected cells (\FIG{viralbarcodes}E). 
 The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
-At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}).
+At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (\FIGSUPP[viralbarcodes]{cumulfracflu}).
 One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\citep{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
-Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
+\FIGSUPP[viralbarcodes]{cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
 The Gini coefficients are $\ge$0.64 for all samples.
@@
 However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
-First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (Figure~\ref{fig:viralbarcodes}E, Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
+First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (\FIG{viralbarcodes}E, \FIGSUPP[viralbarcodes]{cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
 Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\jdbcomment{Alistair, can you add relevant citation for this method?}.
-As shown in Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
-Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since Figure~\ref{fig:viralbarcodes}B does not show the pervasive presence of mixed barcodes that would occur in this case.
+As shown in \FIGSUPP[viralbarcodes]{synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
+Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since \FIG{viralbarcodes}B does not show the pervasive presence of mixed barcodes that would occur in this case.
 Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.
 
-Notably, Figure~\ref{fig:viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
-In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
-This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:flowdose}).
+Notably, \FIG{viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
+In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (\FIGSUPP[viralbarcodes]{flowdose}).
+This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (\FIGSUPP[viralbarcodes]{flowdose}).
 
@@
 Each viral gene segment is associated with one RNP in incoming infecting virions, but secondary transcription by newly synthesized RNPs requires the presence of the viral genes encoding each of the four RNP proteins~\citep{Vreede:2004ip,eisfeld2015centre}.
-This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}).
+This secondary transcription is a major source of viral mRNAs, as evidenced by the fact that blocking synthesis of the RNP proteins reduces the amount of viral mRNA by several orders of magnitude in bulk cells (\FIGSUPP[fluburdenbyflugene]{cyclohexamide}).
 
@@
 Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
-Plots for individual samples are in Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}.
+Plots for individual samples are in \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}.
 {\bf (B)}
@@
 A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
-The trends are similar if we look only at the 10-hour sample (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:10hrfluburdenbyflugene}) or exclude known co-infected cells (Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdenbyflugene_nocoinfection}).
+The trends are similar if we look only at the 10-hour sample (\FIGSUPP[fluburdenbyflugene]{10hrfluburdenbyflugene}) or exclude known co-infected cells (\FIGSUPP[fluburdenbyflugene]{fluburdenbyflugene_nocoinfection}).
 {\bf (C)}
@@
 }
+
 \figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
@@
 }{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
+
 \figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
 The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
-Points are colored as in Figure~\ref{fig:fluburdenbyflugene}A.
+Points are colored as in \FIG{fluburdenbyflugene}A.
 }{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
+
 \figsupp[Like panel (A), but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburden_nocoinfection}]{
@@
 }{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_no_coinfect.pdf}}
+
 \figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
@@
 }{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
+
 \figsupp[Like panel (B) but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburdenbyflugene_nocoinfection}]{
-All findings in Figure~\ref{fig:fluburdenbyflugene}B remain unchanged if we exclude cells called as coinfected based on the presence of mixed viral barcodes.
+All findings in \FIG{fluburdenbyflugene}B remain unchanged if we exclude cells called as coinfected based on the presence of mixed viral barcodes.
 }{\includegraphics[width=\linewidth]{figures/p_no_coinfect_flu_burden_flu_gene_test}}
+
 \figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
+
 \end{figure}
 
-We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
+We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
 Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively~\citep{dubois2014influenza}.
@@
 
-Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
+Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
 However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
-These results are statistically supported by Figure~\ref{fig:fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA.
+These results are statistically supported by \FIG{fluburdenbyflugene}B, which shows that absence of any RNP gene but \emph{not} any other viral gene is associated with reduced amounts of viral mRNA. 
 However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
-Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (Figure~\ref{fig:fluburdenbyflugene}A, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburdensamplebreakdown}, Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:fluburden_nocoinfection}).
+Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
 
@@
 We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
-At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (Figure~\ref{fig:fluburdenbyflugene}C and Figure~\ref{fig:fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
-The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (Figure~\ref{fig:fluburdenbyflugene}C and Figure~\ref{fig:fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
+At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (\FIG{fluburdenbyflugene}C, \FIG{fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
+The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (\FIG{fluburdenbyflugene}C and \FIG{fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
 If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 35-50\% of infected cells express mRNAs from all eight genes.
@@
 
-In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (Figure~\ref{fig:fluexpr}A).
-Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (Figure~\ref{fig:fluexpr}B).
+In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (\FIG{fluexpr}A).
+Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (\FIG{fluexpr}B).
 The relative levels of each viral mRNA in our cells are similar to prior bulk measurements made by Northern blots~\citep{hatada1989control}, which also found an expression hierarchy of M $>$ NS $\gg$ NP $>$ NA $>$ HA $\gg$ PB2 $\sim$ PB1 $\sim$ PA.
-The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (Figure~\ref{fig:fluexpr}C,D).
+The cell-to-cell consistency in the relative expression of different viral genes is even tighter if we limit the analysis only to cells that express all eight viral genes (\FIG{fluexpr}C,D).
 Therefore, with the exception of complete gene absence, the factors that drive the dramatic cell-to-cell variability in the amount of viral mRNA have roughly similar effects on all viral genes in a given cell.
@@
 Our sequencing enables us to identify the rare cells that were co-infected with both wild-type and synonymously barcoded viral variants.
-Overall, we captured 10 such co-infected cells that had $>$5\% of their mRNA derived from virus (Figure~\ref{fig:coexpression}).
+Overall, we captured 10 such co-infected cells that had $>$5\% of their mRNA derived from virus (\FIG{coexpression}).
 Seven of these 10 cells expressed all eight viral genes.
 The majority (4 of 7) of these cells would \emph{not} have expressed all the viral genes in the absence of co-infection, since they have at least one gene exclusively derived from each viral variant.
-For instance, the cell with 11.2\% of its mRNA from virus in the upper right of Figure~\ref{fig:coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
+For instance, the cell with 11.2\% of its mRNA from virus in the upper right of \FIG{coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
 Our data therefore provide the first direct single-cell observation of the fact that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,brooke2014influenza,Fonville:2015cg,aguilera2017plaques}.
@@
 \end{figure}
-Another observation from Figure~\ref{fig:coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
+
+Another observation from \FIG{coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
 This observation is consistent with the finding by \citet{Dou:2017cp} and \citet{Huang:2008gy} that the temporal window for co-infection is short -- if both viral variants infect a cell at about the same time, then neither will have a headstart and so each will have a roughly equal opportunity to transcribe its genes.
@@
 To support this idea with a larger dataset albeit at lower resolution, we generated a virus in which the HA coding sequence was replaced by GFP.
-We then co-infected cells with a mix of wildtype and $\Delta$HA-GFP virus and used flow cytometry to score cells for the presence of HA only (infection by wildtype virus), GFP only (infection by $\Delta$HA-GFP virus), or both (co-infection) as shown in Figure~\ref{fig:coexpression}~Figure~supplement~\ref{figsupp:flowcyto}.
+We then co-infected cells with a mix of wildtype and $\Delta$HA-GFP virus and used flow cytometry to score cells for the presence of HA only (infection by wildtype virus), GFP only (infection by $\Delta$HA-GFP virus), or both (co-infection) as shown in \FIGSUPP[coexpression]{flowcyto}.
 As in our single-cell sequencing data, we found that expression of HA and GFP were highly correlated, indicating that co-infected cells typically expressed roughly equal amounts of transcript from each viral variant.
@@
 To examine whether there were distinct sub-populations of virus-infected cells, we used a semi-supervised t-SNE approach~\citep{VanderMaaten:2008tm} to cluster cells by genes that co-varied with viral infection status.
-As shown in Figure~\ref{fig:tsne}A,B, this approach effectively grouped cells by the amount of viral mRNA that they expressed.
-Sample-to-sample variation was regressed away during the clustering, as cells did not obviously group by time-point, with expected exception that the uninfected and 6-hour samples had few cells in the region of the plot corresponding to large amounts of viral mRNA (Figure~\ref{fig:tsne}C).
+As shown in \FIG{tsne}A,B, this approach effectively grouped cells by the amount of viral mRNA that they expressed.
+Sample-to-sample variation was regressed away during the clustering, as cells did not obviously group by time-point, with expected exception that the uninfected and 6-hour samples had few cells in the region of the plot corresponding to large amounts of viral mRNA (\FIG{tsne}C).
 
@@
 To investigate further, we annotated each cell by the total number of type I and III interferon transcripts detected.
-Remarkably, only a single cell expressed detectable interferon (Figure~\ref{fig:tsne}D).
+Remarkably, only a single cell expressed detectable interferon (\FIG{tsne}D).
 We also examined interferon-stimulated genes, which are induced by autocrine and paracrine interferon signaling.
-Figure~\ref{fig:tsne}E shows expression of one such gene, IFIT1~\citep{Fensterl:2011fp}.
+\FIG{tsne}E shows expression of one such gene, IFIT1~\citep{Fensterl:2011fp}.
 As with interferon itself, expression of IFIT1 was rare and most prominent in the single interferon-positive cell, presumably due to the higher efficiency of autocrine versus paracrine signaling.
-Notably, interferon and interferon-stimulated genes were also relatively ineffective at blocking viral transcription in the single cell in which they were potently induced, since $>$10\% of the mRNA in this cell was derived from virus (Figure~\ref{fig:tsne}A,B,D,E).
+Notably, interferon and interferon-stimulated genes were also relatively ineffective at blocking viral transcription in the single cell in which they were potently induced, since $>$10\% of the mRNA in this cell was derived from virus (\FIG{tsne}A,B,D,E).
 
 We posited that the paucity of interferon induction might be due to the activity of influenza virus's major interferon antagonist, the NS1 protein~\citep{garcia1998influenza,hale2008multifunctional}.
-We therefore identified cells that expressed substantial amounts of viral mRNA but lacked the NS gene (Figure~\ref{fig:tsne}F).
+We therefore identified cells that expressed substantial amounts of viral mRNA but lacked the NS gene (\FIG{tsne}F).
 Consistent with the idea that NS1 is important for suppressing interferon, the one interferon-positive cell lacked detectable expression of the NS gene.
-But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having a substantial amount of their mRNA derived from virus (Figure~\ref{fig:tsne}).
+But other cells that lacked NS expression still failed to induce a detectable interferon response, despite often having a substantial amount of their mRNA derived from virus (\FIG{tsne}).
 This result is in line with other work showing that NS1-deficient influenza virus does not deterministically induce interferon~\citep{Killip:2017ef,Kallfass:2013kp}.
@@
 We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
-We identified 43 cellular genes that co-varied with viral mRNA expression at a false discovery rate of 0.1 (Figure~\ref{fig:cellulargenes}, Figure~\ref{fig:cellulargenes}-source~data~\ref{figdata:cellulargenes}).
+We identified 43 cellular genes that co-varied with viral mRNA expression at a false discovery rate of 0.1 (\FIG{cellulargenes}, \FIG{cellulargenes}-source~data~\ref{figdata:cellulargenes}).
 
@@
 \figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
-Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
+Table delineating genes in \FIG{cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
 }{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
@@
 
-A gene-set analysis shows that many cellular genes that are associated with the amount of viral mRNA are involved in the response to reactive oxygen species or hypoxia (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figdata:geneset}).
-Genes known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress are often expressed at higher levels in cells with more viral mRNA (Figure~\ref{fig:cellulargenes}).
-These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:ROStable}).
+A gene-set analysis shows that many cellular genes that are associated with the amount of viral mRNA are involved in the response to reactive oxygen species or hypoxia (\FIG{cellulargenes}-source~data~\ref{figdata:geneset}).
+Genes known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress are often expressed at higher levels in cells with more viral mRNA (\FIG{cellulargenes}).
+These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (\FIGSUPP[cellulargenes]{ROStable}).
 We additionally see reduced expression of the nitric oxide synthase interacting protein (NOSIP). 
@@
 To directly test the effect of transient oxidative stress, we compared the fraction of cells that expressed detectable viral protein when infected either with or without pre-treatment to induce oxidative stress.
-Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figsupp:oxidative} shows that the cells pre-treated to induce oxidative stress exhibited more frequent detectable expression of viral protein.
-These results, in conjunction with the differential expression test in Figure~\ref{fig:cellulargenes} and the prior work mentioned above, suggest that oxidative stress acts in a proviral fashion.
+\FIGSUPP[cellulargenes]{oxidative} shows that the cells pre-treated to induce oxidative stress exhibited more frequent detectable expression of viral protein.
+These results, in conjunction with the differential expression test in \FIG{cellulargenes} and the prior work mentioned above, suggest that oxidative stress acts in a proviral fashion.
 
-The gene-set analysis also found that the amount of viral mRNA was associated with the expression of genes involved in the G2-M cell-cycle checkpoint (Figure~\ref{fig:cellulargenes}-Figure~supplement~\ref{figdata:geneset}).
-The cell-cycle associated genes CCND3, MKI67, UBE2S, and CENPF are all expressed at significantly lower levels in cells with more viral mRNA (Figure~\ref{fig:cellulargenes}).
+The gene-set analysis also found that the amount of viral mRNA was associated with the expression of genes involved in the G2-M cell-cycle checkpoint (\FIG{cellulargenes}-source~data~\ref{figdata:geneset}).
+The cell-cycle associated genes CCND3, MKI67, UBE2S, and CENPF are all expressed at significantly lower levels in cells with more viral mRNA (\FIG{cellulargenes}).
 However, our data are not sufficient to determine whether the lower expression of these genes is a cause or effect of the reduction in viral mRNA.
@@
 Wildtype A/WSN/1933 (H1N1) influenza virus was generated by reverse genetics using the plasmids pHW181-PB2, pHW182-PB1, pHW183-PA, pHW184-HA, pHW185-NP, pHW186-NA, pHW187-M, and pHW188-NS~\citep{hoffmann2000dna}.
-The sequences of the viral RNAs encoded in these plasmids are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
+The sequences of the viral RNAs encoded in these plasmids are in \FIG{workflow}-source~data~\ref{figdata:viralsequences}.
 Reverse-genetics plasmids encoding the synonymously barcoded WSN virus were created by using site-directed mutagenesis to introduce two synonymous mutations near the 3' end of the mRNA for each viral gene.
-The sequences of the synonymously barcoded viral RNAs are in Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}.
+The sequences of the synonymously barcoded viral RNAs are in \FIG{workflow}-source~data~\ref{figdata:viralsequences}.
 
@@
 To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus grow for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
-The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.
+The high-MOI passage (high-defective particle) stock used in \FIG{viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.
 
-For the experiments in Figure~\ref{fig:coexpression}-Figure~supplement~\ref{figsupp:flowcyto}, we created a virus that carried an HA gene segment in which GFP replaced most of the HA coding sequence, following a scheme first described by \citet{marsh2007specific}.
-Briefly, we created a plasmid encoding a viral RNA with GFP in place of the HA coding sequence in the context of the pHH21~\citep{Neumann:1999ws} reverse-genetics plasmid, removing potential start codons upstream of the GFP (see Figure~\ref{fig:coexpression}-source~data~\ref{figdata:HAflank} for the sequence of the viral RNA).
+For the experiments in \FIGSUPP[coexpression]{flowcyto}, we created a virus that carried an HA gene segment in which GFP replaced most of the HA coding sequence, following a scheme first described by \citet{marsh2007specific}.
+Briefly, we created a plasmid encoding a viral RNA with GFP in place of the HA coding sequence in the context of the pHH21~\citep{Neumann:1999ws} reverse-genetics plasmid, removing potential start codons upstream of the GFP (see \FIG{coexpression}-source~data~\ref{figdata:HAflank} for the sequence of the viral RNA).
 We then generated GFP-carrying virus by reverse-genetics in cells constitutively expressing HA~\citep{Doud:2016gm}.
@@
 \subsection{qPCR}
-For the qPCR in Figure~\ref{fig:viruspopulations} and Figure~\ref{fig:fluburdenbyflugene}-Figure~supplement~\ref{figsupp:cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells per well in a 6-well tissue culture plate in D10 the day prior to infection.
+For the qPCR in \FIG{viruspopulations} and \FIGSUPP[fluburdenbyflugene]{cyclohexamide}, A549 cells were seeded at 3$\times$10\textsuperscript{5} cells per well in a 6-well tissue culture plate in D10 the day prior to infection.
 On the day of infection, a single well was trypsinized and the cells were counted in order to determine the appropriate amount of virus to use to achieve the intended MOI.
@@
 
-For the measurements of viral genomic HA content in Figure~\ref{fig:viruspopulations}A, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol.
+For the measurements of viral genomic HA content in \FIG{viruspopulations}A, vRNA was harvested from 80 \si{\micro}l of viral supernatant by the addition of 600 \si{\micro}l of RLT plus before proceeding with the standard QIAGEN RNeasy Plus Mini kit protocol.
 The cDNA was generated using SuperScript\textsuperscript{TM} III first-strand synthesis supermix using the manufacturer's protocol, and using the universal vRNA primers of \citet{Hoffmann:2001vj} with the modifications described in  \citet{Xue:2017dl}.
@@
 \subsection{Flow cytometry titering and analyses}
-To determine viral titers in terms of HA-expressing units and for the flow cytometry in Figure~\ref{fig:coexpression}, Figure~supplement~\ref{figsupp:flowcyto}, and Figure~supplement~\ref{figsupp:flowdose} A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
+To determine viral titers in terms of HA-expressing units and for the flow cytometry in \FIGSUPP[coexpression]{flowcyto} and \FIGSUPP[viralbarcodes]{flowdose}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
 Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
@@
 We then determined the fraction of cells that were HA positive and calculated the HA-expressing units.
-For NS1 staining in Figure~supplement~\ref{figsupp:flowdose}, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
+For NS1 staining in \FIGSUPP[viralbarcodes]{flowdose}, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
 washed, and analyzed.
-For Figure~\ref{fig:coexpression}, Figure~supplement~\ref{figsupp:flowcyto}, and  Figure~supplement~\ref{figsupp:flowdose}), after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
-For  Figure~supplement~\ref{figsupp:flowdose}) channels were additionally compensated in FLowJo.
+For \FIGSUPP[coexpression]{flowcyto} and \FIGSUPP[viralbarcodes]{flowdose}, after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
+For \FIGSUPP[viralbarcodes]{flowdose} channels were additionally compensated in FLowJo.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
-For Figure~supplement~\ref{figsupp:flowdose}), the percentage of influenza-infected cells was determined by HA staining alone, and the top quantile of NS1-stained cells matching that percentage were taken as the NS1 positive population.
+For \FIGSUPP[viralbarcodes]{flowdose}, the percentage of influenza-infected cells was determined by HA staining alone, and the top quantile of NS1-stained cells matching that percentage were taken as the NS1 positive population.
 
@@
 The human transcriptome was generated by filtering genome assembly GRCh38 for protein coding genes defined in the GTF file GRCh38.87.
-The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (Figure~\ref{fig:workflow}-source~data~\ref{figdata:viralsequences}).
+The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (\FIG{workflow}-source~data~\ref{figdata:viralsequences}).
 The aligned deep sequencing data are available on the GEO repository under accession GSE108041 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE108041}).
@@
 We used custom Python code to annotate the cells in this matrix by the number of viral reads that could be assigned to the wildtype and synonymously barcoded virus.
-Only about half of the viral reads overlapped the barcoded regions of the genes (Figure~\ref{fig:workflow}A) and could therefore be assigned to a viral barcode (Figure~\ref{fig:viralbarcodes}-Figures~supplement~\ref{figsupp:barcodescalled}).
+Only about half of the viral reads overlapped the barcoded regions of the genes (\FIG{workflow}A) and could therefore be assigned to a viral barcode (\FIGSUPP[viralbarcodes]{barcodescalled}).
 So for calculations of the number of reads in a cell derived from each viral barcode for each viral gene, the total number of detected molecules of that gene are multiplied by the fraction of those molecules with assignable barcodes that are assigned to that barcode.
@@
 A Jupyter notebook that performs these analyses is in Supplementary~file~\ref{suppfile:code}.
-For each sample, cell barcodes that had $>$2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were excluded from downstream analyses (see red vertical lines in Figure~\ref{fig:cells}B).
+For each sample, cell barcodes that had $>$2.5-fold fewer or more UMI counts mapping to cellular transcripts than the sample mean were excluded from downstream analyses (see red vertical lines in \FIG{cells}B).
 
-In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}).
+In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (\FIG{viralbarcodes}B,C and \FIGSUPP[viralbarcodes]{thresholds}).
 We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
 This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
-As is apparent from Figure~\ref{fig:viralbarcodes}B, only the 10-hour sample and the 8hr-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
+As is apparent from \FIG{viralbarcodes}B, only the 10-hour sample and the 8hr-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
 The likely reason is that these samples have more total viral mRNA (and so there is more available mRNA to be acquired from lysed cells); in addition, there is always some experimental variability in the amount of cell lysis during the 10X sequencing process, and these samples may simply have the most.
 So the above threshold procedure is appropriate for those two samples.
-For the other samples, we simply set a minimum threshold of requiring at least a fraction 2$\times$$10^{-4}$ reads to come from viral mRNA as explained in the legend to Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
-The thresholds for each sample are shown in Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
+For the other samples, we simply set a minimum threshold of requiring at least a fraction 2$\times$$10^{-4}$ reads to come from viral mRNA as explained in the legend to \FIGSUPP[viralbarcodes]{thresholds}.
+The thresholds for each sample are shown in \FIG{viralbarcodes}C and \FIGSUPP[viralbarcodes]{thresholds}.
 This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
@@
 Cellular genes with expression that co-varied with the fraction of viral mRNAs in a cell were then determined using the Monocle differentialGeneTest, after removing variance explained by sample to sample variation.
-Figure~\ref{fig:cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
+\FIG{cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
 We performed the gene set analysis using the P -alues from the Monocle differentialGeneTest with \texttt{piano}~\citep{varemo2013enriching} using the hallmark gene set from GSEA v6~\citep{subramanian2005gene} and Fisher's method.

### 2018-01-14 paper/ResponseToReviews.tex
added simple simulation model requested by reviewers

@@
 As a fun point of comparison, the Gini coefficients for the variability of viral mRNA across cells ($\ge$0.64 for all samples) exceed the Gini coefficient for the unevenness of income distribution in the United States.
-}
 
-{\color{red}
-Still need to think about a simple model.
+We have also simulated the expected distribution under a simple model where the heterogeneity is due only to Poisson co-infection and whether or not a cell has the full complement of RNP genes.
+This simulation and its results are in Figure 5--source data 2. 
+The simulation gives a Gini coefficient as large as the actual one, but the shape of the simulated distribution is essentially bimodal, whereas the actual distribution is much more continuous and broad.
+Therefore, there are also other sources of heterogeneity that broaden the distribution beyond what is due only to RNP presence / absence and co-infection. 
 }

### 2018-01-14 paper/paper.tex
added simple simulation model requested by reviewers

@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
+\centerline{\includegraphics[width=0.83\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
 \caption{\label{fig:fluburdenbyflugene}
@@
 
+\figdata{\label{figdata:simplemodel} Simulation with a simple model for the expected heterogeneity due to Poisson co-infection and presence / absence of the full RNP is in \texttt{simple\_Poisson\_model.html}.}
+
 \end{figure}
@@
 Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
+Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) is does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.
 

### 2018-01-15 paper/ResponseToReviews.tex
revised last paragraph of Discussion as requested by reviewers

@@
 
-{\color{red} needs to be re-worded}
+{\color{black} 
+We have re-worded the sentences in question (the last paragraph of the Discussion).
+We now make clear that we find little effect of innate-immune activation in our experiments.
+We do still mention that we find \emph{at least} a minor role for host-cell state, since the new experiments described above show that activation of oxidative stress response genes does appear to play at least some causative role.
+}
 

### 2018-01-15 paper/paper.tex
revised last paragraph of Discussion as requested by reviewers

@@
 Overall, our work shows the power and importance of characterizing cellular infection at the level of single cells~\citep{avraham2015pathogen}.
-The dynamics of viral infection in any given cell is shaped by the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes initiated by a single copy of each viral gene.
-We have shown how the confluence of these factors leads to extreme cell-to-cell heterogeneity in the transcriptional outcome of influenza virus infection.
-Further deconstruction of the contributions of each factor will enable a deeper understanding of how the bulk features of infection emerge from the processes occurring within individual virus-infected cells.
+Viral infection can involve heterogeneity in the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes initiated by a single copy of each viral gene.
+In our experiments with short-timeframe and low-MOI infections with a relatively pure stock of influenza virus, we find only a minor role for innate-immune activation, but a substantial role for heterogeneity in the complement of viral genes that are expressed in individual cells and at least some contribution of host-cell state.
+Our current experiments are not able to quantify the role of other possibly important factors such as mutations in viral genes, but we suspect that they may also contribute.
+Future extensions of the approaches described here should enable further deconstruction of the sources of cell-to-cell heterogeneity during viral infection, thereby enabling a deeper understanding of how the bulk features of infection emerge from processes within individual virus-infected cells.
 

### 2018-01-15 paper/paper.tex
cleaned up comments in paper

@@
 \caption{\label{fig:fluexpr}
-\jdbcomment{add expression wedges}
+\jdbcomment{Alistair, can you add expression wedges for panels A and C?}
 Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
@@
 \caption{\label{fig:tsne}
-\jdbcomment{add arrow}
+\jdbcomment{Alistair, can you add an arrow pointing to the IFN positive cell in panels D, E, and F.}
 A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
@@
 Cells colored by the experimental sample.
-{\bf (D), (E)}
+{\bf (D)}
 Cells colored by the number of detected transcripts from type I and III interferons (IFN).
@@
 \caption{\label{fig:cellulargenes}
-\jdbcomment{add gray arrow wedge}
+\jdbcomment{Alistair, can you add a gray arrow wedge indicating expression increases from left to right.}
 Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
@@
 }
-{\jdbcomment{add figure}}
+{\jdbcomment{Alistair, and you add the figure.}}
 
@@
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}.
+Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.}
 
@@
 This ZIP file contains a Jupyter notebook that runs CellRanger to align and annotate the reads, and a Jupyter notebook that uses Monocle to analyze the cell-gene matrix.
-The ZIP file also includes associated custom scripts.}
+The ZIP file also includes associated custom scripts.
+To just run the Monocle analysis in \texttt{monocle\_analysis.ipynb} on a pre-generated cell-gene matrix, unpack Supplementary~file~\ref{suppfile:cellgenematrix} into \texttt{./results/cellgenecounts/}.}
 \end{suppfile}
@@
 The annotated cell-gene matrix in Matrix Market Format.
-\jdbcomment{This file is too large for the \textit{eLife} submission system.
-We will communicate with the editors to get it uploaded for a final accepted version, or we will post it on DataDryad.
-If you are a reviewer and need access, please contact the editor.}
+This is the matrix generated in \texttt{./results/cellgenecounts/} by running the CellRanger analysis in \texttt{align\_and\_annotate.ipynb} in Supplementary~file~\ref{suppfile:code}.
 }

### 2018-01-15 paper/paper.tex
one last manuscript comment

@@
 \figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
+\jdbcomment{Alistair, can you remove GPX2, which is no longer significant.}
 Table delineating genes in \FIG{cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.

### 2018-01-15 paper/eLife_original_submission.tex
add eLife digest answers

--- 2018-01-15 paper/paper.tex
@@
 We validated that our stocks exhibited greater purity of infectious particles than a stock propagated at high MOI by verifying that they had a higher ratio of infectious particles to virion RNA (\FIG{viruspopulations}A) and to particles capable of inducing expression of a single viral protein (\FIG{viruspopulations}B).
-In addition, viral stocks with many defective particles are more immunostimulatory per infectious unit (e.g., TCID50) than low-defective stocks~\citep{Tapia:2013kf,Lopez:2014en}, in part simply because there are more physical virions per infectious unit (\FIG{viruspopulations}A,B).
-We confirmed that our viral stocks induced less interferon per infectious unit than a stock propagated at higher MOI (\FIG{viruspopulations}C).
+In addition, viral stocks with many defective particles are more immunostimulatory~\citep{Tapia:2013kf,Lopez:2014en}.
+We confirmed that our viral stocks induced less interferon than a stock propagated at higher MOI (\FIG{viruspopulations}C).
 
@@
 {\bf (C)} 
-Our viral stocks are less immunostimulatory than virus propagated at high MOI when used at the same number of infectious units as calculated by TCID50.
-Note that this fact does not necessarily imply that they are more immunostimulatory per virion, as the high-MOI stocks also have more virions per infectious unit as shown in the first two panels. 
+Our viral stocks are less immunostimulatory than virus propagated at high MOI. 
 Measurements of \textit{IFNB1} transcript by qPCR normalized to the housekeeping gene L32 in A549 cells at 10 hours post infection at an MOI of 0.5.
@@
 \figsupp[Full flow cytometry data for panel B . \label{figsupp:HAstain}]{
-Full flow cytometry data for  \FIG{viruspopulations}B. 
+Full flow cytometry date for \FIG{viruspopulations}B. 
 A549 cells were infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
@@
 {\bf (D)} Cells infected with a stock of wild-type virus propagated at a high MOI, and therefore enriched in defective particles.
-}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}
-\end{figure}
+}{\includegraphics[width=0.7\linewidth]{figures/Validating_barcode_virus/HA_stain_supplement.png}}\end{figure}
 
 \subsection{Single cells show an extremely wide range of expression of viral mRNA.}
-We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, performing two slightly different variants of the experiment for the 8-hour timepoint.
-For most of the samples, we replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
-However, for the second 8-hour sample (which we denote as ``8hr-2'' in the figures), we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
-The rationale for including a sample without a media change was to determine the importance of synchronicity of the timing of infection as discussed later in this subsection.
-
+We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
+We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
+The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
 We recovered between 3,000 and 4,000 cells for each sample (\FIG{cells}A). 
-As expected for a low-MOI infection, most cells expressed little or no viral mRNA (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
-Also as expected, the amount of viral mRNA per cell among infected cells increased over time (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
+As expected for a low-MOI infection, most cells expressed little or no viral mRNA (\FIG{cells}B).
+Also as expected, the amount of viral mRNA per cell among infected cells increased over time (\FIG{cells}B).
 But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
-While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (\FIG{cells}B, \FIGSUPP[cells]{viral_mRNA_cumulfrac}).
+While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (\FIG{cells}B).
 
@@
 Cells outside the dotted green lines were considered outliers with suspiciously low or high amounts of cellular mRNA (possibly derived from two cells per droplet), and were excluded from all subsequent analyses.
-\FIGSUPP[cells]{viral_mRNA_cumulfrac} shows the exact distributions of the fraction of viral mRNA per cell.
 }
-\figsupp[Cumulative fraction plot of proportion of total mRNA from virus. \label{figsupp:viral_mRNA_cumulfrac}]{
-For each sample, this plot shows the fraction of all cells that derive at least the indication fraction of their mRNA from influenza virus.
-}{\includegraphics[width=0.5\linewidth]{figures/p_mRNA_per_cell_cumul.pdf}}
 \end{figure}
@@
 However, cells with a small amount of viral mRNA often have viral transcripts from both variants, as expected from the random partitioning associated with simple mRNA leakage.
-Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, likely reflecting co-infection.
+Finally, a few cells with large amounts of viral mRNA have viral transcripts from both variants, reflecting co-infection.
 
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.82\linewidth]{figures/p_frac_flu_summary.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_frac_flu_summary.pdf}}
 \caption{\label{fig:viralbarcodes}
@@
 {\bf (C)}
-We determined a threshold for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
+We determined a cutoff for calling ``true'' infections by finding the amount of viral mRNA per cell at which the viral barcode purity no longer increases with more viral mRNA.
 The purity is the fraction of all viral mRNA in a cell derived from the most abundant viral barcode in that cell.
-We fit a curve (orange line) to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff (dotted green line) at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
-This plot illustrates the process for the 10-hour sample, see \FIGSUPP[viralbarcodes]{thresholds} for similar plots for other samples.
+We fit a curve to the mean purity of all cells with more than the indicated amount of viral mRNA, and drew the cutoff at the point where this curve stopped increasing with the fraction of total mRNA derived from virus.
 See the Methods for details.
@@
 Distribution of the fraction of mRNA per cell derived from virus for both infected and co-infected cells.
-\FIGSUPP[viralbarcodes]{cumulfracflu} shows these same data in a cumulative fraction plots and calculates Gini coefficients to quantify the heterogeneity in viral mRNA load.
 }
-
 \figsupp[Number of viral barcodes called.\label{figsupp:barcodescalled}]{
@@
 }{\includegraphics[width=\linewidth]{figures/synbarcodes_umistats.jpg}}
-
-\figsupp[Thresholds for calling infected cells. \label{figsupp:thresholds}]{
-Cell lysis can lead cells to the spurious association of small amounts of extraneous mRNA with individual cells.
-We wanted to avoid classifying as infected cells that had simply acquired such lysis-derived viral mRNA.
-The amount of lysis-derived viral mRNA will vary among samples as a function of both the lysis rate during the cell preparation (which always varies slightly from sample to sample in the 10X procedure) and with the amount of total viral mRNA for that sample (the more viral mRNA, the more there is to be acquired from lysed cells).
-As is shown in \FIG{viralbarcodes}B, the 8hr-2 and 10hr sample clearly have an enrichment of mixed barcodes in cells with small numbers of viral mRNA.
-For each sample, we calculated the mean purity of all cells with at least the indicated amount of viral mRNA, and determined the threshold amount of viral mRNA where purity no longer increased by finding the first maxima in a loess curve fit (orange line).
-We called the threshold at this point of maximum purity (dotted green line).
-For the 6hr and 8hr samples there is no indication of contamination from lysis-derived reads, as \FIG{viralbarcodes}B shows no increase in mixed barcodes in low viral mRNA cells.
-Therefore, for these samples we simply set a threshold of requiring at least 2$\times$10$^{-4}$ of the total mRNA to come from virus, which corresponds to $\sim$2 viral mRNAs for the typical cell with $10^4$ total reads (\FIG{cells}B).
-}{\includegraphics[width=\linewidth]{figures/p_purity_threshold.pdf}}
-
-\figsupp[Cumulative distributions of viral mRNA per cell and Gini coefficients.\label{figsupp:cumulfracflu}]{
+\figsupp[Fraction of total viral mRNA derived from a given fraction of infected cells.\label{figsupp:cumulfracflu}]{
 The total fraction of all viral mRNA among infected cells that is attributable to a given fraction of these cells.
-For instance, the plot for the 8hrs sample shows that $\sim$50\% of all viral mRNA is derived from $\sim$8\% of the infected cells.
-The facet titles above each plot also give the Gini coefficient~\citep{gini1921measurement} that calculates the heterogeneity in the distribution of viral mRNA among infected cells.
-Gini coefficients of 0 indicate a perfectly even distribution across cells, and Gini coefficients of 1 indicate a maximally skewed distribution.
+For instance, the plot for the 8-hour sample shows that roughly 50\% of all viral mRNA is derived from 10\% of the infected cells.
 }{\includegraphics[width=\linewidth]{figures/p_cumul_flu.pdf}}
-
-\figsupp[Synchronization of infection does not greatly affect heterogeneity. \label{figsupp:synchronize}]
-{\jdbcomment{Alistair, can you add caption}
-}{\jdbcomment{Alistair, can you add add figure}}
-
-\figsupp[Effects of infectious dose or coinfection state. \label{figsupp:flowdose}]
-{
-{\bf (A)}
-Flow cytometry analysis of expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
-Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
-HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
-While a higher dose leads to more cells expressing high amounts of viral protein, it does not greatly increase the amount of viral protein in either the low-expressing or high-expressing cells. 
-Therefore, higher viral dose does not lead to a large continuous increase in viral protein production among all cells -- rather, it mostly changes the proportions of cells that fall in different parts of the highly heterogeneous distribution.
-{\bf (B)}
-Cells were co-infected with a mix of wild-type virus and pseudovirus in which the HA gene was replaced by GFP flanked by the terminal regions of the HA gene segment at an MOI of 0.1 for each virus. 
-At 10 hours post-infection, cells were stained for NS1 and HA expression and analyzed by flow cytometry for these proteins and GFP.
-Cells could be annotated as infected by virions of the same type (wild-type infection indicated by presence of HA, or pseudovirus infection indicated by the presence of GFP) or both types of virions (indicated by presence of HA and GFP).
-Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
-Replicates shown.
-}{\includegraphics[width=\linewidth]{figures/Coinf.jpg}}
 \end{figure}
 
-We determined the threshold amount of viral mRNA per cell for each sample at which the barcode partitioning clearly resulted from infection rather than leakage (\FIG{viralbarcodes}C, \FIGSUPP[viralbarcodes]{thresholds}), and used these thresholds to annotate cells that we were confident were truly infected.
+We determined the threshold amount of viral mRNA per cell at which the barcode partitioning clearly resulted from infection rather than leakage (\FIG{viralbarcodes}C), and used this threshold to annotate cells that we were confident were truly infected.
 We also annotated as co-infected cells above this threshold that had mRNA from both viral variants.
 \FIG{viralbarcodes}D shows the number of cells annotated as infected and co-infected for each sample -- these cells are just a small fraction of the number of cells with any viral read.
-These annotation thresholds are conservative, and may miss some true low-level infections.
-However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the possible loss of some low-level infections in order to avoid false positives.
-In addition, the synonymous viral barcodes only identify co-infections by viruses with different barcodes -- since the barcodes are at roughly equal proportion, we expect to miss about half of the co-infections.
-Since we annotate about $\sim$10\% of the infected cells as co-infected by viruses with different barcodes (\FIG{viralbarcodes}D), we expect another $\sim$10\% of the infected cells to also be co-infected but not annotated as so by our approach.
+These annotation thresholds are conservative, and will miss some true low-level infections, as well as any co-infections with the same viral variant.
+However, it is important that the analyses below are restricted to cells that are truly infected with virus, so we accepted the loss of some low-level infections in order to avoid false positives.
 Because most cells are not infected, we subsampled the uninfected cells to the numbers shown in \FIG{viralbarcodes}D to balance the proportions of infected and uninfected cells for all subsequent analyses.
@@
 The fraction of viral mRNA per infected cell follows a roughly exponential distribution, with many cells having few viral transcripts and a few cells having many.
-At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $<$15\% of infected cells produce over half the viral transcripts (\FIGSUPP[viralbarcodes]{cumulfracflu}).
-One way to quantify the heterogeneity of a distribution is to calculate the Gini coefficient~\citep{gini1921measurement}, which ranges from 0 for a completely uniform distribution, to 1 for a maximally skewed distribution.
-\FIGSUPP[viralbarcodes]{cumulfracflu} shows the Gini coefficients for the distribution of viral mRNA across infected cells for each sample.
-The Gini coefficients are $\ge$0.64 for all samples.
-As a fun point of comparison, these Gini coefficients indicate that the distribution of viral mRNA across infected cells is more uneven than the distribution of income in the United States~\citep{alvaredo2011note}.
-
-One possible source of heterogeneity in the amount of viral mRNA per cell is variability in the timing of infection.
-If some cells are infected earlier in the experiment than others, then they might have substantially more viral mRNA.
-However, several lines of evidence indicate that this is not the major cause of heterogeneity across cells.
-First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (\FIG{viralbarcodes}E, \FIGSUPP[viralbarcodes]{cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
-Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\jdbcomment{Alistair, can you add relevant citation for this method?}.
-As shown in \FIGSUPP[viralbarcodes]{synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
-Finally, viral mRNA expression from the secondary spread of virus from infected cells does not appreciably occur during the timeframes of our experiments, since \FIG{viralbarcodes}B does not show the pervasive presence of mixed barcodes that would occur in this case.
-Therefore, variability in the timing of infection is not the dominant cause of the cell-to-cell heterogeneity in our experiments.
-
-Notably, \FIG{viralbarcodes}E shows that there are co-infected cells with both low and high amounts of viral mRNA, suggesting that the initial infectious dose does not drive a simple continuous increase in viral transcript production.
-In support of this view, we used flow cytometry to quantify the levels of individual viral proteins in cells infected at various MOIs or for which we could delineate co-infection status (\FIGSUPP[viralbarcodes]{flowdose}).
-This analysis shows that sub-populations of cells that express similarly low and high levels of viral proteins persist across a wide range of infectious doses, although co-infection can influence the relative proportion of infected cells that fall into these sub-populations (\FIGSUPP[viralbarcodes]{flowdose}).
+At 6 and 8 hours $<$10\% of infected cells are responsible for over half the viral transcripts, while at 10 hours $\approx$15\% of infected cells produce over half the viral transcripts (\FIGSUPP[viralbarcodes}]{cumulfracflu}).
+Notably, co-infected cells do not express dramatically more viral mRNA than other cells (\FIG{viralbarcodes}), suggesting that the initial infectious dose is not the major driver of cell-to-cell variation in our low-MOI infections.
 
@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.83\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_burden_flu_gene_merge.pdf}}
 \caption{\label{fig:fluburdenbyflugene}
@@
 {\bf (A)} 
-The normalized expression of each viral gene as a function of the total fraction of mRNA in each infected cell derived from virus, taken over all time points.
+Fraction of mRNA in each infected cell derived from virus as a function of the normalized expression of each viral gene in that cell, taken over all time points.
 Cells with high viral burden always express all RNP genes, but some cells with high viral burden lack each of the other genes.
-Plots for individual samples are in \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, and a plot that excludes known coinfected cells is in \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}.
 {\bf (B)}
@@
 A Wilcoxon signed-rank test was used to test the null hypothesis that absence of each gene does not affect viral burden: **** = $P < 10^{-4}$, *** = $P < 10^{-3}$,  * = $P < 0.05$, ns = not significant.
-The trends are similar if we look only at the 10-hour sample (\FIGSUPP[fluburdenbyflugene]{10hrfluburdenbyflugene}) or exclude known co-infected cells (\FIGSUPP[fluburdenbyflugene]{fluburdenbyflugene_nocoinfection}).
 {\bf (C)}
@@
 }
-
 \figsupp[Secondary transcription is a major source of viral mRNA during bulk infections.\label{figsupp:cyclohexamide}]{
@@
 }{\includegraphics[width=0.3\linewidth]{figures/Primary_transcription_measurements/Cyclohexamide.pdf}}
-
 \figsupp[Like panel (A), but shows samples individually.\label{figsupp:fluburdensamplebreakdown}]{
-The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
+The fraction of mRNA in each infected cell derived as a function of the normalized expression of each viral gene, shown for the 10-hour and 8-hour samples individually (the other samples had too few infected cells for this analysis to be useful).
 Points are colored as in \FIG{fluburdenbyflugene}A.
 }{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_by_sample.pdf}}
-
-\figsupp[Like panel (A), but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburden_nocoinfection}]{
-The normalized expression of each viral gene as a function of the fraction of total mRNA derived from virus, excluding cells that were annotated as coinfected based on the presence of both viral barcodes.
-}{\includegraphics[width=0.8\linewidth]{figures/p_flu_burden_flu_gene_no_coinfect.pdf}}
-
 \figsupp[Like panel (B) but for the 10-hr sample only.\label{figsupp:10hrfluburdenbyflugene}]{
@@
 }{\includegraphics[width=\linewidth]{figures/p_10hr_flu_burden_flu_gene_test}}
-
-\figsupp[Like panel (B) but excludes coinfected cells with mixed viral barcodes.\label{figsupp:fluburdenbyflugene_nocoinfection}]{
-All findings in \FIG{fluburdenbyflugene}B remain unchanged if we exclude cells called as coinfected based on the presence of mixed viral barcodes.
-}{\includegraphics[width=\linewidth]{figures/p_no_coinfect_flu_burden_flu_gene_test}}
-
-\figdata{\label{figdata:fracmissing} The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
-
-\figdata{\label{figdata:simplemodel} Simulation with a simple model for the expected heterogeneity due to Poisson co-infection and presence / absence of the full RNP is in \texttt{simple\_Poisson\_model.html}.}
-
+\figdata{The numerical data for panel (C) are in \texttt{p\_missing\_genes.csv}.}
 \end{figure}
 
-We examined the total amount of viral mRNA versus the expression of the genes from each viral segment (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
-Note that influenza virus expresses ten major gene transcripts from its eight gene segments, as the M and NS segments are alternatively spliced to produce the M1 / M2 and NS1 / NEP transcript, respectively~\citep{dubois2014influenza}.
-However, an inherent limitation of current established single-cell mRNA sequencing techniques is that they only sequence the 3' end of the transcript~\citep{zheng2017massively,macosko2015highly,Klein:2015ki,cao2017comprehensive}.
-Since the alternative spliceoforms M1 / M2 and NS1 / NEP share the same 3' ends, we cannot distinguish them and therefore will refer simply to the combined counts of transcripts from each of these alternatively spliced segments as the M and NS genes.
-
-Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
+We examined the total amount of viral mRNA versus the expression of each viral gene (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}). 
+Cells that lack an RNP gene never derive more than a few percent of their mRNAs from virus, confirming the expected result that all four RNP genes are essential for high levels of viral transcription.
 However, we observe cells that lack each of the other non-RNP genes but still derive $\approx$40\% of their mRNAs from virus, suggesting that none of the other genes are important for high levels of viral transcription.
@@
 However, gene absence clearly does not explain all of the variability in viral gene expression, since even cells expressing all viral genes exhibit a very wide distribution in the amount of viral mRNA that they express.
-Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
-Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) is does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.
+Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}).
 
@@
 We limited this analysis to examining the presence / absence of the non-RNP genes in cells expressing all four RNP genes, since we might fail to detect viral transcripts that are actually present at low levels in RNP-deficient cells due to the lower viral burden in these cells.
-At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (\FIG{fluburdenbyflugene}C, \FIG{fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
-The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (\FIG{fluburdenbyflugene}C and \FIG{fluburdenbyflugene}-source~data~\ref{figdata:fracmissing}).
-If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 35-50\% of infected cells express mRNAs from all eight genes.
+At the 8- and 10-hour time points, between 5\% and 17\% of cells fail to express any one of the four non-RNP genes (\FIG{fluburdenbyflugene}C).
+The absence of a given gene appears to be an independent event, as the probability of observing all four non-RNP genes in a cell is well predicted by simply multiplying the probabilities of observing each gene individually (\FIG{fluburdenbyflugene}C).
+If we extrapolate the frequencies at which cells lack non-RNP genes to the RNP genes, then we would predict that 40-50\% of infected cells express mRNAs from all eight genes.
 This estimate of the frequency at which infected cells express mRNAs from all eight gene segments is slightly higher than previous estimates of 13\%~\citep{Brooke:2013kb} and 20\%~\citep{Dou:2017cp}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_manualedits.pdf}}
 \caption{\label{fig:fluexpr}
-\jdbcomment{add expression wedges}
 Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
@@
 Seven of these 10 cells expressed all eight viral genes.
-The majority (4 of 7) of these cells would \emph{not} have expressed all the viral genes in the absence of co-infection, since they have at least one gene exclusively derived from each viral variant.
+Remarkably, the majority (4 of 7) of these cells would \emph{not} have expressed all the viral genes in the absence of co-infection, since they have at least one gene exclusively derived from each viral variant.
 For instance, the cell with 11.2\% of its mRNA from virus in the upper right of \FIG{coexpression} expresses M only from the wildtype viral variant, and NP and HA only from the synonymously barcoded variant.
-Our data therefore provide the first direct single-cell observation of the fact that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,brooke2014influenza,Fonville:2015cg,aguilera2017plaques}.
+Our data therefore provide direct single-cell support for the idea that co-infection can rescue missing viral genes~\citep{Brooke:2013kb,brooke2014influenza,Fonville:2015cg,aguilera2017plaques}.
 
@@
 \end{figure} 
-
 Another observation from \FIG{coexpression} is that co-infected cells usually express roughly equal amounts of transcripts from each of the two viral variants.
@@
 Influenza virus infection can trigger innate-immune sensors that lead to the transcriptional induction of type I and III interferons, and subsequently of anti-viral interferon-stimulated genes~\citep{Killip:2015dw,Iwasaki:2014dw,crotta2013type}.
+%In bulk cells, this interferon response is one of the most prominent transcriptional signatures of influenza-virus infection~\citep{Geiss:2002a}.
 However, activation of the interferon response is stochastic and bi-modal at the level of single cells~\citep{Chen:2010cr,shalek2013single,shalek2014single,PerezCidoncha:2014jr,bhushal2017cell,hagai2017gene}.
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge_manualarrow.pdf}}
 \caption{\label{fig:tsne}
-\jdbcomment{add arrow}
 A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
@@
 We restricted this analysis to infected cells with all eight viral genes in order to focus on cellular genes that were associated with viral mRNA burden independent of effects due to the presence or absence of particular viral transcripts.
-We identified 43 cellular genes that co-varied with viral mRNA expression at a false discovery rate of 0.1 (\FIG{cellulargenes}, \FIG{cellulargenes}-source~data~\ref{figdata:cellulargenes}).
+We identified 45 cellular genes that co-varied with viral gene expression at a false discovery rate of 0.1 (\FIG{cellulargenes}).
 
-\begin{figure}[t!]
-\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap.pdf}}
+\begin{figure}
+\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap_manualedit.pdf}}
 \caption{\label{fig:cellulargenes}
-\jdbcomment{add gray arrow wedge}
 Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
@@
 }
-
-\figdata{The full results of the differential expression test are in \texttt{p\_sig\_cellular\_genes.csv}. \label{figdata:cellulargenes}}
-
-\figdata{A gene-set analysis for pathways associated with the amount of viral mRNA is in \texttt{p\_pathway\_enrichment.csv}. \label{figdata:geneset}}
-
+\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
+\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
 \figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
@@
 }{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
-
-\figsupp[Pre-treating to induce oxidative stress increases the fraction of infected cells expressing detectable viral protein. \label{figsupp:oxidative}]
-{
-\jdbcomment{Alistair, can you add figure and caption showing the increase in the percent of flu positive cells after treatment.
-It should just be possible to put the methods for this experiment in the figure caption.}
-}
-{\jdbcomment{add figure}}
-
 \end{figure}
 
-A gene-set analysis shows that many cellular genes that are associated with the amount of viral mRNA are involved in the response to reactive oxygen species or hypoxia (\FIG{cellulargenes}-source~data~\ref{figdata:geneset}).
-Genes known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress are often expressed at higher levels in cells with more viral mRNA (\FIG{cellulargenes}).
+Many of the genes with increased expression in cells with more viral mRNA are known or suspected to be regulated by the Nrf2 master regulator in response to oxidative stress.
 These genes produce proteins that are involved in detoxification of reactive oxygen species or resultant products, the management of misfolded proteins, the electron transport chain, or a general stress response (\FIGSUPP[cellulargenes]{ROStable}).
@@
 The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
-To directly test the effect of transient oxidative stress, we compared the fraction of cells that expressed detectable viral protein when infected either with or without pre-treatment to induce oxidative stress.
-\FIGSUPP[cellulargenes]{oxidative} shows that the cells pre-treated to induce oxidative stress exhibited more frequent detectable expression of viral protein.
-These results, in conjunction with the differential expression test in \FIG{cellulargenes} and the prior work mentioned above, suggest that oxidative stress acts in a proviral fashion.
-
-The gene-set analysis also found that the amount of viral mRNA was associated with the expression of genes involved in the G2-M cell-cycle checkpoint (\FIG{cellulargenes}-source~data~\ref{figdata:geneset}).
-The cell-cycle associated genes CCND3, MKI67, UBE2S, and CENPF are all expressed at significantly lower levels in cells with more viral mRNA (\FIG{cellulargenes}).
-However, our data are not sufficient to determine whether the lower expression of these genes is a cause or effect of the reduction in viral mRNA.
-
-Interestingly, none of the cellular genes that are significantly associated with the amount of viral mRNA in our study are among the 128 genes that \citet{watanabe2010cellular} report as having been identified multiple times in genome-wide screens for factors affecting influenza virus replication.
-One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral factors, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.
+Our data do not reveal whether the expression of genes involved in the response to oxidative stress are a cause or a symptom of higher levels of viral mRNA, and further investigation of this topic is an interesting area for future work.
 
@@
 
-Overall, our work shows the power and importance of characterizing cellular infection at the level of single cells~\citep{avraham2015pathogen}.
-Viral infection can involve heterogeneity in the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes initiated by a single copy of each viral gene.
-In our experiments with short-timeframe and low-MOI infections with a relatively pure stock of influenza virus, we find only a minor role for innate-immune activation, but a substantial role for heterogeneity in the complement of viral genes that are expressed in individual cells and at least some contribution of host-cell state.
-Our current experiments are not able to quantify the role of other possibly important factors such as mutations in viral genes, but we suspect that they may also contribute.
-Future extensions of the approaches described here should enable further deconstruction of the sources of cell-to-cell heterogeneity during viral infection, thereby enabling a deeper understanding of how the bulk features of infection emerge from processes within individual virus-infected cells.
+Overall, our work shows the power and importance of charactering cellular infection at the level of single cells~\citep{avraham2015pathogen}.
+The dynamics of viral infection in any given cell is shaped by the genetic composition of the incoming virion, the host-cell state, the bi-modality of innate-immune activation, and the inherent stochasticity of molecular processes initiated by a single copy of each viral gene.
+We have shown how the confluence of these factors leads to extreme cell-to-cell heterogeneity in the transcriptional outcome of influenza virus infection.
+Further deconstruction of the contributions of each factor will enable a deeper understanding of how the bulk features of infection emerge from the processes occurring within individual virus-infected cells.
 
@@
 We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
-To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus grow for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
+To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
 The high-MOI passage (high-defective particle) stock used in \FIG{viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.
@@
 \subsection{Flow cytometry titering and analyses}
-To determine viral titers in terms of HA-expressing units and for the flow cytometry in \FIGSUPP[coexpression]{flowcyto} and \FIGSUPP[viralbarcodes]{flowdose}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
+To determine viral titers in terms of HA-expressing units and for the flow cytometry in \FIGSUPP[coexpression]{flowcyto}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
 Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
-After washing in PBS supplemented with 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
+After washing in PBS plus 2\% FBS, the cells were stained with a goat anti-mouse IgG antibody conjugated to APC.
 Cells were then washed, fixed in 1\% formaldehyde, and washed further before a final resuspension and analysis. 
 We then determined the fraction of cells that were HA positive and calculated the HA-expressing units.
-For NS1 staining in \FIGSUPP[viralbarcodes]{flowdose}, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
-washed, and analyzed.
-For \FIGSUPP[coexpression]{flowcyto} and \FIGSUPP[viralbarcodes]{flowdose}, after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
-For \FIGSUPP[viralbarcodes]{flowdose} channels were additionally compensated in FLowJo.
+
+For \FIGSUPP[coexpression]{flowcyto}, after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
-For \FIGSUPP[viralbarcodes]{flowdose}, the percentage of influenza-infected cells was determined by HA staining alone, and the top quantile of NS1-stained cells matching that percentage were taken as the NS1 positive population.
 
@@
 
-All time points except for the second 8-hour sample (8hr-2) were prepared on the same day.
+All time points except for the 8-hour replicate were prepared on the same day.
 For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
@@
 All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
-For 8hr-2 sample, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
+For our replicate 8-hour time point, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
 After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
@@
 Samples were sequenced on an Illumina HiSeq. 
+The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Upload in progress.}
 
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.
-If you are a reviewer who needs access, please contact the editor.}.
+Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.}
 
@@
 The influenza virus transcriptome was generated from the reverse-complement of the wildtype WSN viral RNA sequences as encoded in the reverse-genetics plasmids (\FIG{workflow}-source~data~\ref{figdata:viralsequences}).
-The aligned deep sequencing data are available on the GEO repository under accession GSE108041 (\url{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE108041}).
-
 CellRanger calls cells based on the number of observed cell barcodes, and creates a cell-gene matrix.
@@
 
-In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (\FIG{viralbarcodes}B,C and \FIGSUPP[viralbarcodes]{thresholds}).
+In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (\FIG{viralbarcodes}B,C).
 We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
 This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
-As is apparent from \FIG{viralbarcodes}B, only the 10-hour sample and the 8hr-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
-The likely reason is that these samples have more total viral mRNA (and so there is more available mRNA to be acquired from lysed cells); in addition, there is always some experimental variability in the amount of cell lysis during the 10X sequencing process, and these samples may simply have the most.
-So the above threshold procedure is appropriate for those two samples.
-For the other samples, we simply set a minimum threshold of requiring at least a fraction 2$\times$$10^{-4}$ reads to come from viral mRNA as explained in the legend to \FIGSUPP[viralbarcodes]{thresholds}.
-The thresholds for each sample are shown in \FIG{viralbarcodes}C and \FIGSUPP[viralbarcodes]{thresholds}.
+The 10-hour sample has the most viral mRNA, and so is expected to have the most lysis-associated influenza reads.
+We therefore determined the threshold using just this sample, and then applied this threshold to all samples.
+The threshold is shown in \FIG{viralbarcodes}C.
 This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
@@
 Candidate marker genes for t-SNE dimensionality reduction were then determined using the Monocle function markerDiffTable, excluding the effects of sample variation and the number of genes identified in a given cell, using a q-value cutoff of 0.01.
-The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 50 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
+The specificity of these markers was determined using the function calculateMarkerSpecificity -- the top 40 markers were retained, and used to place populations in a two-dimensional plane based on tSNE dimensionality reduction.
 
@@
 \FIG{cellulargenes} shows all genes that were significantly associated with the fraction of mRNA from virus at a false discovery rate of 0.1.
-We performed the gene set analysis using the P -alues from the Monocle differentialGeneTest with \texttt{piano}~\citep{varemo2013enriching} using the hallmark gene set from GSEA v6~\citep{subramanian2005gene} and Fisher's method.
 
@@
 
+%\nocite{*} % This command displays all refs in the bib file
 \bibliography{references}

### 2018-01-16 paper/paper.tex
Addressed remaining supplement and small comments. Changed wording of antioxidant section. Updated flow methods, took out specific figure references as was getting too cumbersome and added in NAC treatment to get a few important details inappropriate for caption. Hopefully ready to go.

--- 2018-01-15 paper/paper.tex
@@
 \figsupp[Synchronization of infection does not greatly affect heterogeneity. \label{figsupp:synchronize}]
-{\jdbcomment{Alistair, can you add caption}
-}{\jdbcomment{Alistair, can you add add figure}}
+{Flow cytometry analysis of expression of viral proteins in cells infected at an MOI of 0.1 (unsynchronized) or 0.2 (synchronized) as calculated by TCID50 on MDCK-SIAT1 cells. 
+The higher MOI for synchronized infections was to attempt to account for loss of virus in washing steps.
+Infections were synchronized by pre-adsorbing virus at 4$^{\circ}$C for 1h prior to initiation of infection by shifting temperature to 37$^{\circ}$C using pre-warmed media.
+Cells were concurrently stained for HA and NS1 proteins 10 hours post initiation of infection.
+Synchronization resulted in a small decrease in variability, particularly in intermediately-stained cells, but not a massive shift in distribution. 
+Replicates shown.
+}{\includegraphics[width=\linewidth]{figures/synchronize.jpg}}
 
@@
 First, the sample for which the infection inoculum was never removed (8hr-2) only shows slightly more heterogeneity than samples for which the inoculum was washed away after one hour (\FIG{viralbarcodes}E, \FIGSUPP[viralbarcodes]{cumulfracflu}), despite the fact that the potential time window for infection is much longer in the former sample.
-Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\jdbcomment{Alistair, can you add relevant citation for this method?}.
+Second, in an independent experiment, we performed completely synchronized infections by pre-binding virus to cells on ice and then washing away unbound virus before bringing the cells to 37$^{\circ}$C~\citep{Dapat:2014kv}.
 As shown in \FIGSUPP[viralbarcodes]{synchronize}, flow cytometry staining found that the heterogeneity in the levels of individual viral proteins was not markedly different for these synchronized infections than in the absence of pre-binding and washing.
@@
 \caption{\label{fig:fluexpr}
-\jdbcomment{Alistair, can you add expression wedges for panels A and C?}
 Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
@@
 \caption{\label{fig:tsne}
-\jdbcomment{Alistair, can you add an arrow pointing to the IFN positive cell in panels D, E, and F.}
 A t-SNE plot created by semi-supervised clustering using genes that co-vary with viral infection status.
@@
 \caption{\label{fig:cellulargenes}
-\jdbcomment{Alistair, can you add a gray arrow wedge indicating expression increases from left to right.}
 Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
@@
 
-\figsupp[Pre-treating to induce oxidative stress increases the fraction of infected cells expressing detectable viral protein. \label{figsupp:oxidative}]
-{
-\jdbcomment{Alistair, can you add figure and caption showing the increase in the percent of flu positive cells after treatment.
-It should just be possible to put the methods for this experiment in the figure caption.}
-}
-{\jdbcomment{Alistair, and you add the figure.}}
+\figsupp[Pre-treating to reduce oxidative stress decreases the fraction of infected cells expressing detectable viral protein. \label{figsupp:oxidative}]
+{Cells were treated as indicated with 15 mM N-acetylcysteine (NAC), an antioxidant, and infected at an MOI of 0.1 as calculated by TCID50 on MDCK-SIAT1 cells. 
+At 10 hours post-infection, cells were analyzed by flow cytometry for HA expression.
+Percentage HA-positive cells and gating shown.
+Replicates shown.
+}{\includegraphics[width=0.7\linewidth]{figures/antiox.jpg}}
 
@@
 The antioxidant response is thought to be largely antiviral, potentially through inhibition of MAPK activity \citep{Lin:2016ec,Sgarbanti:2014ht}.
-To directly test the effect of transient oxidative stress, we compared the fraction of cells that expressed detectable viral protein when infected either with or without pre-treatment to induce oxidative stress.
-\FIGSUPP[cellulargenes]{oxidative} shows that the cells pre-treated to induce oxidative stress exhibited more frequent detectable expression of viral protein.
+To directly test the effect of transient oxidative stress, we compared the fraction of cells that expressed detectable viral protein when infected either with or without pre-treatment to suppress oxidative stress.
+\FIGSUPP[cellulargenes]{oxidative} shows that the cells pre-treated with an antioxidant exhibited less frequent detectable expression of viral protein.
 These results, in conjunction with the differential expression test in \FIG{cellulargenes} and the prior work mentioned above, suggest that oxidative stress acts in a proviral fashion.
@@
 \subsection{Flow cytometry titering and analyses}
-To determine viral titers in terms of HA-expressing units and for the flow cytometry in \FIGSUPP[coexpression]{flowcyto} and \FIGSUPP[viralbarcodes]{flowdose}, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
+To determine viral titers in terms of HA-expressing units and for the flow cytometry, A549 cells were seeded in a 6-well plate and infected as described above for the qPCR analyses.
 Cells were harvested by trypsinization, resuspended in phosphate-buffered saline supplemented with 2\% heat-inactivated FBS, and stained with 10 \si{\micro}g/ml of H17-L19, a mouse monoclonal antibody confirmed to bind to WSN HA in a prior study~\citep{Doud:2017bw}.
@@
 We then determined the fraction of cells that were HA positive and calculated the HA-expressing units.
-For NS1 staining in \FIGSUPP[viralbarcodes]{flowdose}, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
+For NS1 staining, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
 washed, and analyzed.
-For \FIGSUPP[coexpression]{flowcyto} and \FIGSUPP[viralbarcodes]{flowdose}, after gating to exclude multiplets in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
-For \FIGSUPP[viralbarcodes]{flowdose} channels were additionally compensated in FLowJo.
+To analyze the effect of N-acetylcysteine, the compound was added to cells in D10 9h prior to media change and infection, and included in infection media.
+Stocks of N-acetylcysteine were reconstituted immediately prior to use, and the pH of growth media supplemented with this compound was adjusted using sodium hydroxide.
+After channels were compensated and cells gated to exclude multiplets and debris in FlowJo, data were extracted using the R package flowCore~\citep{LeMeur:2007uo} and analyzed using a custom Python script.
 Guassian kernel density estimates were obtained using the scipy stats package method, guassian\_kde, using automatic bandwidth determination~\citep{vanderWalt:2017dp}.
-For \FIGSUPP[viralbarcodes]{flowdose}, the percentage of influenza-infected cells was determined by HA staining alone, and the top quantile of NS1-stained cells matching that percentage were taken as the NS1 positive population.
+For gating on NS1 positive cells, the percentage of influenza-infected cells was determined by HA staining alone, and the top quantile of NS1-stained cells matching that percentage were taken as the NS1 positive population.
 

### 2018-01-16 paper/paper.tex
Renamed manually edited figures with suffix `_manualedits.pdf` so that they do not accidentally get overwritten by auto-created figures if analysis notebook is re-run

@@
 \begin{figure}
-\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge.pdf}}
+\centerline{\includegraphics[width=0.9\linewidth]{figures/p_small_tsne_merge_manualedits.pdf}}
 \caption{\label{fig:tsne}
@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap.pdf}}
+\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap_manualedits.pdf}}
 \caption{\label{fig:cellulargenes}

### 2018-01-16 paper/paper.tex
A few small tweaks to legends for new supplementary figures. Primary change is to better explain the dual staining strategy for HA / NS1

@@
 Infections were synchronized by pre-adsorbing virus at 4$^{\circ}$C for 1h prior to initiation of infection by shifting temperature to 37$^{\circ}$C using pre-warmed media.
-Cells were concurrently stained for HA and NS1 proteins 10 hours post initiation of infection.
-Synchronization resulted in a small decrease in variability, particularly in intermediately-stained cells, but not a massive shift in distribution. 
-Replicates shown.
+Cells were concurrently stained for HA and NS1 proteins at 10 hours after initiation of infection, and then analyzed by flow cytometry.
+The level of HA protein was quantified in cells that were identified as infected based on being positive for NS1 protein (top), and the level of NS1 protein was quantified in cells that were identified as infected based on being positive for HA protein (bottom)
+Synchronization resulted in a small decrease in variability in the levels of each viral protein, particularly in cells with intermediate levels.
+But the effects were small compared to the overall variability in the protein levels, indicating timing of infection makes only a small contribution to the observed heterogeneity. 
+Data are shown for three independent replicates.
 }{\includegraphics[width=\linewidth]{figures/synchronize.jpg}}
@@
 Flow cytometry analysis of expression of viral proteins in cells infected at high (MOI 5 as calculated by TCID50 on MDCK-SIAT1 cells) or low (MOI 0.1) initial infectious dose.
-Cells were concurrently stained for HA and NS1 proteins 10 hours post infection.
-HA staining was analyzed in cells positive for NS1 (left), and NS1 staining was analyzed in cells positive for HA (right).
+Cells were concurrently stained for HA and NS1 proteins at 10 hours after initiation of infection, and then analyzed by flow cytometry.
+The level of HA protein was quantified in cells that were identified as infected based on being positive for NS1 protein (left), and the level of NS1 protein was quantified in cells that were identified as infected based on being positive for HA protein (right)
 While a higher dose leads to more cells expressing high amounts of viral protein, it does not greatly increase the amount of viral protein in either the low-expressing or high-expressing cells. 
@@
 Coinfected cells, like cells infected at a higher infectious dose, occupy different positions in the distribution of viral protein production but do not not exhibit a continuous increase in viral protein production.
-Replicates shown.
+Data are shown for three independent replicates.
 }{\includegraphics[width=\linewidth]{figures/Coinf.jpg}}
@@
 At 10 hours post-infection, cells were analyzed by flow cytometry for HA expression.
-Percentage HA-positive cells and gating shown.
-Replicates shown.
+The percentage of HA-positive cells is indicated on the flow cytometry plots.
+Data are shown for three independent replicates.
 }{\includegraphics[width=0.7\linewidth]{figures/antiox.jpg}}

### 2018-01-16 paper/paper.tex
Fixed a missing comma in methods

@@
 We then determined the fraction of cells that were HA positive and calculated the HA-expressing units.
-For NS1 staining, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405.
-washed, and analyzed.
+For NS1 staining, cells stained for HA as described above were permeabilized using BD Cytofix/Cytoperm following manufacturer's instructions, stained with anti-NS1 (GTX125990, Genetex) at 4.4 \si{\micro}g/ml, washed,  stained with a  goat anti-rabbit IgG antibody conjugated to Alexa Fluor 405, washed, and analyzed.
 To analyze the effect of N-acetylcysteine, the compound was added to cells in D10 9h prior to media change and infection, and included in infection media.

### 2018-01-16 paper/paper.tex
a few small typo / wording fixes

@@
 Specifically, at both 8 and 10 hours, the amount of viral mRNA in individual cells expressing all eight viral genes still ranges from $<$1\% to $>$50\% (\FIG{fluburdenbyflugene}A, \FIGSUPP[fluburdenbyflugene]{fluburdensamplebreakdown}, \FIGSUPP[fluburdenbyflugene]{fluburden_nocoinfection}).
-Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) is does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.
+Furthermore, the actual distribution of viral mRNA per infected cell (\FIG{viralbarcodes}E) does not match the mostly bi-modal shape expected under a simple model where RNP gene absence and Poisson co-infection are the only factors (\FIG{fluburdenbyflugene}--source~data~\ref{figdata:simplemodel}), indicating that there are additional sources of variability beyond whether cells have full complement of RNP genes.
 
@@
 
-In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this mRNA derived from each gene is much more consistent across cells (\FIG{fluexpr}A).
+In contrast to the extreme variability in the total viral mRNA per cell, the fraction of this viral mRNA derived from each gene is much more consistent across cells (\FIG{fluexpr}A).
 Total viral mRNA varies by orders of magnitude, but the fraction from any given viral gene is fairly tightly clustered around the median value for all cells (\FIG{fluexpr}B).
@@
 \begin{figure}[t!]
-\centerline{\includegraphics[width=0.63\linewidth]{figures/p_cellular_heatmap_manualedits.pdf}}
+\centerline{\includegraphics[width=0.65\linewidth]{figures/p_cellular_heatmap_manualedits.pdf}}
 \caption{\label{fig:cellulargenes}
@@
 Interestingly, none of the cellular genes that are significantly associated with the amount of viral mRNA in our study are among the 128 genes that \citet{watanabe2010cellular} report as having been identified multiple times in genome-wide screens for factors affecting influenza virus replication.
-One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral factors, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.
+One possible explanation is that most of the cell-to-cell heterogeneity in our experiments might arise from viral segment absence or mutations, pure stochasticity, or more subtle alterations in host-cell state -- not due to changes in expression of the type of single large-effect genes that are usually identified in genome-wide knockdown / knockout studies.
 

### 2018-01-17 paper/paper.tex
added extra info requested by *eLife* for submission of revision

@@
 The following cell lines were used in this study: the human lung epithelial carcinoma line A549 (ATCC CCL-185), the MDCK-SIAT1 variant of the Madin Darby canine kidney cell line overexpressing human SIAT1 (Sigma-Aldrich 05071502), and the human embryonic kidney cell line 293T (ATCC CRL-3216).
+The A549 cells were tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core.
 All cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
@@
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell} \jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.}
+Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell}~\citep{russell2018github}.
+\jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.}
 
@@
 This is the matrix generated in \texttt{./results/cellgenecounts/} by running the CellRanger analysis in \texttt{align\_and\_annotate.ipynb} in Supplementary~file~\ref{suppfile:code}.
+\jdbcomment{This file is too large for the \textit{eLife} submission system, and so is in the process of being uploaded to DataDryad.}
 }

### 2018-01-17 paper/paper_nofigs.tex
other versions of files needed for submission
added extra info requested by *eLife* for submission of revision

--- 2018-01-16 paper/paper.tex
@@
 % Please note that these options may affect formatting.
+
+\renewcommand\includegraphics[2][]{}
 

### 2018-02-06 paper/paper.tex
added link to cell-gene matrix provided by *eLife*
removed extra `\end{document}`
updated repository commit

@@
 Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell}~\citep{russell2018github}.
-\jdbcomment{This GitHub repository will be made public upon acceptance of the manuscript.}
 
@@
 This is the matrix generated in \texttt{./results/cellgenecounts/} by running the CellRanger analysis in \texttt{align\_and\_annotate.ipynb} in Supplementary~file~\ref{suppfile:code}.
-\jdbcomment{This file is too large for the \textit{eLife} submission system, and so is in the process of being uploaded to DataDryad.}
+This file is available on DataDryad at \url{https://doi.org/10.5061/dryad.qp0t3}.
 }
@@
 \end{document}
-
-\end{document}

### 2018-02-13 paper/paper.tex
added STR profiling and eLife Github link

@@
 The following cell lines were used in this study: the human lung epithelial carcinoma line A549 (ATCC CCL-185), the MDCK-SIAT1 variant of the Madin Darby canine kidney cell line overexpressing human SIAT1 (Sigma-Aldrich 05071502), and the human embryonic kidney cell line 293T (ATCC CRL-3216).
-The A549 cells were tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core.
+The A549 cells were tested as negative for mycoplasma contamination by the Fred Hutch Genomics Core, and authenticated using the ATCC STR profiling service.
 All cells were maintained in D10 media (DMEM supplemented with 10\% heat-inactivated fetal bovine serum, 2 mM L-glutamine, 100 U of penicillin/ml, and 100 \si{\micro}g of streptomycin/ml) at 37 \si{\degreeCelsius} at 5 \% CO\textsubscript{2}.
@@
 \subsection{Computational analysis of single-cell mRNA sequencing data}
-Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell}~\citep{russell2018github}.
+Jupyter notebooks that perform all of the computational analyses are available in Supplementary~file~\ref{suppfile:code} and at \url{https://github.com/jbloomlab/flu_single_cell}~\citep[][copy archived at \url{https://github.com/elifesciences-publications/flu_single_cell}]{russell2018github}.
 

### 2018-02-13 paper/paper_nofigs.tex
added STR profiling and eLife Github link

@@
 
-\renewcommand\includegraphics[2][]{}
+> \renewcommand\includegraphics[2][]{}
 
